### 2014-03-28 tex/abstract.tex
initial commit.

%!TEX root = flock-comment-main.tex

      We show that the program {\sc flock} \citep{Duc&Tur2009} employs 
a model that is identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. The computational routine in {\sc flock} is 
formally equivalent to searching for the maximum-a-posteriori (MAP) 
estimate of this restricted {\sc structure} model via a simulated 
annealing algorithm with an extreme (and likely suboptimal) cooling 
schedule.  Understanding {\sc flock} as a special case of the model in 
{\sc structure} allows a principled perspective on its expected 
strengths and weaknesses.

### 2014-03-28 tex/author-title-etc.tex
initial commit.

%!TEX root = flock-comment-main.tex

%%%%% THIS IS THE SECTION WHERE THE AUTHOR PUTS IN ALL OF THEIR TITLE AND AFFILIATION
%%%%% INFORMATION AND  A FEW OTHER THINGS
\newcommand{\myTitle}{A statistically-grounded interpretation of {\sc flock}}

% redefine this to make author list. Note affiliation symbols are done manually.
% All caps tends to look better
\newcommand{\myAuthors}{ ERIC C.~ANDERSON$^{*,\dagger,\S}$ and PATRICK BARRY$^\ddag$}

% redefine to make the affiliation list.  Note symbols are done manually
\newcommand{\myAffiliations}{$^*$Fisheries Ecology Division, 
    Southwest Fisheries Science Center, National Marine Fisheries Service, NOAA,
    110 Shaffer Road,
    Santa Cruz, CA 95060, USA, $^\dagger$Department of Applied Math and Statistics, University of
California, Santa Cruz, CA, $^\ddag$University of Alaska, Fairbanks, Fisheries Academic Program,
17101 Point Lena Loop Rd.,
Juneau, AK 99801, USA }

% the email address for the corresponding author
\newcommand{\myEmailAddress}{eric.anderson@noaa.gov}
\newcommand{\myEmailFootnote}{$^\S$}

% here you can put your very own copyright notice
\newcommand{\myCopyright}{\copyright US Federal Government work in the public domain in the USA}

% here you can put a running title (a short title that goes on the left of the 
% even pages)
\newcommand{\myRunningTitle}{{\sc flock} as a special case of {\sc structure}}

% and here you put the running author (short listing of authors for the
% upper right header on the odd pages)
\newcommand{\myRunningAuthor}{Anderson {\em et al.}}

%%%% DONE WITH AUTHOR/TITLE/ETC INFORMATION DEFINITION

### 2014-03-28 tex/main-body-text.tex
initial commit.

%!TEX root = flock-comment-main.tex

\section*{Introduction}
In most fields, proposed methods that are 
perceived to be fundamentally novel or new garner more
attention and accolades---and have a higher chance of publication---
than methods that are clearly minor elaborations upon existing work. 
Not surprisingly, then, in articles and manuscripts, authors may 
emphasize the differences and downplay the similarities between their 
work and the published literature.   In molecular ecology, this 
tendency amongst the creators of statistical methodology can make it 
difficult for end-users to understand the relationship between 
different methods.

Of course, new methods almost always build upon pre-existing ones, and 
much is to be gained by understanding the close relationship between 
different statistical methods in use in molecular ecology today.  
Indeed, some authors downplay the similarities between their method and 
existing ones not because they wish to make their method seem more 
novel, but rather because they are genuinely unaware of the close ties 
between their work and existing methods.  Helping to identify the 
similarities between new and existing work should help the field 
progress more quickly by 1) reducing end-user confusion about whether 
it is necessary to analyze data with a (sometimes overwhelming) variety 
of computer programs; 2) establishing a common language upon which to 
compare different methods; and 3) providing a principled perspective 
from which to argue about the expected strengths and weaknesses of any 
method.

### 2014-03-28 tex/main-body-text.tex
expand introduction and start putting in basic structure

@@
 method.
+
+Methods for the unsupervised clustering of genotypes has received 
+considerable attention in the molecular ecology literature.  
+The best known example of this class of methods is {\sc structure} 
+\citep{Pritchardetal2000}.  {\sc structure} itself can be interpreted 
+as an elaboration of earlier models used to identify the spawning 
+stock of salmon \citep{Smouseetal1990}, and a variety of other 
+clustering methods have been developed that are all closely related to 
+{\sc structure}, for example {\sc NewHybrids} \citep{And&Tho2002}, {\sc 
+BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
+\citep{Coranderetal2004}. (An overview of these similarities can be 
+found in \citealt{Anderson2009PGAC})
+
+\section*{Comparison of methods}
+
+\subsection*{{\sc flock}}
+One issue I have with it is that it is described as a ``procedure'' or 
+an ``algorithm,'' without an attempt to see it as a special case of 
+well known inference frameworks.
+
+Some quotations to ponder.
+\begin{itemize}
+\item ``FLOCK is a non- Bayesian method and therefore differs 
+substantially from previous clustering algorithms''(p.~1333)
+
+\item ``However, its working principle does not involve a probabilistic 
+walk through the space of all possible k-partitions of sample S as is 
+the case with clustering algorithms such as Structure (Pritchard et al. 
+2000), BAPS (Corander et al. 2003) and NewHybrids (Anderson \& Thompson 
+2002).''(p.~1334).
+
+\item ``Partition (Dawson \& Belkhir 2001) is another clustering 
+program. However, since it is assuming Hardy–Weinberg and linkage 
+equilibrium, it is not suitable for the analysis of samples originating 
+from genetic groups in the process of becoming admixed.''(p.~1334).
+
+\item In \citet{Duc&Tur2012}: ``Note that FLOCK is very different from 
+other clustering 
+programs. It does not sample the space of partitions through small 
+random step walks as in MCMC, and it does not try to optimize some 
+target function, such as HWLE. Briefly stated, it is not based on a 
+probabilistic search algorithm. On the contrary, FLOCK is 
+entirely deterministic.''(p.~736).
+
+\end{itemize}
+
+\subsection*{{\sc structure}}
+
+\subsection*{{\sc flock} as a special case of {\sc structure}}
+
+\section*{Conclusions}

### 2014-03-28 tex/main-body-text.tex
more work on paper.  added directions to pdflatex it to README

--- 2014-03-28 tex/main-body-text.tex
@@
 novel, but rather because they are genuinely unaware of the close ties 
-between their work and existing methods.  Helping to identify the 
+between their work and existing methods.  Identifying the 
 similarities between new and existing work should help the field 
@@
 method.
+
+Methods for the unsupervised clustering of genotypes have received 
+considerable attention in the molecular ecology literature.  
+The best known example of this class of methods is {\sc structure} 
+\citep{Pritchardetal2000}.  {\sc structure} itself can be interpreted 
+as an elaboration of earlier models used to identify the spawning 
+stock of salmon \citep{Smouseetal1990}, and a variety of other 
+clustering methods have been developed that are all closely related to 
+{\sc structure}, for example {\sc NewHybrids} \citep{And&Tho2002}, {\sc 
+BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
+\citep{Coranderetal2004}. (An overview of these similarities can be 
+found in \citealt{Anderson2009PGAC}.)
+
+Recently, a description of the software {\sc flock} was published, 
+billing the method as a ``non-Bayesian method [that] differs substantially from previous 
+clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. A following paper asserts that
+\begin{quote}
+``{\sc flock} is very different from  other clustering 
+programs. It does not sample the space of partitions through small 
+random step walks as in MCMC, and it does not try to optimize some 
+target function, such as HWLE\@. Briefly stated, it is not based on a 
+probabilistic search algorithm. On the contrary, {\sc flock} is 
+entirely deterministic'' \citep[][p.~736]{Duc&Tur2012}.
+\end{quote}
+These papers describe the rationale behind the {\sc flock} algorithm in
+intuitive terms with analogies to the coalescence of flocks of birds
+or the accretion of snow on a moving snowball, \etc.
+
+Here, we provide a different perspective on the {\sc flock} algorithm
+showing it to be a special, restricted case of the simulated annealing algorithm for finding the Bayesian maximum {\em a posteriori} (MAP) estimate from a marginalized form of the {\sc structure} model with no admixture and non-correlated allele frequencies.
+
+\section*{Comparison of methods}
+We start with a succinct mathematical description of the {\sc flock} algorithm, and then\ldots
+
+\subsection*{{\sc flock}}
+One issue I have with it is that it is described as a ``procedure'' or 
+an ``algorithm,'' without an attempt to see it as a special case of 
+well known inference frameworks.
+
+\subsection*{{\sc structure}}
+
+\subsection*{{\sc flock} as a special case of {\sc structure}}
+
+\section*{Conclusions}

### 2014-03-31 tex/main-body-text.tex
minor edits during a brief read-through

@@
 The best known example of this class of methods is {\sc structure} 
-\citep{Pritchardetal2000}.  {\sc structure} itself can be interpreted 
+\citep{Pritchardetal2000}.  {\sc structure} itself can be described 
 as an elaboration of earlier models used to identify the spawning 
@@
 BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
-\citep{Coranderetal2004}. (An overview of these similarities can be 
-found in \citealt{Anderson2009PGAC}.)
+\citep{Coranderetal2004}. An overview of these similarities can be 
+found in \citet{Anderson2009PGAC}.
 
@@
 
-Here, we provide a different perspective on the {\sc flock} algorithm
+Here, we provide a different perspective on the {\sc flock} algorithm,
 showing it to be a special, restricted case of the simulated annealing algorithm for finding the Bayesian maximum {\em a posteriori} (MAP) estimate from a marginalized form of the {\sc structure} model with no admixture and non-correlated allele frequencies.
@@
 \section*{Comparison of methods}
-We start with a succinct mathematical description of the {\sc flock} algorithm, and then\ldots
+We start with a succinct mathematical description of the {\sc structure} algorithm, and then\ldots
+
+\subsection*{{\sc structure}}
 
@@
 
-\subsection*{{\sc structure}}
-
 \subsection*{{\sc flock} as a special case of {\sc structure}}

### 2014-04-21 tex/main-body-text.tex
Just added some line endings to clean it up a bit.
added some results still working on figures

@@
 Recently, a description of the software {\sc flock} was published, 
-billing the method as a ``non-Bayesian method [that] differs substantially from previous 
-clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. A following paper asserts that
+billing the method as a ``non-Bayesian method [that] 
+differs substantially from previous 
+clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. A following
+paper asserts that
 \begin{quote}
@@
 Here, we provide a different perspective on the {\sc flock} algorithm,
-showing it to be a special, restricted case of the simulated annealing algorithm for finding the Bayesian maximum {\em a posteriori} (MAP) estimate from a marginalized form of the {\sc structure} model with no admixture and non-correlated allele frequencies.
+showing it to be a special, restricted case of the simulated annealing
+algorithm for finding the Bayesian maximum {\em a posteriori} (MAP)
+estimate from a marginalized form of the {\sc structure} model with no
+admixture and non-correlated allele frequencies.
 
 \section*{Comparison of methods}
-We start with a succinct mathematical description of the {\sc structure} algorithm, and then\ldots
+We start with a succinct mathematical description of the {\sc structure}
+algorithm, and then\ldots
 
@@
 
+\section*{Results}
+
+\subsection*{{\sc flock} analysis of coasta California steelhead} 
+{\sc flock} was run four times using the default setting for all run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50, LLOD threshold = 0)} for values of \textit{k} from 2 to 6.
+The stopping point of \textit{k}=6 was chosen because four consecutive runs with 0 length plateaus were produced.
+Run times were relatively short.
+The four groups of runs for the default settings averaged 217 minutes (30min SD) on an i7 3.40GHz processor with 8GB of RAM.
+While no plateaus were observed in any of the four groups for a \textit{k} greater than two, the plateau record varied greatly among each group for \textit{k}=2 (plateau sequences- Run1=3,2,5,6,3; Run2=2,2,2,4,3,2; Run3=2,3,2,4; and Run4=2,3,3,4,2,2,5,2).
+Using the plateau analysis procedure suggested by \citet{Duc&Tur2012} either \textit{k}=2 is the lower bound estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4).
+Despite different conclusions, the allocation of individuals for the 'best run' from each group differed by a only a single individual.
+When \textit{k}=2, {\sc flock} identifies a boundary in the Russian Gulch area clustering individuals into northern and southern groups.
+With increased \textit{k} {\sc flock} splits individuals collected from the same stream between clusters.
+Increasing the number of iterations from 20 to 100 did not yield an increase in plateau length.
+
 \section*{Conclusions}
+The partition space over which {\sc flock} evaluates how individuals are clustered greatly increases with both the number of individuals and the number of clusters into which they can be partitioned.
+The total partition space for any given k can be defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the second kind.
+The results of {\sc flock} depend primarily on the initial partition of individuals.
+It is not surprising that with 2,596 individuals we could have drastically different plateau sequences with just two clusters.

### 2014-04-21 tex/main-body-text.tex
Lets try to push to my repo and then do a pull request
Add description of structure modeling and Gibbs samlping therein.
I have sort of just put that in there.  It could use some editing, no
doubt.

--- 2014-03-31 tex/main-body-text.tex
@@
 Here, we provide a different perspective on the {\sc flock} algorithm,
-showing it to be a special, restricted case of the simulated annealing algorithm for finding the Bayesian maximum {\em a posteriori} (MAP) estimate from a marginalized form of the {\sc structure} model with no admixture and non-correlated allele frequencies.
+showing it to be a special, restricted case of the simulated annealing
+algorithm for finding the Bayesian maximum {\em a posteriori} (MAP)
+estimate from a marginalized form of the {\sc structure} model with no
+admixture and non-correlated allele frequencies.
 
 \section*{Comparison of methods}
-We start with a succinct mathematical description of the {\sc structure} algorithm, and then\ldots
+We start with a succinct mathematical description of the {\sc structure}
+algorithm, then describe the {\sc flock} algorithm, and finally explain
+the close relationship between the two.
 
 \subsection*{{\sc structure}}
+In the {\sc structure} model with no admixture, the unknown ``subpopulation'' that the 
+$i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, where $K$ is the 
+number of subpopulations (or clusters, as they are often referred to).  
+In a diploid individual from cluster $k$, the 
+allelic types of the two gene copies at the $\ell\thh$ locus are assumed
+to be drawn independently from the vector of allele frequencies at
+locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where $A_\ell$ 
+denotes the number of alleles observed in the data set at locus $\ell$.
+Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the number
+of copies of each of the $A_\ell$ alleles at locus $j$ in individual $i$, and 
+$Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
+with $A_\ell$  components and cell probabilities given by the allele frequencies: 
+\begin{equation}
+(Y_{i\ell}~|~Z_i=k) \sim \mathrm{Mult}_{A_\ell}(2, \theta_{k\ell}).
+\end{equation}
+In the model without physical linkage, the genotypes at the loci are assumed to
+be independent of one another, so the probability of the genotype data at
+all $L$ loci---$Y_i=(Y_{i1},\ldots,Y_{iL})$---is simply a product of multinomial probabilities.
+In the uncorrelated allele frequencies model, the prior on each $\theta_{k\ell}$ is a
+Dirichlet distribution with parameters $(\lambda_{k\ell1},\ldots,\lambda_{k\ell A_\ell})$,
+which are usually set to a value like $1/A_\ell$.
+
+Inference in the model proceeds by sampling from the joint posterior of the 
+$Z_i$'s and $\theta = (\theta_1,\ldots,\theta_K)$ using Gibbs sampling.  That is,
+at iteration $t$:
+\begin{enumerate}
+\item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
+the full conditional distribution of $Z_i$ given
+$Y_i$ and $\theta^{(t)}$ (the current estimate of the allele frequencies).  This
+distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
+prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
+\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each cluster
+$k$, and locus $\ell$ the full conditional for $\theta_{k\ell}$ is independently a 
+Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for $j=1,\ldots, A_\ell$,
+where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
+individuals whose $Z^{(t+1)}_i = k$.   
+\end{enumerate}
+
+It will be useful for our comparison with {\sc flock} to point out that another way 
+of pursuing inference in this model would be to first integrate out the 
+Dirichlet priors on the allele frequencies and then sample from the posterior
+for the $Z_i$'s by Gibbs sampling.  When $\theta$ is integrated out, the genotypes 
+of the individuals are no longer conditionally independent (they were originally
+independent {\em conditional} on $\theta$ and the $Z_i$'s), so the calculation of 
+the full conditional distribution of $Z_i$ is now somewhat more involved: the full 
+conditional distribution of $Y_{i\ell}$ given $Z_i=k$, and $Z_{(-i)}$ and $Y_{(-i)\ell}$---the
+cluster memberships and genotypes, respectively,  of all the remaining individuals---is
+now a Compound Dirichlet Multinomial distribution (CDM):
+\begin{eqnarray}
+\lefteqn{(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell}) \sim} \label{eq:cdm}\\
+& & \mathrm{CDM}(\lambda_{k\ell 1} + \#_{(-i)}(k,\ell,1), \ldots,
+\lambda_{k\ell A_\ell} + \#_{(-i)}(k,\ell,A\ell)), \nonumber
+\end{eqnarray}
+where $\#_{(-i)}(k,\ell,1)$ is the number of alleles of type $j$ at locus $\ell$
+found in individuals {\em other than individual $i$} that currently
+belong to cluster $k$. Calculating (\ref{eq:cdm}) for each value of $Z_i=k \in \{1,\ldots,K\}$ 
+and normalizing to sum to one gives the full conditional distribution for 
+$Z_i$:
+\begin{equation}
+P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})~~,~~k=1,\ldots,K.
+\label{eq:fc}
+\end{equation}
+A new value of $Z_i$ would be drawn from (\ref{eq:fc}) if doing Gibbs sampling in this
+version of the {\sc structure} model in which $\theta$ has been integrated out.  In fact,
+this is the approach (with a slightly different prior weight on the $Z_i$'s) taken to update 
+$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} when not
+proposing changes to the number of clusters.
 

### 2014-04-22 tex/main-body-text.tex
Added some results and a figure under images in a Figures-Pat folder

@@
 While no plateaus were observed in any of the four groups for a \textit{k} greater than two, the plateau record varied greatly among each group for \textit{k}=2 (plateau sequences- Run1=3,2,5,6,3; Run2=2,2,2,4,3,2; Run3=2,3,2,4; and Run4=2,3,3,4,2,2,5,2).
-Using the plateau analysis procedure suggested by \citet{Duc&Tur2012} either \textit{k}=2 is the lower bound estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4).
+The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to two different conclusions for the plateau sequences observed: \textit{k}=2 is the lower bound estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4).
 Despite different conclusions, the allocation of individuals for the 'best run' from each group differed by a only a single individual.
-When \textit{k}=2, {\sc flock} identifies a boundary in the Russian Gulch area clustering individuals into northern and southern groups.
-With increased \textit{k} {\sc flock} splits individuals collected from the same stream between clusters.
-Increasing the number of iterations from 20 to 100 did not yield an increase in plateau length.
+When \textit{k}=2, {\sc flock} identifies a boundary in the Russian Gulch area clustering individuals into northern and southern groups (Fig.
+1).
+
+\begin{figure}
+    \includegraphics[width=8.0in]{FLOCK_OneSampDefault.tif}
+    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created in the program {\sc flock}.
+Each horizontal plot represents the 'best run' from a series where individuals are reallocated into \textit{k} clusters.
+Within each plot the normalized individual likelihood for each reference population is represented by a colored vertical bar.
+Major geographic boundaries that were attributed to genetic structure in \citet{Garzaetal_norcal} are indicated with black vertical lines.
+Results from the 'best runs' showed similar geographic differentiation to \citet{Garzaetal_norcal} analysis.
+The plateau analysis developed by \citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 is the lower bound estimate.}
+    \label{Fig.1}
+\end{figure}
+
+With increased \textit{k} {\sc flock} continues to splits individuals approximately based on geographical boundaries identified by \citet{Garzaetal_norcal} using {\sc structure}.
+Once \textit{k} = 6 is reached no new regions are identified.
+
+To investigate the ability of the arrangement of individuals within the reference populations to move around the partition space we varied the number of iterations that each run goes through to find the 'optimal' allocation of individuals.
+Unfortuantely only the reallocation matrix for the 'best run' is given as output.
+Increasing the number of iterations did not affect the overall inference of \textit{k}.
+It was observed that often during the last iterations individuals would alternate between reference groups during subsequent iterations.
+Increasing the log likelihood difference threshold for reallocation to 0.18 (there needs to be at least a 1.5 times higher likelihood to reallocate the individual) minimized this, but did not affect the inference of \textit{k} (lower bound estimate of \textit{k} = 2).
 
@@
 It is not surprising that with 2,596 individuals we could have drastically different plateau sequences with just two clusters.
+It is difficult to access how well the reallocation process moves the cluster of individuals within the partition space due to the fact that the reallocation matrix from only the 'best-run' is reported.

### 2014-04-22 tex/main-body-text.tex
Made figure a pdf and made the extension work - formatting help needed
Minor changes to figures and references.
Hard wrap lines in tex document.

@@
 In the {\sc structure} model with no admixture, the unknown ``subpopulation'' that the 
-$i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, where $K$ is the 
+$i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, 
+where $K$ is the 
 number of subpopulations (or clusters, as they are often referred to).  
@@
 to be drawn independently from the vector of allele frequencies at
-locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where $A_\ell$ 
+locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where 
+$A_\ell$ 
 denotes the number of alleles observed in the data set at locus $\ell$.
-Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the number
+Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the 
+number
 of copies of each of the $A_\ell$ alleles at locus $j$ in individual $i$, and 
@@
 be independent of one another, so the probability of the genotype data at
-all $L$ loci---$Y_i=(Y_{i1},\ldots,Y_{iL})$---is simply a product of multinomial probabilities.
+all $L$ loci---$Y_i=(Y_{i1},\ldots,Y_{iL})$---is simply a product of multinomial 
+probabilities.
 In the uncorrelated allele frequencies model, the prior on each $\theta_{k\ell}$ is a
-Dirichlet distribution with parameters $(\lambda_{k\ell1},\ldots,\lambda_{k\ell A_\ell})$,
+Dirichlet distribution with parameters $(\lambda_{k\ell1},\ldots,\lambda_{k\ell A_
+\ell})$,
 which are usually set to a value like $1/A_\ell$.
@@
 prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
-\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each cluster
+\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each 
+cluster
 $k$, and locus $\ell$ the full conditional for $\theta_{k\ell}$ is independently a 
-Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for $j=1,\ldots, A_\ell$,
+Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
+$j=1,\ldots, A_\ell$,
 where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
@@
 the full conditional distribution of $Z_i$ is now somewhat more involved: the full 
-conditional distribution of $Y_{i\ell}$ given $Z_i=k$, and $Z_{(-i)}$ and $Y_{(-i)\ell}$---the
-cluster memberships and genotypes, respectively,  of all the remaining individuals---is
+conditional distribution of $Y_{i\ell}$ given $Z_i=k$, and $Z_{(-i)}$ and $Y_{(-i)
+\ell}$---the
+cluster memberships and genotypes, respectively,  of all the remaining individuals---
+is
 now a Compound Dirichlet Multinomial distribution (CDM):
@@
 found in individuals {\em other than individual $i$} that currently
-belong to cluster $k$. Calculating (\ref{eq:cdm}) for each value of $Z_i=k \in \{1,\ldots,K\}$ 
+belong to cluster $k$. Calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
+\{1,\ldots,K\}$ 
 and normalizing to sum to one gives the full conditional distribution for 
@@
 A new value of $Z_i$ would be drawn from (\ref{eq:fc}) if doing Gibbs sampling in this
-version of the {\sc structure} model in which $\theta$ has been integrated out.  In fact,
-this is the approach (with a slightly different prior weight on the $Z_i$'s) taken to update 
-$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} when not
+version of the {\sc structure} model in which $\theta$ has been integrated out.  In 
+fact,
+this is the approach (with a slightly different prior weight on the $Z_i$'s) taken to 
+update 
+$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} 
+when not
 proposing changes to the number of clusters.
@@
 \subsection*{{\sc flock} analysis of coasta California steelhead} 
-{\sc flock} was run four times using the default setting for all run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50, LLOD threshold = 0)} for values of \textit{k} from 2 to 6.
-The stopping point of \textit{k}=6 was chosen because four consecutive runs with 0 length plateaus were produced.
-Run times were relatively short.
-The four groups of runs for the default settings averaged 217 minutes (30min SD) on an i7 3.40GHz processor with 8GB of RAM.
-While no plateaus were observed in any of the four groups for a \textit{k} greater than two, the plateau record varied greatly among each group for \textit{k}=2 (plateau sequences- Run1=3,2,5,6,3; Run2=2,2,2,4,3,2; Run3=2,3,2,4; and Run4=2,3,3,4,2,2,5,2).
-The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to two different conclusions for the plateau sequences observed: \textit{k}=2 is the lower bound estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4).
-Despite different conclusions, the allocation of individuals for the 'best run' from each group differed by a only a single individual.
-When \textit{k}=2, {\sc flock} identifies a boundary in the Russian Gulch area clustering individuals into northern and southern groups (Fig.
-1).
+{\sc flock} was run four times using the default setting for all run parameters 
+\textit{(initial partition = random, number of iterations = 20, number of runs = 50, 
+LLOD threshold = 0)} for values of \textit{k} from 2 to 6. The stopping point of 
+\textit{k}=6 was chosen because four consecutive runs with 0 length plateaus were 
+produced. 
+Run times were relatively short. The four groups of runs for the default settings 
+averaged 217 minutes (30min SD) on an i7 3.40GHz processor with 8GB of RAM. While no 
+plateaus were observed in any of the four groups for a \textit{k} greater than two, 
+the plateau record varied greatly among each group for \textit{k}=2 (plateau 
+sequences- Run1=3,2,5,6,3; Run2=2,2,2,4,3,2; Run3=2,3,2,4; and Run4=2,3,3,4,2,2,5,2). 
+The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to two different 
+conclusions for the plateau sequences observed: \textit{k}=2 is the lower bound 
+estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4). Despite 
+different conclusions, the allocation of individuals for the 'best run' from each 
+group differed by a only a single individual. When \textit{k}=2, {\sc flock} 
+identifies a boundary in the Russian Gulch area clustering individuals into northern 
+and southern groups (Fig. 1).
 
-\begin{figure}
-    \includegraphics[width=8.0in]{FLOCK_OneSampDefault.tif}
-    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created in the program {\sc flock}.
-Each horizontal plot represents the 'best run' from a series where individuals are reallocated into \textit{k} clusters.
-Within each plot the normalized individual likelihood for each reference population is represented by a colored vertical bar.
-Major geographic boundaries that were attributed to genetic structure in \citet{Garzaetal_norcal} are indicated with black vertical lines.
-Results from the 'best runs' showed similar geographic differentiation to \citet{Garzaetal_norcal} analysis.
-The plateau analysis developed by \citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 is the lower bound estimate.}
+\begin{figure*}
+\begin{center}
+    \includegraphics[width=\textwidth]{images/Figures-Pat/FLOCK_OneSampDefault.pdf}
+    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 
+in the program {\sc flock}. Each horizontal plot represents the 'best run' from a 
+series where individuals are reallocated into \textit{k} clusters. Within each plot 
+the normalized individual likelihood for each reference population is represented by a 
+colored vertical bar. Major geographic boundaries that were attributed to genetic 
+structure in \citet{Garzaetal_norcal} are indicated with black vertical lines. Results 
+from the 'best runs' showed similar geographic differentiation to 
+\citet{Garzaetal_norcal} analysis. The plateau analysis developed by 
+\citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 
+is the lower bound estimate.}
     \label{Fig.1}
-\end{figure}
+\end{center}
+\end{figure*}
 
-With increased \textit{k} {\sc flock} continues to splits individuals approximately based on geographical boundaries identified by \citet{Garzaetal_norcal} using {\sc structure}.
-Once \textit{k} = 6 is reached no new regions are identified.
+With increased \textit{k} {\sc flock} continues to splits individuals approximately 
+based on geographical boundaries identified by \citet{Garzaetal_norcal} using {\sc 
+structure}. Once \textit{k} = 6 is reached no new regions are identified.
 
-To investigate the ability of the arrangement of individuals within the reference populations to move around the partition space we varied the number of iterations that each run goes through to find the 'optimal' allocation of individuals.
-Unfortuantely only the reallocation matrix for the 'best run' is given as output.
-Increasing the number of iterations did not affect the overall inference of \textit{k}.
-It was observed that often during the last iterations individuals would alternate between reference groups during subsequent iterations.
-Increasing the log likelihood difference threshold for reallocation to 0.18 (there needs to be at least a 1.5 times higher likelihood to reallocate the individual) minimized this, but did not affect the inference of \textit{k} (lower bound estimate of \textit{k} = 2).
+To investigate the ability of the arrangement of individuals within the reference 
+populations to move around the partition space we varied the number of iterations that 
+each run goes through to find the 'optimal' allocation of individuals. Unfortuantely 
+only the reallocation matrix for the 'best run' is given as output. Increasing the 
+number of iterations did not affect the overall inference of \textit{k}. It was 
+observed that often during the last iterations individuals would alternate between 
+reference groups during subsequent iterations. Increasing the log likelihood 
+difference threshold for reallocation to 0.18 (there needs to be at least a 1.5 times 
+higher likelihood to reallocate the individual) minimized this, but did not affect the 
+inference of \textit{k} (lower bound estimate of \textit{k} = 2).
 
 \section*{Conclusions}
-The partition space over which {\sc flock} evaluates how individuals are clustered greatly increases with both the number of individuals and the number of clusters into which they can be partitioned.
-The total partition space for any given k can be defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the second kind.
-The results of {\sc flock} depend primarily on the initial partition of individuals.
-It is not surprising that with 2,596 individuals we could have drastically different plateau sequences with just two clusters.
-It is difficult to access how well the reallocation process moves the cluster of individuals within the partition space due to the fact that the reallocation matrix from only the 'best-run' is reported.
+The partition space over which {\sc flock} evaluates how individuals are clustered 
+greatly increases with both the number of individuals and the number of clusters into 
+which they can be partitioned. The total partition space for any given k can be 
+defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
+second kind. The results of {\sc flock} depend primarily on the initial partition of 
+individuals. It is not surprising that with 2,596 individuals we could have 
+drastically different plateau sequences with just two clusters. It is difficult to 
+access how well the reallocation process moves the cluster of individuals within the 
+partition space due to the fact that the reallocation matrix from only the 'best-run' 
+is reported.

### 2014-05-06 tex/main-body-text.tex
Minor edits. Started writing flock description.

@@
 $Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
-with $A_\ell$  components and cell probabilities given by the allele frequencies: 
+$A_\ell$  components and cell probabilities given by the allele frequencies: 
 \begin{equation}
@@
 cluster
-$k$, and locus $\ell$ the full conditional for $\theta_{k\ell}$ is independently a 
+$k$ and locus $\ell$, the full conditional for $\theta_{k\ell}$ is independently a 
 Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
@@
 where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
-individuals whose $Z^{(t+1)}_i = k$.   
+individuals with $Z^{(t+1)}_i = k$.   
 \end{enumerate}
@@
 independent {\em conditional} on $\theta$ and the $Z_i$'s), so the calculation of 
-the full conditional distribution of $Z_i$ is now somewhat more involved: the full 
-conditional distribution of $Y_{i\ell}$ given $Z_i=k$, and $Z_{(-i)}$ and $Y_{(-i)
-\ell}$---the
-cluster memberships and genotypes, respectively,  of all the remaining individuals---
-is
-now a Compound Dirichlet Multinomial distribution (CDM):
+the full conditional distribution of $Z_i$ becomes somewhat more involved,
+but can be computed by the following reasoning. First, the full 
+conditional distribution of $Y_{i\ell}$, given that individual $i$ is from subpopulation $k$
+(\ie $Z_i=k$) depends on the 
+cluster memberships and genotypes  of all the remaining individuals,
+which we denote by  $Z_{(-i)}$ and $Y_{(-i)\ell}$, respectively.
+This full conditional for $Y_{i\ell}$ follows 
+a Compound Dirichlet Multinomial distribution (CDM):
 \begin{eqnarray}
@@
 found in individuals {\em other than individual $i$} that currently
-belong to cluster $k$. Calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
+belong to cluster $k$. Thus, calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
 \{1,\ldots,K\}$ 
@@
 \subsection*{{\sc flock}}
+Here we translate the recipe given by \citeauthor{Duc&Tur2009} for the 
+{\sc flock} algorithm into a concise specification in terms of the variables
+defined in the previous section.  We provide the description for case when
+it is assumed there are $K$ clusters.
+
+The first step of the flock algorithm is initialization.  [Now echo what the
+authors say about this, and then translate that into an equation on $Z_i$]
+
 One issue I have with it is that it is described as a ``procedure'' or 

### 2014-05-09 tex/main-body-text.tex
Revised figures added
Added references, made some slight changes to results section

@@
 
-\section*{Results}
+\section*{Performance Comparison with {\sc structure}}
 
-\subsection*{{\sc flock} analysis of coasta California steelhead} 
-{\sc flock} was run four times using the default setting for all run parameters 
+\subsection*{Analysis of coasta California steelhead} 
+To evaluate the preformance and computational efficiency of {\sc flock}  we used a published 
+dataset from \citet{Garzaetal_norcal} comprised of 2596 steelhead \textit{Oncorhynchus mykiss}
+genotyped at 15 microsatellite loci.  {\sc flock} was run six times using the default setting for all run parameters 
 \textit{(initial partition = random, number of iterations = 20, number of runs = 50, 
@@
 \textit{k}=6 was chosen because four consecutive runs with 0 length plateaus were 
-produced. 
+produced. The non-admixture, uncorrelated allele frequency model implemented in {\sc structure}
+was run with six iterations of a 50,000 sweep burn-in followed by 150,000 sweep sample 
+from the posterior of \textit{k} from 2 to 6. To facilitate the comparison of the two programs,
+the program {\sc clump} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
+while the program Distruct \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+
+\subsection*{Results} 
 Run times were relatively short. The four groups of runs for the default settings 
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/FLOCK_OneSampDefault.pdf}
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}
     \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 

### 2014-05-09 tex/main-body-text.tex
Some minor additions to flock method description.

--- 2014-05-06 tex/main-body-text.tex
@@
 {\sc flock} algorithm into a concise specification in terms of the variables
-defined in the previous section.  We provide the description for case when
+defined in the previous section.  We provide the description for a case in which
 it is assumed there are $K$ clusters.
 
-The first step of the flock algorithm is initialization.  [Now echo what the
-authors say about this, and then translate that into an equation on $Z_i$]
+The first step of the flock algorithm is initialization, during which 
+the program randomly allocates each of the $N$ individuals to one of 
+$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
+these steps, each individual is given the chance to be reallocated 
+(\ie moved to a different cluster).  This is done on the basis of 
+maximum likelihood:
 
@@
 
-\subsection*{{\sc flock} analysis of coasta California steelhead} 
+\subsection*{{\sc flock} analysis of coastal California steelhead} 
 {\sc flock} was run four times using the default setting for all run parameters 

### 2014-05-09 tex/main-body-text.tex

--- 2014-05-09 tex/main-body-text.tex
@@
 
-\subsection*{Analysis of coasta California steelhead} 
+\subsection*{Analysis of coastal California steelhead} 
 To evaluate the preformance and computational efficiency of {\sc flock}  we used a published 
@@
 genotyped at 15 microsatellite loci.  {\sc flock} was run six times using the default setting for all run parameters
+
 \textit{(initial partition = random, number of iterations = 20, number of runs = 50, 

### 2014-05-13 tex/main-body-text.tex
Write technical description of flock
Also some minor edits to eric's earlier sections
and I made a jpg out of Patrick's big figure for
previewing, just because my computer can't handle the big
PDF file elegantly enough to have it there during
editing.

@@
 \end{equation}
-In the model without physical linkage, the genotypes at the loci are assumed to
+In the {\sc structure} model without physical linkage, the genotypes at the loci are assumed to
 be independent of one another, so the probability of the genotype data at
@@
 
-Inference in the model proceeds by sampling from the joint posterior of the 
+After initializing the unknown allele frequencies to randomly drawn values,
+$\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
 $Z_i$'s and $\theta = (\theta_1,\ldots,\theta_K)$ using Gibbs sampling.  That is,
-at iteration $t$:
+at iteration $t = 0, 1, 2, \ldots$:
 \begin{enumerate}
@@
 conditional distribution of $Y_{i\ell}$, given that individual $i$ is from subpopulation $k$
-(\ie $Z_i=k$) depends on the 
+(\ie $Z_i=k$), depends on the 
 cluster memberships and genotypes  of all the remaining individuals,
@@
 (\ie moved to a different cluster).  This is done on the basis of 
-maximum likelihood:
+maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
+performed following multilocus maximum likelihood (Paetkau et al. 1995).''
 
-One issue I have with it is that it is described as a ``procedure'' or 
-an ``algorithm,'' without an attempt to see it as a special case of 
-well known inference frameworks.
+From that description, it is not clear how \citeauthor{Duc&Tur2009} treat 
+alleles that appear in an
+individual but do not appear within any other individuals within a cluster.
+The original approach of \citet{Paetkauetal1995} merely added 0.01 or some other
+small value to each allele frequency that was 0, and then renormalized 
+the allele frequencies to sum to 1.  In later works, Paetkau and colleagues
+also employed the ``Bayesian'' approach of \citet{Ran&Mou1997} in their 
+software {\sc geneclass2}  \citep{Piryetal2004}.  If \citeauthor{Duc&Tur2009}
+use the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
+be reallocated to whichever cluster gives the highest value to 
+$P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$, which is exactly the probability
+defined in (\ref{eq:cdm}).  
+If \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
+formulation of \citet{Paetkauetal1995}, then their reallocations are based
+on maximizing a likelihood that, although it may not be formally identical to
+that in (\ref{eq:cdm}), is not substantively different from it.
+
+It is worth pointing out that, since the
+{\sc structure} model without admixture assumes a uniform prior over 
+the $K$ different clusters, $P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$
+is exactly proportional to $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$, which 
+we have shown is the
+full conditional distribution for $Z_i$ in the {\sc structure} model after
+integrating out $\theta$.  Thus, we have shown that, doing Gibbs
+sampling for $Z_i$ in  the {\sc structure} in which $\theta$ has been
+integrated out would involve sampling a new value of $Z_i$ from the 
+probability distribution,
+\[
+P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
+\]
+and that the updates that
+{\sc flock} makes to each $Z_i$ involve assigning to $Z_i$
+\begin{equation}
+\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
+\end{equation}
+\ie, the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$
+(or, as stated above, a very similar function of $k$).
 
 \subsection*{{\sc flock} as a special case of {\sc structure}}
+The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
+$Z$'s in proportion to their posterior probability.  If, all that was desired
+was a point estimate of the $Z_i$'s that maximized the posterior probability,
+then one standard approach to finding that maximimum {\em a posteriori}, or MAP, 
+estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
+To do simulated annealing in the {\sc structure} model with no admixture and
+no correlated allele frequency prior, after integrating out the unknown
+allele frequencies, the updates for each $Z_i$ would be made from the 
+distribution
+\[
+\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})\biggr]^\beta
+\]
+which is the same full conditional distribution as before, raised
+to the power $\beta$.  In typical applications of simulated annealing,
+$\beta$ is initially set to a low value $<1$ and as the algorithm proceeds
+$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
+the algorithm finds a local maximum in the posterior probability surface.  
+When $\beta$ is small, the process can easily traverse ``valleys''
+in the posterior probability surface and have a better chance of searching
+more of the space for the maximum value.  As $\beta$ gets larger, the process
+tends to move uphill on the probability surface, searching for the
+maximum near where it currently is.
+
+It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
+algorithm for the {\sc structure} model in which the cooling schedule is
+``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
+the algorithm.''  As such, we can expect that it may not provide the best 
+mechanism for searching for the MAP estimate and it may be more susceptible 
+to becoming trapped in local modes of the posterior probability. 
+In the following section we show that {\sc structure} and {\sc flock} find similar
+solutions in the analysis of a large microsatellite data set, though, as expected,
+{\sc flock} seems to find the same solution less consistently than does {\sc structure}.
+
+\subsection*{boneyard}
+ On the other hand, in practice, the 
+posterior probability surface may have several modes that are separated by regions
+of low posterior probability, such that the solution
+an algorithm converges to (whether it is {\sc structure}  doing MCMC or {\sc flock} doing 
+``very cold simulated annealing'') depends almost entirely on the initial random
+starting values.
 
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}
+	 % NOTE. Eric made Flock-Fig1.jpg as a Preview because Flock-Fig1.pdf
+	 % ate up too many resources while editing the document.  For final 
+	 % production, it should be changed back to Flock-Fig1.pdf
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.jpg}  
     \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 

### 2014-05-13 tex/main-body-text.tex
added q-val plot & some results

--- 2014-05-09 tex/main-body-text.tex
@@
 dataset from \citet{Garzaetal_norcal} comprised of 2596 steelhead \textit{Oncorhynchus mykiss}
-genotyped at 15 microsatellite loci.  {\sc flock} was run six times using the default setting for all run parameters
-
-\textit{(initial partition = random, number of iterations = 20, number of runs = 50, 
+genotyped at 15 microsatellite loci.
+{\sc flock} was run six times using the default setting for all run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50,
 LLOD threshold = 0)} for values of \textit{k} from 2 to 6. The stopping point of 
-\textit{k}=6 was chosen because four consecutive runs with 0 length plateaus were 
-produced. The non-admixture, uncorrelated allele frequency model implemented in {\sc structure}
-was run with six iterations of a 50,000 sweep burn-in followed by 150,000 sweep sample 
-from the posterior of \textit{k} from 2 to 6. To facilitate the comparison of the two programs,
+\textit{k}=6 was chosen after four consecutive runs yielded no plateau.
+The non-admixture, uncorrelated allele frequency model implemented in {\sc structure} was run with six iterations of a 50,000 sweep burn-in followed by 150,000 sweep sample from the posterior of \textit{k} from 2 to 6.
+All runs were performed with an i7 3.40GHz processor and 8GB of RAM.
+To facilitate the comparison of the two programs,
 the program {\sc clump} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
@@
 \subsection*{Results} 
-Run times were relatively short. The four groups of runs for the default settings 
-averaged 217 minutes (30min SD) on an i7 3.40GHz processor with 8GB of RAM. While no 
-plateaus were observed in any of the four groups for a \textit{k} greater than two, 
-the plateau record varied greatly among each group for \textit{k}=2 (plateau 
-sequences- Run1=3,2,5,6,3; Run2=2,2,2,4,3,2; Run3=2,3,2,4; and Run4=2,3,3,4,2,2,5,2). 
-The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to two different 
-conclusions for the plateau sequences observed: \textit{k}=2 is the lower bound 
-estimate (Run1) or there is no estimate (\textit{undecided}: Runs 2-4). Despite 
-different conclusions, the allocation of individuals for the 'best run' from each 
-group differed by a only a single individual. When \textit{k}=2, {\sc flock} 
-identifies a boundary in the Russian Gulch area clustering individuals into northern 
-and southern groups (Fig. 1).
+Run times were relatively short. The six groups of runs for the default settings 
+averaged 188.8 minutes (1.5 min SD).
+While no  plateaus were observed in any of the six groups for a \textit{k} greater than two, the plateau record varied greatly among each group for \textit{k}=2 (plateau
+sequences- Run1= 3,2,3,5,2,4; Run2=2,5,2,3,2,3; Run3=2,2,2,2,2,5; Run4=2,3,4,2,2,2,3; Run5=2,2,2,4,3,2; 
+and; Run6=2,3,3,4,2,2,5,2). The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to 
+an \textit{undecided} estimate of K as a consequence of no plateau lengths being greater than 6. Despite 
+the K estimation based on rules of thumb devloped by \citet{Duc&Tur2012}, Distruct plots for each group of runs 
+for \textit{k}=2-4 lead to nearly identical clustering of individuals (See supplementary materials).
+While only two of the six 'best runs' when \textit{k}=2 ended with identical clustering (same mean LLOD score), all other runs differed by at most the allocation of 13 individuals.
+When \textit{k} was set to three and four all runs provided very similar clustering.
 
@@
 based on geographical boundaries identified by \citet{Garzaetal_norcal} using {\sc 
-structure}. Once \textit{k} = 6 is reached no new regions are identified.
+structure}.
+
+\begin{figure*}
+\begin{center}
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-StrFloc.pdf}
+    \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. {\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty attributed to them.}
+    \label{Fig.2}
+\end{center}
+\end{figure*}
 

### 2014-05-13 tex/main-body-text.tex
minor results added

@@
 dataset from \citet{Garzaetal_norcal} comprised of 2596 steelhead \textit{Oncorhynchus mykiss}
-genotyped at 15 microsatellite loci.
-{\sc flock} was run six times using the default setting for all run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50,
+genotyped at 15 microsatellite loci.  {\sc flock} was run six times using the default setting for all 
+run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50, 
 LLOD threshold = 0)} for values of \textit{k} from 2 to 6. The stopping point of 
-\textit{k}=6 was chosen after four consecutive runs yielded no plateau.
-The non-admixture, uncorrelated allele frequency model implemented in {\sc structure} was run with six iterations of a 50,000 sweep burn-in followed by 150,000 sweep sample from the posterior of \textit{k} from 2 to 6.
-All runs were performed with an i7 3.40GHz processor and 8GB of RAM.
-To facilitate the comparison of the two programs,
-the program {\sc clump} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
+\textit{k}=6 was chosen after four consecutive runs yielded no plateau. The non-admixture, 
+uncorrelated allele frequency model implemented in {\sc structure} was run with six iterations of a 
+50,000 sweep burn-in followed by 150,000 sweep sample from the posterior of \textit{k} from 2 to 6. 
+All runs were performed with an i7 3.40GHz processor and 8GB of RAM. To facilitate the comparison of
+ the two programs, the program {\sc clump} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
 while the program Distruct \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
@@
 \subsection*{Results} 
-Run times were relatively short. The six groups of runs for the default settings 
-averaged 188.8 minutes (1.5 min SD).
-While no  plateaus were observed in any of the six groups for a \textit{k} greater than two, the plateau record varied greatly among each group for \textit{k}=2 (plateau
-sequences- Run1= 3,2,3,5,2,4; Run2=2,5,2,3,2,3; Run3=2,2,2,2,2,5; Run4=2,3,4,2,2,2,3; Run5=2,2,2,4,3,2; 
-and; Run6=2,3,3,4,2,2,5,2). The plateau analysis procedure suggested by \citet{Duc&Tur2012} leads to 
-an \textit{undecided} estimate of K as a consequence of no plateau lengths being greater than 6. Despite 
-the K estimation based on rules of thumb devloped by \citet{Duc&Tur2012}, Distruct plots for each group of runs 
-for \textit{k}=2-4 lead to nearly identical clustering of individuals (See supplementary materials).
-While only two of the six 'best runs' when \textit{k}=2 ended with identical clustering (same mean LLOD score), all other runs differed by at most the allocation of 13 individuals.
-When \textit{k} was set to three and four all runs provided very similar clustering.
+Run times were relatively short for both programs. The six groups of runs for {\sc flock} 
+using the default settings required 1133 minutes to complete. Each individual group  of runs 
+averaged 188.8 minutes (1.5 min SD). The six iterations of {\sc structure} completed in 
+XXXX minutes with each iterations averaging XXXX (SD).
+
+While no  plateaus were observed in any of the six groups for a \textit{k} greater than 
+two, the plateau record varied greatly among each
+ group for \textit{k}=2 (plateau  sequences- Run1= 3,2,3,7,4; Run2=2,5,2,3,2,3; Run3=2,2,2,2,2,5; 
+Run4=2,3,4,2,4,3; Run5=2,2,5,5,3; and; Run6=2,2,2,2,2,3,3). The plateau analysis 
+procedure suggested by \citet{Duc&Tur2012} leads to 
+either an estimate of K$\geq$2 or \textit{undecided} as a consequence of no plateau 
+lengths being greater than 6. Distruct plots for each group of runs for \textit{k}=2-4 
+lead to nearly identical clustering of individuals (See supplementary materials). While 
+only two of the six 'best runs' when 
+\textit{k}=2 ended with identical clustering (same mean LLOD score), all other runs 
+differed by at most the allocation of 13 individuals. When \textit{k} was set to three 
+and four all runs provided very similar clustering.
 
@@
     \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-StrFloc.pdf}
-    \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. {\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty attributed to them.}
+    \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 
+{\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
+ attributed to them.}
     \label{Fig.2}

### 2014-05-15 ForPipeline/Flock2Structure.R
Added code for analysis located in ForPipeline. If you open from Rstudio wd should be set and you can strip the flock results from their workbooks and easily paste into the structure output file

@@ -1,0 +1,263 @@
+######Code for Converting Flock pop Q values to Structure formated pop - Q values
+######Written for Flock Comment with Eric Anderson
+
+###### Note: individuals parts of the code rely on Java and should be run individually 
+###### If they are all run together you will probably experience a heap error. 
+
+### packages that we will need that probably aren't installed
+install.packages("XLConnect")
+install.packages("stringr")
+
+### 
+require("XLConnect")
+require("stringr")
+
+#We are going to work in a folder called 'ForPipeline'
+#So long as the R scrip launches from this point the working directory 
+#should be whereever that script is stored. 
+
+#first let's make sure the wd is what we expect, change it if not
+wd<-getwd()
+
+#Where are flock results stored?
+FOR<-"Flock-output-Raw"
+#Note: we are going to do a lot of importing of different runs. 
+# The generic naming of flock output is 'k2 Steelhead_Run1_Defaults.xls'
+# where k and Run # will change
+
+#how many clusters?
+k.max=6
+#how many reps did we do?
+rep=6
+
+# First need to save population based allocation of individuals page in flock output as 
+# a .csv with only one header line!
+# Flock saves the population allocation information in the sheet labelled "Samples allocation matrix"
+options(java.parameters = "-Xmx4g" )
+
+#read in the excel file and output a .csv 
+#Writing all these .csv files may be time intensive, but 
+#I already wrote the code to process the .csv file, so let's just
+#live with this inefficency.
+for (k in 2:k.max){
+for (r in 1:rep){
+excel.file <- file.path(FOR,paste("k",k," Steelhead_Run",r,"_Defaults.xls",sep=""))
+assign(x="temp",value=(readWorksheetFromFile(excel.file, sheet="Samples allocation matrix", header=TRUE)))
+temp<-temp[-(1:3),]
+write.csv(x=temp,file=file.path(FOR,paste("k",k," Steelhead_Run",r,"_Defaults-popq.csv",sep="")),row.names=F)
+}#over r reps
+}#over k clusters
+
+########################################################################################
[…]

### 2014-05-18 tex/main-body-text.tex
Combined all Q-val plots into a single axis
Edited the results section and added Fig1 in Grey Scale

@@
 Run times were relatively short for both programs. The six groups of runs for {\sc flock} 
-using the default settings required 1133 minutes to complete. Each individual group  of runs 
+required 1133 minutes to complete. Each individual group of runs 
 averaged 188.8 minutes (1.5 min SD). The six iterations of {\sc structure} completed in 
-XXXX minutes with each iterations averaging XXXX (SD).
+486 minutes with each of the six iterations averaging 80.8 minutes (0.8 min SD). Increasing
+the number of putative clusters from \textit{k} of two to six yielded a 2.4 fold increase 
+in processing time for {\sc flock}, whereas a 1.6 fold increase was observed for runs of 
+{\sc structure}.
 
-While no  plateaus were observed in any of the six groups for a \textit{k} greater than 
-two, the plateau record varied greatly among each
+The allocation of indivividuals into K clusters is congruent for each program. For all six groups 
+of runs from \textit{k}=2-4, {\sc flock} lead to nearly identical clustering of individuals 
+(See supplementary materials). It should be noted
+that of the 50 starting allocations of individuals in the {\sc flock} procedure only the LLOD for the 
+ 'best run' is reported. Of these 'best runs' for a \textit{k} of 2 only two  
+ended with identical clustering (same mean LLOD score), all other runs 
+differed by at most the allocation of 13 individuals. As \textit{k} increases so do the number of solutions
+that {\sc flock} finds (Fig 1). These solutions, however, are suprisingly similar to
+ the partions identified by {\sc structure}. Both programs idenfity clusters 
+which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal}.
+
+Despite the similar allocation of individuals into \textit{k} clusters, the 
+plateau analysis in {\sc flock} leads to different conclusions among groups of runs, as well as 
+one that differs from both the pure likelihood 
+and delta K approach in {\sc structure}. While no plateaus were observed in any of the six groups 
+for a \textit{k} $\geq$2, the plateau record varied greatly among each
  group for \textit{k}=2 (plateau  sequences- Run1= 3,2,3,7,4; Run2=2,5,2,3,2,3; Run3=2,2,2,2,2,5; 
@@
 procedure suggested by \citet{Duc&Tur2012} leads to 
-either an estimate of K$\geq$2 or \textit{undecided} as a consequence of no plateau 
-lengths being greater than 6. Distruct plots for each group of runs for \textit{k}=2-4 
-lead to nearly identical clustering of individuals (See supplementary materials). While 
-only two of the six 'best runs' when 
-\textit{k}=2 ended with identical clustering (same mean LLOD score), all other runs 
-differed by at most the allocation of 13 individuals. When \textit{k} was set to three 
-and four all runs provided very similar clustering.
+either an estimate of K$\geq$2 (Run1) or \textit{undecided} as a consequence of no plateau 
+lengths being greater than 6. Interpreting the results from {\sc structure} can either be done 
+with the estimated likelihood of the data from the program itself or by the delta k method 
+suggest by Evanno et al XXXXX. Both methods indicate that the most likely \textit{k} for the dataset is 6.
 
@@
 
-With increased \textit{k} {\sc flock} continues to splits individuals approximately 
-based on geographical boundaries identified by \citet{Garzaetal_norcal} using {\sc 
-structure}.
-
 \begin{figure*}
 \begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-StrFloc.pdf}
+	% NOTE. Pat made Q-val-StrFloc.pdf that has all runs individually, replaced with 
+	% all of them on same axis. It will look fine as a black and white with different 
+	% characters.
+ \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-1plot.pdf}
     \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 

### 2014-05-20 tex/main-body-text.tex

@@
 
-The allocation of indivividuals into K clusters is congruent for each program. For all six groups 
+The allocation of indivividuals into K clusters was congruent for each program. For all six groups 
 of runs from \textit{k}=2-4, {\sc flock} lead to nearly identical clustering of individuals 
 (See supplementary materials). It should be noted
-that of the 50 starting allocations of individuals in the {\sc flock} procedure only the LLOD for the 
- 'best run' is reported. Of these 'best runs' for a \textit{k} of 2 only two  
+that of the 50 starting allocations of individuals in the {\sc flock} procedure likelihoods of each individual in 
+each reference is only reported for the 
+ 'best run'. Of these 'best runs' for a \textit{k} of 2 only two  
 ended with identical clustering (same mean LLOD score), all other runs 
@@
 plateau analysis in {\sc flock} leads to different conclusions among groups of runs, as well as 
-one that differs from both the pure likelihood 
-and delta K approach in {\sc structure}. While no plateaus were observed in any of the six groups 
+one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
 for a \textit{k} $\geq$2, the plateau record varied greatly among each
@@
 either an estimate of K$\geq$2 (Run1) or \textit{undecided} as a consequence of no plateau 
-lengths being greater than 6. Interpreting the results from {\sc structure} can either be done 
-with the estimated likelihood of the data from the program itself or by the delta k method 
-suggest by Evanno et al XXXXX. Both methods indicate that the most likely \textit{k} for the dataset is 6.
+lengths being greater than 6. Interpreting the results from {\sc structure} is often accomplished by 
+looking at the estimated log likelihood of the data given K, the delta k method \citep{Evannoetal2005}, and plots of 
+individual \textit{$q_i$} values. Both the estimated likelihood of the data
+and the plots of individuals \textit{$q_i$} values suggest that 
+the most likely \textit{k} for the dataset is 6.
 
@@
 \section*{Conclusions}
+
+One of the major claimed advantages of {\sc flock} over MCMC methods for solving the 
+K-partition problem is that, '[t]he very short processing times for each run of {\sc flock} allows for
+ comparing partitions from many runs of each \textit{k}'' \citep[][p.~735]{Duc&Tur2012}.
+While the authors observed improved process time in their 
+comparisions we did not observe the same advantage in our analysis. In their comparison, the 
+run parameters for {\sc flock} were 50 runs and 20 iterations for each \textit{k} = 1 to K+4. 
+These process times were compared to the admixture model with correlated allele 
+frequencies in {\sc structure} with ten iterations of a 50,000 sweep burn-in followed 
+by 200,000 sweep sample from the posterior of \textit{k} from 1 to K+3. 
+It seems reasonable, for each \textit{k}, to compare one iteration of {\sc structure} 
+to one iteration of {\sc flock} with 50 different starting allocations that are reallocated 20 times 
+as both evaluate the partition space once. Process speed advantages also may only been observed 
+with low sample sizes. The authors evaluated simulated data sets with sample sizes of N = 60, 120, 
+and 240.
+
 The partition space over which {\sc flock} evaluates how individuals are clustered 
@@
 individuals. It is not surprising that with 2,596 individuals we could have 
-drastically different plateau sequences with just two clusters. It is difficult to 
-access how well the reallocation process moves the cluster of individuals within the 
-partition space due to the fact that the reallocation matrix from only the 'best-run' 
-is reported.
+drastically different plateau sequences with just two clusters and that as \textit{k} increases
+{\sc flock} finds suboptimal allocations.
+
+For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the adhoc rules 
+to estimate K may be overly stringent. While there has been ample 
+discourse concering choosing the optimal K in  {\sc structure} \citep{Evannoetal2005, 
+Pricharedetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
+to be cautious of interpreting K, no such care in interpreting results 
+is suggested in running {\sc flock}. The authors developed estimation rules based on simulated data 
+sets with different migration regimes and rates, number of loci, and K. 
+By relying on identical allocations (mean LLOD values) as a means of supporting 
+a particular \textit{k},  similar but not idential 
+allocations will provide no support to a given \textit{k}. This may not be as big an issue with small datasets,
+however, as sample sizes increase this could have a drastic influence on the estimation of 
+K.

### 2014-05-20 tex/main-body-text.tex
Fixed permissions on grayscale figure, and put it in the document.
Eric edited the intro and Comparison of Methods.

@@
 perceived to be fundamentally novel or new garner more
-attention and accolades---and have a higher chance of publication---
-than methods that are clearly minor elaborations upon existing work. 
+attention and accolades---and have a higher chance of 
+publication---than methods that are clearly minor elaborations upon existing work. 
 Not surprisingly, then, in articles and manuscripts, authors may 
@@
 between their work and existing methods.  Identifying the 
-similarities between new and existing work should help the field 
-progress more quickly by 1) reducing end-user confusion about whether 
+similarities between new and existing work should help our field
+by 1) reducing end-user confusion about whether 
 it is necessary to analyze data with a (sometimes overwhelming) variety 
@@
 intuitive terms with analogies to the coalescence of flocks of birds
-or the accretion of snow on a moving snowball, \etc.
+or the accretion of snow on a moving snowball, \etc
 
@@
 showing it to be a special, restricted case of the simulated annealing
-algorithm for finding the Bayesian maximum {\em a posteriori} (MAP)
+algorithm for finding the Bayesian maximum {\em a posteriori}
 estimate from a marginalized form of the {\sc structure} model with no
-admixture and non-correlated allele frequencies.
+admixture and non-correlated allele frequencies.  For illustration, we subsequently compare
+the results obtained from {\sc flock} and {\sc structure} on a real data
+set with $>$2,500 individuals.
 
@@
 We start with a succinct mathematical description of the {\sc structure}
-algorithm, then describe the {\sc flock} algorithm, and finally explain
+model, then describe the {\sc flock} algorithm, and finally explain
 the close relationship between the two.
@@
 to be drawn independently from the vector of allele frequencies at
-locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where 
+locus $\ell$ in subpopulation $k$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where 
 $A_\ell$ 
 denotes the number of alleles observed in the data set at locus $\ell$.
-Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the 
+Hence, if $Y_{i\ell}$ denotes a vector of length $A_\ell$ whose components denote the 
 number
-of copies of each of the $A_\ell$ alleles at locus $j$ in individual $i$, and 
+of copies of each of the $A_\ell$ alleles at locus $\ell$ in individual $i$, and 
 $Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
@@
 \ell})$,
-which are usually set to a value like $1/A_\ell$.
+which are usually all set to a value such as 1 or $1/A_\ell$.
 
-After initializing the unknown allele frequencies to randomly drawn values,
+After initializing the unknown allele frequencies, $\theta = (\theta_1,\ldots,\theta_K)$, to randomly drawn values,
 $\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
-$Z_i$'s and $\theta = (\theta_1,\ldots,\theta_K)$ using Gibbs sampling.  That is,
+$Z_i$'s and $\theta$ using Gibbs sampling.  That is,
 at iteration $t = 0, 1, 2, \ldots$:
@@
 prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
-\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each 
+\item $\theta^{(t)}$ is updated to $\theta^{(t+1)}$ from its full conditional distribution.  For each 
 cluster
@@
 & & \mathrm{CDM}(\lambda_{k\ell 1} + \#_{(-i)}(k,\ell,1), \ldots,
-\lambda_{k\ell A_\ell} + \#_{(-i)}(k,\ell,A\ell)), \nonumber
+\lambda_{k\ell A_\ell} + \#_{(-i)}(k,\ell,A_\ell)), \nonumber
 \end{eqnarray}
-where $\#_{(-i)}(k,\ell,1)$ is the number of alleles of type $j$ at locus $\ell$
+where $\#_{(-i)}(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$
 found in individuals {\em other than individual $i$} that currently
@@
 Here we translate the recipe given by \citeauthor{Duc&Tur2009} for the 
-{\sc flock} algorithm into a concise specification in terms of the variables
+{\sc flock} algorithm into a specification in terms of the variables
 defined in the previous section.  We provide the description for a case in which
@@
 From that description, it is not clear how \citeauthor{Duc&Tur2009} treat 
-alleles that appear in an
+alleles that appear in the focal
 individual but do not appear within any other individuals within a cluster.
@@
 on maximizing a likelihood that, although it may not be formally identical to
-that in (\ref{eq:cdm}), is not substantively different from it.
+that in (\ref{eq:cdm}), is nearly identical to it.
 
@@
 \begin{equation}
-\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
+\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell}),
 \end{equation}
-\ie, the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$
-(or, as stated above, a very similar function of $k$).
+\ie the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$
+(or, as stated above, a nearly identical function of $k$, depending on how
+the program {\sc flock} treats alleles that are not observed in certain clusters).
 
@@
 \]
-which is the same full conditional distribution as before, raised
+which is the same full conditional distribution as the marginalized {\sc structure}
+model (equation~\ref{eq:fc}), raised
 to the power $\beta$.  In typical applications of simulated annealing,
-$\beta$ is initially set to a low value $<1$ and as the algorithm proceeds
+$\beta$ is set to a starting value less than 1, and, as the algorithm proceeds,
 $\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
@@
 When $\beta$ is small, the process can easily traverse ``valleys''
-in the posterior probability surface and have a better chance of searching
-more of the space for the maximum value.  As $\beta$ gets larger, the process
-tends to move uphill on the probability surface, searching for the
+in the posterior probability surface and has a better chance of searching
+more of the space for the maximum value.  As $\beta$ gets larger, it becomes
+more probable that the process will
+move uphill on the probability surface, tending toward the local
 maximum near where it currently is.
@@
 {\sc flock} seems to find the same solution less consistently than does {\sc structure}.
-
-\subsection*{boneyard}
- On the other hand, in practice, the 
-posterior probability surface may have several modes that are separated by regions
-of low posterior probability, such that the solution
-an algorithm converges to (whether it is {\sc structure}  doing MCMC or {\sc flock} doing 
-``very cold simulated annealing'') depends almost entirely on the initial random
-starting values.
 
@@
 	 % production, it should be changed back to Flock-Fig1.pdf
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.jpg}  
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1-GreyScale.pdf} % Flock-Fig1.jpg}  
     \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 

### 2014-05-20 tex/main-body-text.tex
Edited Fig2, made a greyscale version, and pointed to the GS version in the .tex

--- 2014-05-20 tex/main-body-text.tex
@@
 	% characters.
- \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-1plot.pdf}
+ \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-1plotGreyScale.pdf}
     \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 

### 2014-05-20 tex/main-body-text.tex
Eric edited Comparison of performance section up through the \hrule.

--- 2014-05-20 tex/main-body-text.tex
@@
 \subsection*{Analysis of coastal California steelhead} 
-To evaluate the preformance and computational efficiency of {\sc flock}  we used a published 
-dataset from \citet{Garzaetal_norcal} comprised of 2596 steelhead \textit{Oncorhynchus mykiss}
-genotyped at 15 microsatellite loci.  {\sc flock} was run six times using the default setting for all 
-run parameters \textit{(initial partition = random, number of iterations = 20, number of runs = 50, 
-LLOD threshold = 0)} for values of \textit{k} from 2 to 6. The stopping point of 
-\textit{k}=6 was chosen after four consecutive runs yielded no plateau. The non-admixture, 
-uncorrelated allele frequency model implemented in {\sc structure} was run with six iterations of a 
-50,000 sweep burn-in followed by 150,000 sweep sample from the posterior of \textit{k} from 2 to 6. 
-All runs were performed with an i7 3.40GHz processor and 8GB of RAM. To facilitate the comparison of
- the two programs, the program {\sc clump} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
-while the program Distruct \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+To compare the behavior of {\sc flock} to that of {\sc structure} we used a published 
+dataset of 2596 steelhead \textit{Oncorhynchus mykiss}
+genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal}.  {\sc flock} was run six times using the default setting for all 
+run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
+LLOD threshold = 0) for values of $K$ from 2 to 6. The stopping point of 
+$K=6$ was chosen after four consecutive runs (at $K=6$) % <-- Patrick, is that right. were the non-plateau-ing runs at K=6?
+yielded no plateau (see below). For each value of $K$ from 2 to 6, the no-admixture, 
+uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times using 
+50,000 sweeps of burn-in followed by 150,000 sweeps of sample collection from the posterior. 
+All runs were performed on a (WHAT TYPE OF COMPUTER?) with an i7 3.40GHz processor and 8GB
+of RAM. To facilitate the comparison of results from
+the two programs, we derived $q_i$ values 
+from {\sc flock}'s output by exponentiating the log-likelihood values for cluster membership  % <-- Patrick, please verify.
+and rescaling so they summed to one.  These are then comparable to the $q_i$ values
+returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
+The program {\sc clumpp} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
+while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+
+\citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
+subpopulations.  Their approach involves INSERT SHORT DESCRIPTION, INCLUDING DISCUSSION OF PLATEAUS.
+
+It is now common practice to use the approximate marginal likelihood ($\ln P(D)$) from {\sc structure}
+or a derived quantity such as $\Delta K$ \citep{Evannoetal2005} to try to estimate $K$.
+While we take the stance that estimates of $K$ made with {\em any} unsupervised clustering algorithm
+from data on real (non-idealized) populations should always be interpreted and used
+cautiously, it is nonetheless instructive to compare \citet{Duc&Tur2012}'s method for
+estimating $K$ to results obtained using $\ln P(D)$ and $\Delta K$, and we do so.
 
 \subsection*{Results} 
-Run times were relatively short for both programs. The six groups of runs for {\sc flock} 
-required 1133 minutes to complete. Each individual group of runs 
-averaged 188.8 minutes (1.5 min SD). The six iterations of {\sc structure} completed in 
-486 minutes with each of the six iterations averaging 80.8 minutes (0.8 min SD). Increasing
-the number of putative clusters from \textit{k} of two to six yielded a 2.4 fold increase 
-in processing time for {\sc flock}, whereas a 1.6 fold increase was observed for runs of 
-{\sc structure}.
+Run times were relatively short for both programs. {\sc flock} 
+required 1133 minutes to complete all the runs, averaging 188.8 minutes for 
+each of six groups of runs over $K$ from 2 to 6. {\sc structure} completed all the runs in 
+486 minutes with each of the six groups of runs averaging 80.8 minutes. Increasing
+$K$ from of 2 to 6 increased {\sc flock}'s run time by a factor of 2.4, but {\sc structure}'s
+run time was increased by a factor of only 1.6.  
+\hrule
 
@@
 discourse concering choosing the optimal K in  {\sc structure} \citep{Evannoetal2005, 
-Pricharedetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
+Prichardetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
 to be cautious of interpreting K, no such care in interpreting results 

### 2014-05-26 tex/main-body-text.tex
Filled in results that were unclear/lacking

@@
 LLOD threshold = 0) for values of $K$ from 2 to 6. The stopping point of 
-$K=6$ was chosen after four consecutive runs (at $K=6$) % <-- Patrick, is that right. were the non-plateau-ing runs at K=6?
+$K=6$ was chosen after four consecutive values of $K$
 yielded no plateau (see below). For each value of $K$ from 2 to 6, the no-admixture, 
@@
 50,000 sweeps of burn-in followed by 150,000 sweeps of sample collection from the posterior. 
-All runs were performed on a (WHAT TYPE OF COMPUTER?) with an i7 3.40GHz processor and 8GB
-of RAM. To facilitate the comparison of results from
-the two programs, we derived $q_i$ values 
-from {\sc flock}'s output by exponentiating the log-likelihood values for cluster membership  % <-- Patrick, please verify.
-and rescaling so they summed to one.  These are then comparable to the $q_i$ values
+All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
+of RAM running 64-bit Windows 7 OS. For each run {\sc flock} reports the 
+normalized likelihood values (the exponenetiated log-likelihood value for an individuals 
+membership to each $K$ clusters scaled to sum to one). 
+This value is comparable to the $q_i$ values
 returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
@@
 \citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
-subpopulations.  Their approach involves INSERT SHORT DESCRIPTION, INCLUDING DISCUSSION OF PLATEAUS.
+subpopulations. The user must first specify the number of runs (different initial allocation of individuals 
+to each cluster) as well as the number of iterations individual has the chance to be reallocated 
+to a different cluster. For each run the log-likelihood value of membership to all $K$ clusters is computed
+and the difference between the cluster with the highest likelihood and second highest likelihood is reported (LLOD)
+for each individual. Runs that converge on the same solution are indicated by the same mean LLOD score. and 
+are referred to as a plateau. If for a given $k$ there may be multiple local maxima in addition to 
+the global maximum, depending on the initial allocation of individuals the alorithm will result in finding
+different plateaus. The plateau record is a list of how many runs ended up at these different solutions. 
+\citet{Duc&Tur2012} developed two stopping rules that lead to the estimation of $K$. In the abscence of 
+reasonable values for $k$ \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
+$\geq$6 is reached or four consecutive $k$ values yield no plateaus. If the first stopping condition is satisfied,
+the $K$ for which there is a a single plateau $\geq$6 is the point estimate for $K$. Alternatively, 
+if multiple plateaus are observed with at least one plateau$\geq$6, the largest value of $K$ that satisfies 
+this conditon should be taken as the lower bound estimate of $K$. If no plateaus \geq$6 are observed $K$
+is \textit{undecided}.
 

### 2014-05-27 ForPipeline/Flock2Structure.R
Added skeleton files and some code to make the final formatted files. Sed was pretty fast and required fewer skeleton files

@@ -97,13 +97,6 @@ write(line.temp, file = paste("Flock_PopQ_k=",k,"run",r,".txt", sep = ""))
 }#do for all reps
 }#do for all k clusters
 
-# Let's tidy up the space and put them in a folder 
-# run a system command - Folder = Flock PopQ
-folder<-'Flock\" \"PopQ'
-system(paste0("mkdir ",folder))
-String2pass<-paste('mv ',wd,'/Flock_PopQ_k=*run*.txt ',wd,'/Flock\" \"PopQ',sep="")
-system(String2pass)
-
 #########################################################################################
 # Lets write some code for extracting the individual q values
 # We have already set the number of max clusters and reps above
@@ -182,23 +175,54 @@ write(line.temp2.2, file = paste("Flock_IndQ_k=",k,"run",r,".txt", sep = ""),app
 }#over r
 }#over k
 
-# Let's tidy up the space and put them in a folder 
-# run a system command - Folder = 'Flock IndQ'
-folder<-'Flock\" \"IndQ'
-system(paste0("mkdir ",folder))
-String2pass<-paste('mv ',wd,'/Flock_IndQ_k=*run*.txt ',wd,'/Flock\" \"IndQ',sep="")
-system(String2pass)
 
 #########################################################################################
+#Use sed to paste in the file!
+#sed '/INCLUDE/ r file' <in >out 
+##sed '/INCLUDE/d' file >out 
+#which file to insert
 
-#Now here you have to manually paste in the flock popq and indq values into 
-#The correct location of the Structure output so that Clump and Distruct 
-#can do their thing. 
 
-#If I wanted to write some script one way of doing it would be to split the original 
-#Structure output into 3 sections: Before the popq values, between popq and indq 
-# and after the indq values. I would convert each of those to strings and then 
-# paste all the bits and pieces together. 
+for (k in 2:6){
+for (r in 1:6){
+Flock.pop<- paste("Flock_PopQ_k=",k,"run",r,".txt",sep="") 
[…]

### 2014-05-27 tex/main-body-text.tex
Fixed some $'s so it typeset and changed README to say "use latexmk"
also fixed a pritchardetal

@@
 reasonable values for $k$ \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
-$\geq$6 is reached or four consecutive $k$ values yield no plateaus. If the first stopping condition is satisfied,
-the $K$ for which there is a a single plateau $\geq$6 is the point estimate for $K$. Alternatively, 
-if multiple plateaus are observed with at least one plateau$\geq$6, the largest value of $K$ that satisfies 
-this conditon should be taken as the lower bound estimate of $K$. If no plateaus \geq$6 are observed $K$
+$\geq 6$  is reached or four consecutive $k$ values yield no plateaus. If the first stopping condition is satisfied,
+the $K$ for which there is a a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
+if multiple plateaus are observed with at least one plateau $\geq 6$, the largest value of $K$ that satisfies 
+this conditon should be taken as the lower bound estimate of $K$. If no plateaus $\geq 6$ are observed $K$
 is \textit{undecided}.
@@
 discourse concering choosing the optimal K in  {\sc structure} \citep{Evannoetal2005, 
-Prichardetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
+Pritchardetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
 to be cautious of interpreting K, no such care in interpreting results 

### 2014-05-27 ForPipeline/Flock2Structure.R
minor change to code. Double checked it on another machine and it works.

@@ -51,17 +51,22 @@ write.csv(x=temp,file=file.path(FOR,paste("k",k," Steelhead_Run",r,"_Defaults-po
 ########################################################################################
 # For the next part I am going to need a way of repeating strings with different indexing
 # For this I'll write a function that takes k as an input and outputs a string
-temp.string<-function(k){
-  #need a clever way of repeating stuff inside of a formating loop
-  #this way I don't have to copy and paste for all k
-  string<- vector()
-  for (i in 3:(k+1)){
-    string[i]<-paste('(formatC(Mpopq.str[,',i, '],width=7,flag=','"+"','))',sep="")
-  }
-  string<-string[-(1:2)]
-  string<-paste(string,collapse=",")
-  return(string)
-}
+#temp.string<-function(k){
+#  #need a clever way of repeating stuff inside of a formating loop
+#  #this way I don't have to copy and paste for all k
+#  string<- vector()
+#  for (i in 3:(k+1)){
+#    string[i]<-paste('(formatC(Mpopq.str[,',i, '],width=7,flag=','"+"','))',sep="")
+#  }
+#  string<-string[-(1:2)]
+#  string<-paste(string,collapse=",")
+#  return(string)
+#}
+
+#Eric suggested a vectorized form to replace the loop. 
+temp.string <- function(k) {
+  i<-3:(k+1); paste(paste('(formatC(Mpopq.str[,',i, '],width=7,flag=','"+"','))',sep=""), collapse=",")}
+
 
 # now we are going to read in each of the structure 
 # output files so that we can eventually put our flock 

### 2014-06-23 tex/main-body-text.tex
edited description of flock

@@
 dataset of 2596 steelhead \textit{Oncorhynchus mykiss}
-genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal}.  {\sc flock} was run six times using the default setting for all 
+genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal}.  {\sc flock} 
+was run six times using the default setting for all 
 run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
@@
 subpopulations. The user must first specify the number of runs (different initial allocation of individuals 
-to each cluster) as well as the number of iterations individual has the chance to be reallocated 
-to a different cluster. For each run the log-likelihood value of membership to all $K$ clusters is computed
-and the difference between the cluster with the highest likelihood and second highest likelihood is reported (LLOD)
-for each individual. Runs that converge on the same solution are indicated by the same mean LLOD score. and 
-are referred to as a plateau. If for a given $k$ there may be multiple local maxima in addition to 
-the global maximum, depending on the initial allocation of individuals the alorithm will result in finding
-different plateaus. The plateau record is a list of how many runs ended up at these different solutions. 
+to each cluster) as well as the number of iterations each individual has the opportunity to be reallocated 
+to a different cluster. For each run, the log-likelihood value of membership to all $K$ clusters is computed
+and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
+and second highest likelihood, is reported for each individual. Runs that converge on the same solution 
+are indicated by the same mean LLOD score and 
+are referred to as a plateau. The plateau record is a list of how many runs ended up at these different solutions. 
 \citet{Duc&Tur2012} developed two stopping rules that lead to the estimation of $K$. In the abscence of 
@@
 defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
-second kind. The results of {\sc flock} depend primarily on the initial partition of 
+second kind. The likelihood surface may be characterized by multiple peaks and troughs.
+The peak or plateau that the  {\sc flock} algorithm finds is entirely dependent on the initial allocation of 
 individuals. It is not surprising that with 2,596 individuals we could have 

### 2014-06-23 tex/author-title-etc.tex
minor edits

@@
 % All caps tends to look better
-\newcommand{\myAuthors}{ ERIC C.~ANDERSON$^{*,\dagger,\S}$ and PATRICK BARRY$^\ddag$}
+\newcommand{\myAuthors}{ ERIC C.~ANDERSON$^{*,\dagger,\S}$ and PATRICK D. BARRY$^\ddag$}
 
@@
     Santa Cruz, CA 95060, USA, $^\dagger$Department of Applied Math and Statistics, University of
-California, Santa Cruz, CA, $^\ddag$University of Alaska, Fairbanks, Fisheries Academic Program,
+California, Santa Cruz, CA, $^\ddag$University of Alaska, Fairbanks, School of Fisheries and Ocean Science,
 17101 Point Lena Loop Rd.,
@@
 % upper right header on the odd pages)
-\newcommand{\myRunningAuthor}{Anderson {\em et al.}}
+\newcommand{\myRunningAuthor}{Anderson \& Barry}
 

### 2014-06-23 tex/main-body-text.tex
minor edits

@@
 reasonable values for $k$ \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
-$\geq 6$  is reached or four consecutive $k$ values yield no plateaus. If the first stopping condition is satisfied,
+$\geq 6$  is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
 the $K$ for which there is a a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
 if multiple plateaus are observed with at least one plateau $\geq 6$, the largest value of $K$ that satisfies 
-this conditon should be taken as the lower bound estimate of $K$. If no plateaus $\geq 6$ are observed $K$
+this conditon should be taken as the lower bound estimate of $K$. If no plateaus $\geq 6$ are observed, $K$
 is \textit{undecided}.
@@
 486 minutes with each of the six groups of runs averaging 80.8 minutes. Increasing
-$K$ from of 2 to 6 increased {\sc flock}'s run time by a factor of 2.4, but {\sc structure}'s
-run time was increased by a factor of only 1.6.  
+$K$ from of 2 to 6 increased {\sc flock}'s run time by a factor of 2.4, while the run time 
+for  {\sc structure} increased by a factor of only 1.6.  
 \hrule
@@
 (See supplementary materials). It should be noted
-that of the 50 starting allocations of individuals in the {\sc flock} procedure likelihoods of each individual in 
-each reference is only reported for the 
- 'best run'. Of these 'best runs' for a \textit{k} of 2 only two  
+that of the 50 starting allocations of individuals, the likelihood of membership for every individual 
+to each cluster is only reported for the 'best run'. Of these 'best runs' for a \textit{k} of 2 only two  
 ended with identical clustering (same mean LLOD score), all other runs 
@@
 Despite the similar allocation of individuals into \textit{k} clusters, the 
-plateau analysis in {\sc flock} leads to different conclusions among groups of runs, as well as 
+plateau analysis to estimate $K$ in {\sc flock} leads to different conclusions among groups of runs, as well as 
 one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 

### 2014-06-24 tex/main-body-text.tex
few changes before dinner

@@
 All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
-of RAM running 64-bit Windows 7 OS. For each run {\sc flock} reports the 
+of RAM running 64-bit Windows 7. For each set of runs {\sc flock} reports the 
 normalized likelihood values (the exponenetiated log-likelihood value for an individuals 
-membership to each $K$ clusters scaled to sum to one). 
+membership to each $K$ clusters scaled to sum to one) for the run that belongs to 
+the longest plateau - the 'best run'. 
 This value is comparable to the $q_i$ values
@@
 \citet{Duc&Tur2012} developed two stopping rules that lead to the estimation of $K$. In the abscence of 
-reasonable values for $k$ \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
+reasonable values for $k$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
 $\geq 6$  is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
-the $K$ for which there is a a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
+the $K$ for which there is a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
 if multiple plateaus are observed with at least one plateau $\geq 6$, the largest value of $K$ that satisfies 
@@
 486 minutes with each of the six groups of runs averaging 80.8 minutes. Increasing
-$K$ from of 2 to 6 increased {\sc flock}'s run time by a factor of 2.4, while the run time 
+$K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
 for {\sc structure} increased by a factor of only 1.6.
@@
 
-The allocation of indivividuals into K clusters was congruent for each program. For all six groups 
+The allocation of indivividuals into K clusters was similar for each program. For all six groups 
 of runs from \textit{k}=2-4, {\sc flock} lead to nearly identical clustering of individuals 
-(See supplementary materials). It should be noted
-that of the 50 starting allocations of individuals, the likelihood of membership for every individual 
-to each cluster is only reported for the 'best run'. Of these 'best runs' for a \textit{k} of 2 only two  
+(See supplementary materials) none of which differed substantially from results obtain from 
+{\sc structure} (Fig 1). Of the 'best runs' for the six groups for \textit{k} of 2, two  
 ended with identical clustering (same mean LLOD score), all other runs 
-differed by at most the allocation of 13 individuals. As \textit{k} increases so do the number of solutions
-that {\sc flock} finds (Fig 1). These solutions, however, are suprisingly similar to
+differed by at most the allocation of 13 individuals. As 
+\textit{k} increases so do the number of solutions
+that {\sc flock} finds (Fig 1). For all froups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
+allocations resulted in solutions that differed slightly. These solutions, however, are suprisingly similar to
  the partions identified by {\sc structure}. Both programs idenfity clusters 

### 2014-06-24 tex/main-body-text.tex
few minor edits

@@
 \subsection*{Analysis of coastal California steelhead} 
-To compare the behavior of {\sc flock} to that of {\sc structure} we used a published 
-dataset of 2596 steelhead \textit{Oncorhynchus mykiss}
-genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal}.  {\sc flock} 
+We used a published dataset of 2596 steelhead \textit{Oncorhynchus mykiss}
+genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal} to compare the programs
+{\sc flock} and {\sc structure} .  {\sc flock} 
 was run six times using the default setting for all 
@@
 yielded no plateau (see below). For each value of $K$ from 2 to 6, the no-admixture, 
-uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times using 
-50,000 sweeps of burn-in followed by 150,000 sweeps of sample collection from the posterior. 
+uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times with 
+a 50,000 sweep burn-in followed by 150,000 sweep sample from the posterior. 
 All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
@@
 required 1133 minutes to complete all the runs, averaging 188.8 minutes for 
-each of six groups of runs over $K$ from 2 to 6. {\sc structure} completed all the runs in 
-486 minutes with each of the six groups of runs averaging 80.8 minutes. Increasing
+each of six groups of runs over $K$ from 2 to 6. All runs of {\sc structure} were completed in 
+486 minutes. Each of the six groups of runs averaged 80.8 minutes. Increasing
 $K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
@@
 The allocation of indivividuals into K clusters was similar for each program. For all six groups 
-of runs from \textit{k}=2-4, {\sc flock} lead to nearly identical clustering of individuals 
+of runs from \textit{k} = 2-4, {\sc flock} resulted in nearly identical clustering of individuals 
 (See supplementary materials) none of which differed substantially from results obtain from 
 {\sc structure} (Fig 1). Of the 'best runs' for the six groups for \textit{k} of 2, two  
-ended with identical clustering (same mean LLOD score), all other runs 
+runs yielded identical clustering (same mean LLOD score), all other runs 
 differed by at most the allocation of 13 individuals. As 
 \textit{k} increases so do the number of solutions
-that {\sc flock} finds (Fig 1). For all froups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
-allocations resulted in solutions that differed slightly. These solutions, however, are suprisingly similar to
+that {\sc flock} finds (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
+allocations resulted in solutions that differed. These solutions, however, are suprisingly similar to
  the partions identified by {\sc structure}. Both programs idenfity clusters 

### 2014-06-24 ForPipeline/Flock2Structure.R
added ggplot for fig2 and cleaned up some stuff

@@ -4,11 +4,11 @@
 ###### Note: individuals parts of the code rely on Java and should be run individually 
 ###### If they are all run together you will probably experience a heap error. 
 
-### packages that we will need that probably aren't installed
+### packages that we will need that might not be installed
 install.packages("XLConnect")
 install.packages("stringr")
 
-### 
+### load packages
 require("XLConnect")
 require("stringr")
 
@@ -87,7 +87,7 @@ Flock_popq<-as.matrix(Flock_popq)
 Mpopq.str[,(seq(from=2, to=k+1,by=1))]<- Flock_popq[,c(seq(from=2, to=k+1,by=1))]
 
 # Now I need to write them out with formatting!
-# This will produce some text files will the correct formating to 
+# This will produce some text files will the correct formatting to 
 # paste into the Structure output
 if (k==2){
 line.temp<-paste((formatC(Mpopq.str[,1],width=4,flag="+")),(formatC(Mpopq.str[,2],width=10,flag="+")),(formatC(Mpopq.str[,3],width=7,flag="+")),(formatC(Mpopq.str[,4],width=16,flag="+")),sep="")
@@ -230,6 +230,7 @@ String2pass<-paste('mv ',wd,'/StructOuput_genos_slg_pipe.txt_dat002_k00*_Rep00*.
 system(String2pass)
 
 #########################################################################################
+#########################################################################################
 #For CLUMP and Distruct we will need the names of each of the populations
 #Extract Pop name and number
 Names<-Flock_popq[,1]
@@ -269,7 +270,7 @@ q.floc.string[[r]]<-q.floc.string[[r]][2:(6+k)]
 # make a nice matrix with sample,flock q, structure q values
 q.matrix<-matrix(data=NA, ncol=1+k*2,nrow=2596)
 
-for (n in 1:2596){
+for (n in 1:length(q.str)){
 for (j in 1:k){
   q.matrix[n,j]<-q.str.string[[n]][j+5]
   q.matrix[n,k+j]<-q.floc.string[[n]][j+5]
@@ -301,4 +302,27 @@ points(q.matrix[,k+4],q.matrix[,4],pch=4)
 points(q.matrix[,k+5],q.matrix[,5],pch=5)
 par(xpd=T)
[…]

### 2014-06-24 tex/main-body-text.tex
new fig2 referenced in tex document
more minor edits

@@
 normalized likelihood values (the exponenetiated log-likelihood value for an individuals 
-membership to each $K$ clusters scaled to sum to one) for the run that belongs to 
-the longest plateau - the 'best run'. 
+membership to each $K$ clusters scaled to sum to one) for a run belonging to the longest plateau.
 This value is comparable to the $q_i$ values
@@
 \citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
-subpopulations. The user must first specify the number of runs (different initial allocation of individuals 
+subpopulations. The user must specify both the number of runs (different initial allocation of individuals 
 to each cluster) as well as the number of iterations each individual has the opportunity to be reallocated 
-to a different cluster. For each run, the log-likelihood value of membership to all $K$ clusters is computed
+to a different cluster. For each run, the log-likelihood value for membership to all $K$ clusters is computed
 and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
-and second highest likelihood, is reported for each individual. Runs that converge on the same solution 
+and second highest likelihood, is calculated for each individual. Runs that converge on the same solution 
 are indicated by the same mean LLOD score and 
-are referred to as a plateau. The plateau record is a list of how many runs ended up at these different solutions. 
+are referred to as a plateau. The plateau record is a list of how many runs converged at the same allocation
+of individuals. 
 \citet{Duc&Tur2012} developed two stopping rules that lead to the estimation of $K$. In the abscence of 
@@
 \subsection*{Results} 
-Run times were relatively short for both programs. {\sc flock} 
-required 1133 minutes to complete all the runs, averaging 188.8 minutes for 
-each of six groups of runs over $K$ from 2 to 6. All runs of {\sc structure} were completed in 
-486 minutes. Each of the six groups of runs averaged 80.8 minutes. Increasing
+Run times were relatively short for both programs. All runs of {\sc flock} 
+completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
+All runs of {\sc structure} were completed in 
+486 minutes and averaged 80.8 minutes to complete each of the six groups of runs. Increasing
 $K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
@@
 The allocation of indivividuals into K clusters was similar for each program. For all six groups 
-of runs from \textit{k} = 2-4, {\sc flock} resulted in nearly identical clustering of individuals 
+of runs from \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
 (See supplementary materials) none of which differed substantially from results obtain from 
@@
 for a \textit{k} $\geq$2, the plateau record varied greatly among each
-group for \textit{k}=2 (plateau  sequences- Run1= 3,2,3,7,4; Run2=2,5,2,3,2,3; Run3=2,2,2,2,2,5; 
-Run4=2,3,4,2,4,3; Run5=2,2,5,5,3; and; Run6=2,2,2,2,2,3,3). The plateau analysis 
+group for \textit{k} = 2 (plateau  sequences- Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
+Run4 = 2,3,4,2,4,3; Run5 = 2,2,5,5,3; and; Run6 = 2,2,2,2,2,3,3). The plateau analysis 
 procedure suggested by \citet{Duc&Tur2012} leads to 
-either an estimate of K$\geq$2 (Run1) or \textit{undecided} as a consequence of no plateau 
+either an estimate of K $\geq$ 2 (Run1) or \textit{undecided} as a consequence of no plateau 
 lengths being greater than 6. Interpreting the results from {\sc structure} is often accomplished by 
@@
 \begin{center}
-	% NOTE. Pat made Q-val-StrFloc.pdf that has all runs individually, replaced with 
-	% all of them on same axis. It will look fine as a black and white with different 
-	% characters.
- \includegraphics[width=\textwidth]{images/Figures-Pat/Qval-1plotGreyScale.pdf}
+	% NOTE. Pat replaced old Q-val-StrFloc.pdf with a ggplot version 
+ \includegraphics[width=25em]{images/Figures-Pat/Fig2.pdf}
     \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 

### 2014-06-25 tex/main-body-text.tex
fixed compression issue on Fig2, now it is less ugly

@@
 
-The allocation of indivividuals into K clusters was similar for each program. For all six groups 
-of runs from \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
+We did not observe major differences among the 
+allocation of indivividuals into K clusters between each program. For all six groups 
+of runs for \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
 (See supplementary materials) none of which differed substantially from results obtain from 
@@
 	% NOTE. Pat replaced old Q-val-StrFloc.pdf with a ggplot version 
- \includegraphics[width=25em]{images/Figures-Pat/Fig2.pdf}
+ \includegraphics[width=25em]{images/Figures-Pat/Fig2.eps}
     \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 

### 2014-06-25 ForPipeline/Flock2Structure.R
cleaned up figure folder, made a function for the comparision of runs that outputs a nicer pdf file

@@ -11,6 +11,7 @@ install.packages("stringr")
 ### load packages
 require("XLConnect")
 require("stringr")
+require("ggplot2")
 
 #We are going to work in a folder called 'ForPipeline'
 #So long as the R scrip launches from this point the working directory 
@@ -247,13 +248,18 @@ write(Names,file="Pops.txt")
 
 ##########################################################
 # After running CLUMP and Distruct we can play with the output
-# Lets look at certainty of assignment
-#lets start by pulling two that should look really similar 
-# Read in the clumped files -> clump files should be reorganized.
-k<-5 #which k are we analyzing?
-#which runs do we want to compare
-Str.run<-1
-Flock.run<-5
+# Here is a user defined function to compare the results from different
+#flock and structure runs
+
+Comp.runs<-function(k=k,Str.run=Str.run,Flock.run=Flock.run,greyscale=TRUE,FigOut='Fig2.eps'){
+## Inputs
+#k -> which k are we analyzing?
+#Str.run -> which run from structure do we want to compare
+#Flock.run -> which run from Flock do we want to compare
+#greyscale -> do we want our plot to be greyscale or color?
+#FigOut -> name the Figure... NOTE that the extension determines the type of file we use ggsave
+
+
 # Extract the results from the files in some strings
 q.str<-readLines(con=file.path("intermediate",paste("Output_00",k,".perms_",Str.run,sep="")))
 q.floc<-readLines(con=file.path("intermediate",paste("Output_00",k,".perms_",Flock.run+6,sep="")))
@@ -277,31 +283,6 @@ for (j in 1:k){
 }
 q.matrix[n,2*k+1]<-q.str.string[[n]][1]
 }
-write.csv(q.matrix,"Qval.csv",eol = "\r\n")
-
-par(mfrow=c(1,1),xpd=F)
-plot(q.matrix[,k+1],q.matrix[,1],xlab="Q-val Flock",ylab="Q-val Structure",col="Dark Green")
-abline(0,1)
-points(q.matrix[,k+2],q.matrix[,2],col="Blue")
-points(q.matrix[,k+3],q.matrix[,3],col="Orange")
[…]

### 2014-07-01 tex/abstract.tex
Made a few edits and pushed the hrule a little further down.

@@
 allele frequency priors. The computational routine in {\sc flock} is 
-formally equivalent to searching for the maximum-a-posteriori (MAP) 
+equivalent to searching for the maximum-a-posteriori (MAP) 
 estimate of this restricted {\sc structure} model via a simulated 
 annealing algorithm with an extreme (and likely suboptimal) cooling 
-schedule.  Understanding {\sc flock} as a special case of the model in 
+schedule.  By analysis of a large, real data set we
+confirm that {\sc flock} yields results very similar to the restricted
+{\sc structure} model, albeit with apparently more variability 
+from run to run. Understanding {\sc flock} as a special case of the model in 
 {\sc structure} allows a principled perspective on its expected 

### 2014-07-01 tex/main-body-text.tex
cleaned up figure folder, made a function for the comparision of runs that outputs a nicer pdf file
Made a few edits and pushed the hrule a little further down.

--- 2014-06-24 tex/main-body-text.tex
@@
 intuitive terms with analogies to the coalescence of flocks of birds
-or the accretion of snow on a moving snowball, \etc
+or the accretion of snow on a moving snowball.
 
@@
 number
-of copies of each of the $A_\ell$ alleles at locus $\ell$ in individual $i$, and 
+of copies of each of the $A_\ell$ alleles at locus $\ell$ in a diploid individual $i$, and 
 $Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
@@
 the full conditional distribution of $Z_i$ becomes somewhat more involved,
-but can be computed by the following reasoning. First, the full 
+but can be computed by the following reasoning. First, the 
 conditional distribution of $Y_{i\ell}$, given that individual $i$ is from subpopulation $k$
@@
 integrating out $\theta$.  Thus, we have shown that, doing Gibbs
-sampling for $Z_i$ in  the {\sc structure} in which $\theta$ has been
+sampling for $Z_i$ in  the {\sc structure} model in which $\theta$ has been
 integrated out would involve sampling a new value of $Z_i$ from the 
@@
 \subsection*{Analysis of coastal California steelhead} 
-We used a published dataset of 2596 steelhead \textit{Oncorhynchus mykiss}
+We used a published dataset of 2596 steelhead, \textit{Oncorhynchus mykiss},
 genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal} to compare the programs
-{\sc flock} and {\sc structure} .  {\sc flock} 
+{\sc flock} and {\sc structure}.  {\sc flock} 
 was run six times using the default setting for all 
@@
 uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times with 
-a 50,000 sweep burn-in followed by 150,000 sweep sample from the posterior. 
+a 50,000 sweep burn-in followed by a 150,000 sweep sample from the posterior. 
 All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
-of RAM running 64-bit Windows 7. For each set of runs {\sc flock} reports the 
-normalized likelihood values (the exponenetiated log-likelihood value for an individuals 
-membership to each $K$ clusters scaled to sum to one) for a run belonging to the longest plateau.
+of RAM running 64-bit Windows 7.
+
+For each set of runs, {\sc flock} reports the 
+normalized likelihood values (the exponentiated log-likelihood value for an individual's 
+membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau (see below).
 This value is comparable to the $q_i$ values
@@
 \citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
-subpopulations. The user must specify both the number of runs (different initial allocation of individuals 
-to each cluster) as well as the number of iterations each individual has the opportunity to be reallocated 
-to a different cluster. For each run, the log-likelihood value for membership to all $K$ clusters is computed
+subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
+to each cluster) as well as the number of iterations. (In every iteration each individual has the opportunity to be reallocated 
+to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
 and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
-and second highest likelihood, is calculated for each individual. Runs that converge on the same solution 
-are indicated by the same mean LLOD score and 
-are referred to as a plateau. The plateau record is a list of how many runs converged at the same allocation
+and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
+have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
+{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
+The plateau record is a list of how many runs converged to the same allocation
 of individuals. 
-\citet{Duc&Tur2012} developed two stopping rules that lead to the estimation of $K$. In the abscence of 
-reasonable values for $k$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
+\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
+strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
 $\geq 6$  is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
@@
 cautiously, it is nonetheless instructive to compare \citet{Duc&Tur2012}'s method for
-estimating $K$ to results obtained using $\ln P(D)$ and $\Delta K$, and we do so.
+estimating $K$ to results obtained using $\ln P(D)$ and $\Delta K$, and we do so, below.
 
@@
 for {\sc structure} increased by a factor of only 1.6.
+
+We did not observe major differences among the 
+allocation of individuals into K clusters between each program. All six groups 
+of {\sc flock} runs for $K = 2$ to $K=4$ yielded nearly identical clustering of individuals 
+(See supplementary materials) none of which differed substantially from results obtained with 
+{\sc structure} (Fig.~\ref{fig:qplots}). Of the 'best runs' for the six groups for $K=2$, two  
+runs yielded identical clustering (same mean LLOD score), all other runs 
+differed by at most the allocation of 13 individuals. As 
+$K$ increases so does the number of solutions
+that {\sc flock} finds (Fig.~\ref{fig:qplots}). For all groups of runs for $K\geq 2$, no plateaus were observed - all initial 
+allocations resulted in solutions that differed. These solutions, however, are surprisingly similar to
+the cluster membership posterior means identified by {\sc structure}. Both programs identify clusters 
+which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal}.
+
+Despite the similar allocation of individuals into $K$ clusters, the 
+plateau analysis to estimate $K$ in {\sc flock} leads to different conclusions among groups of runs, as well as 
+one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
+for a $K\geq 2$, the plateau record varied greatly among each
+group for $K = 2$ (plateau sequences: Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
+Run4 = 2,3,4,2,4,3; Run5 = 2,2,5,5,3; and; Run6 = 2,2,2,2,2,3,3). The plateau analysis 
+procedure suggested by \citet{Duc&Tur2012} leads to 
+either an estimate of $K\geq 2$ (Run1) or {\em undecided} as a consequence of no plateau 
+lengths being greater than 6. 
 \hrule
 
-The allocation of indivividuals into K clusters was similar for each program. For all six groups 
-of runs from \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
-(See supplementary materials) none of which differed substantially from results obtain from 
-{\sc structure} (Fig 1). Of the 'best runs' for the six groups for \textit{k} of 2, two  
-runs yielded identical clustering (same mean LLOD score), all other runs 
-differed by at most the allocation of 13 individuals. As 
-\textit{k} increases so do the number of solutions
-that {\sc flock} finds (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
-allocations resulted in solutions that differed. These solutions, however, are suprisingly similar to
- the partions identified by {\sc structure}. Both programs idenfity clusters 
-which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal}.
-
-Despite the similar allocation of individuals into \textit{k} clusters, the 
-plateau analysis to estimate $K$ in {\sc flock} leads to different conclusions among groups of runs, as well as 
-one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
-for a \textit{k} $\geq$2, the plateau record varied greatly among each
-group for \textit{k} = 2 (plateau  sequences- Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
-Run4 = 2,3,4,2,4,3; Run5 = 2,2,5,5,3; and; Run6 = 2,2,2,2,2,3,3). The plateau analysis 
-procedure suggested by \citet{Duc&Tur2012} leads to 
-either an estimate of K $\geq$ 2 (Run1) or \textit{undecided} as a consequence of no plateau 
-lengths being greater than 6. Interpreting the results from {\sc structure} is often accomplished by 
+Interpreting the results from {\sc structure} is often accomplished by 
 looking at the estimated log likelihood of the data given K, the delta k method \citep{Evannoetal2005}, and plots of 
@@
 is the lower bound estimate.}
-    \label{Fig.1}
+    \label{fig:qplots}
 \end{center}
@@
  attributed to them.}
-    \label{Fig.2}
+    \label{fig:qscatter}
 \end{center}

### 2014-07-01 tex/main-body-text.tex
editing the resutls section

--- 2014-06-25 tex/main-body-text.tex
@@
 genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal} to compare the programs
-{\sc flock} and {\sc structure} .  {\sc flock} 
+{\sc flock} and {\sc structure}.  {\sc flock} 
 was run six times using the default setting for all 
@@
 normalized likelihood values (the exponenetiated log-likelihood value for an individuals 
-membership to each $K$ clusters scaled to sum to one) for a run belonging to the longest plateau.
+membership to each $K$ clusters scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012}.
 This value is comparable to the $q_i$ values
@@
 \subsection*{Results} 
-Run times were relatively short for both programs. All runs of {\sc flock} 
+Run times for both programs were relatively short, albeit non-trivial. All runs of {\sc flock} 
 completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
 All runs of {\sc structure} were completed in 
-486 minutes and averaged 80.8 minutes to complete each of the six groups of runs. Increasing
+486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
+The processing time of each algorithm increased with as the number or 
+putative clusters increases. Increasing
 $K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
@@
 
-We did not observe major differences among the 
-allocation of indivividuals into K clusters between each program. For all six groups 
+{\sc flock} and {\sc structure} provide similar results in 
+terms of how individuals are allocated among runs at low values of $K$, 
+while, at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, multiple other solutions. For all six groups 
 of runs for \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
@@
 that {\sc flock} finds (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
-allocations resulted in solutions that differed. These solutions, however, are suprisingly similar to
+allocations resulted in solutions that differed. 
+It is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
+equal to one. Many of these solutions are, however, suprisingly similar to
  the partions identified by {\sc structure}. Both programs idenfity clusters 
-which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal}.
+which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
+although less reliably so in the case of {\sc flock}.
 
@@
 either an estimate of K $\geq$ 2 (Run1) or \textit{undecided} as a consequence of no plateau 
-lengths being greater than 6. Interpreting the results from {\sc structure} is often accomplished by 
-looking at the estimated log likelihood of the data given K, the delta k method \citep{Evannoetal2005}, and plots of 
-individual \textit{$q_i$} values. Both the estimated likelihood of the data
-and the plots of individuals \textit{$q_i$} values suggest that 
-the most likely \textit{k} for the dataset is 6.
+lengths being greater than 6. The ($\ln P(D)$) from {\sc structure}
+was largest at $K$ = 6 which was supported by the $q_i$ plots. The $\Delta K$ method, however, 
+supports a $K$ = 2.
 
@@
     \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1-GreyScale.pdf} % Flock-Fig1.jpg}  
-    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 
-in the program {\sc flock}. Each horizontal plot represents the 'best run' from a 
-series where individuals are reallocated into \textit{k} clusters. Within each plot 
-the normalized individual likelihood for each reference population is represented by a 
-colored vertical bar. Major geographic boundaries that were attributed to genetic 
-structure in \citet{Garzaetal_norcal} are indicated with black vertical lines. Results 
-from the 'best runs' showed similar geographic differentiation to 
-\citet{Garzaetal_norcal} analysis. The plateau analysis developed by 
-\citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 
-is the lower bound estimate.}
+    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
+from {\sc flock} made in the program Distruct.
+Each horizontal plot represents an individual run of the program {\sc structure}, or the 'best run' from a series of runs in {\sc flock}.
+Within each plot
+$q_i$ values from {\sc structure} and the normalized likelihood values 
+from {\sc flock}  are represented by a 
+colored vertical bar.  Results from 
+from the 'best runs' of {\sc flock} showed similar geographic differentiation to 
+\citet{Garzaetal_norcal}.}
     \label{Fig.1}
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}
 	% NOTE. Pat replaced old Q-val-StrFloc.pdf with a ggplot version 
- \includegraphics[width=25em]{images/Figures-Pat/Fig2.eps}
+ \includegraphics[width=25em]{images/Figures-Pat/Fig2.pdf}
     \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 
@@
 \end{center}
-\end{figure*}
+\end{figure}
 

### 2014-07-01 tex/main-body-text.tex
clean up merge conflicts

--- 2014-07-01 tex/main-body-text.tex
@@
 normalized likelihood values (the exponentiated log-likelihood value for an individual's 
-membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau (see below).
+membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
 This value is comparable to the $q_i$ values
@@
 
-We did not observe major differences among the 
-allocation of individuals into K clusters between each program. All six groups 
-of {\sc flock} runs for $K = 2$ to $K=4$ yielded nearly identical clustering of individuals 
-(See supplementary materials) none of which differed substantially from results obtained with 
-{\sc structure} (Fig.~\ref{fig:qplots}). Of the 'best runs' for the six groups for $K=2$, two  
+{\sc flock} and {\sc structure} provide similar results in 
+terms of how individuals are allocated among runs at low values of $K$, 
+while, at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
+multiple other solutions. For all six groups 
+of runs for \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
+(See supplementary materials) none of which differed substantially from results obtain from 
+{\sc structure} (Fig 1). Of the 'best runs' for the six groups for \textit{k} of 2, two  
 runs yielded identical clustering (same mean LLOD score), all other runs 
 differed by at most the allocation of 13 individuals. As 
-$K$ increases so does the number of solutions
-that {\sc flock} finds (Fig.~\ref{fig:qplots}). For all groups of runs for $K\geq 2$, no plateaus were observed - all initial 
-allocations resulted in solutions that differed. These solutions, however, are surprisingly similar to
-the cluster membership posterior means identified by {\sc structure}. Both programs identify clusters 
-which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal}.
+\textit{k} increases so do the number of solutions
+that {\sc flock} finds (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
+allocations resulted in solutions that differed. 
+It is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
+equal to one. Many of these solutions are, however, suprisingly similar to
+ the partions identified by {\sc structure}. Both programs idenfity clusters 
+which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
+although less reliably so in the case of {\sc flock}.
 
@@
 procedure suggested by \citet{Duc&Tur2012} leads to
-either an estimate of $K\geq 2$ (Run1) or {\em undecided} as a consequence of no plateau 
-lengths being greater than 6. 
+either an estimate of K $\geq$ 2 (Run1) or {\em undecided} as a consequence of no plateau 
+lengths being greater than 6. The ($\ln P(D)$) from {\sc structure}
+was largest at $K$ = 6 which was supported by the $q_i$ plots. The $\Delta K$ method, however, 
+supports a $K$ = 2.
+
 \hrule
-
-Interpreting the results from {\sc structure} is often accomplished by 
-looking at the estimated log likelihood of the data given K, the delta k method \citep{Evannoetal2005}, and plots of 
-individual \textit{$q_i$} values. Both the estimated likelihood of the data
-and the plots of individuals \textit{$q_i$} values suggest that 
-the most likely \textit{k} for the dataset is 6.
 
@@
     \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1-GreyScale.pdf} % Flock-Fig1.jpg}  
-    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 
-in the program {\sc flock}. Each horizontal plot represents the 'best run' from a 
-series where individuals are reallocated into \textit{k} clusters. Within each plot 
-the normalized individual likelihood for each reference population is represented by a 
-colored vertical bar. Major geographic boundaries that were attributed to genetic 
-structure in \citet{Garzaetal_norcal} are indicated with black vertical lines. Results 
-from the 'best runs' showed similar geographic differentiation to 
-\citet{Garzaetal_norcal} analysis. The plateau analysis developed by 
-\citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 
-is the lower bound estimate.}
-    \label{fig:qplots}
+    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
+from {\sc flock} made in the program Distruct.
+Each horizontal plot represents an individual run of the program {\sc structure}, or the 'best run' from a series of runs in {\sc flock}.
+Within each plot
+$q_i$ values from {\sc structure} and the normalized likelihood values 
+from {\sc flock}  are represented by a 
+colored vertical bar.  Results from 
+from the 'best runs' of {\sc flock} showed similar geographic differentiation to 
+\citet{Garzaetal_norcal}.}
+    \label{Fig.1}
 \end{center}
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}

### 2014-07-01 tex/main-body-text.tex
minor edits to correct some merge edits that I reverted

@@
 486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
-The processing time of each algorithm increased with as the number or 
+The processing time of each algorithm increased as the number or 
 putative clusters increases. Increasing
 $K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
-for {\sc structure} increased by a factor of only 1.6.
+for {\sc structure} increased by a factor of 1.6.
 
@@
 terms of how individuals are allocated among runs at low values of $K$, 
-while, at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
+while at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
 multiple other solutions. For all six groups 
@@
 (See supplementary materials) none of which differed substantially from results obtain from 
-{\sc structure} (Fig 1). Of the 'best runs' for the six groups for \textit{k} of 2, two  
-runs yielded identical clustering (same mean LLOD score), all other runs 
+{\sc structure} (Fig 1). Of the 'best runs' for the six groups for $K$ = 2, two  
+runs yielded identical clustering (same mean LLOD score) while all other runs 
 differed by at most the allocation of 13 individuals. As 
 \textit{k} increases so do the number of solutions
-that {\sc flock} finds (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
+that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
 allocations resulted in solutions that differed. 
@@
 \citet{Garzaetal_norcal}.}
-    \label{Fig.1}
+    \label{fig:qplots}
 \end{center}
@@
 One of the major claimed advantages of {\sc flock} over MCMC methods for solving the 
-K-partition problem is that, '[t]he very short processing times for each run of {\sc flock} allows for
+K-partition problem is that, 'The very short processing times for each run of {\sc flock} allows for
  comparing partitions from many runs of each \textit{k}'' \citep[][p.~735]{Duc&Tur2012}.

### 2014-07-01 tex/main-body-text.tex
edits to conclusion section

@@
 allocations resulted in solutions that differed. 
-It is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
+{\sc flock} still produces estimates of the LLOD by individual for the 
+'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
 equal to one. Many of these solutions are, however, suprisingly similar to
@@
 lengths being greater than 6. The ($\ln P(D)$) from {\sc structure}
-was largest at $K$ = 6 which was supported by the $q_i$ plots. The $\Delta K$ method, however, 
-supports a $K$ = 2.
+was largest at $K$ = 6 which was further supported by the $q_i$ plots, however,
+the $\Delta K$ method supported a $K$ of 2.
 
@@
 only the reallocation matrix for the 'best run' is given as output. Increasing the 
-number of iterations did not affect the overall inference of \textit{k}. It was 
+number of iterations did not affect the overall inference of \textit{k}. We 
 observed that often during the last iterations individuals would alternate between 
-reference groups during subsequent iterations. Increasing the log likelihood 
-difference threshold for reallocation to 0.18 (there needs to be at least a 1.5 times 
+reference groups. Increasing the log likelihood 
+difference threshold for reallocation to 0.18 (requiring at least a 1.5 times 
 higher likelihood to reallocate the individual) minimized this, but did not affect the 
@@
 frequencies in {\sc structure} with ten iterations of a 50,000 sweep burn-in followed 
-by 200,000 sweep sample from the posterior of \textit{k} from 1 to K+3. 
-It seems reasonable, for each \textit{k}, to compare one iteration of {\sc structure} 
-to one iteration of {\sc flock} with 50 different starting allocations that are reallocated 20 times 
-as both evaluate the partition space once. Process speed advantages also may only been observed 
-with low sample sizes. The authors evaluated simulated data sets with sample sizes of N = 60, 120, 
-and 240.
+by a 200,000 sweep sample from the posterior of \textit{k} from 1 to K+3. 
+It seemed reasonable to compare, for each \textit{k}, one iteration of {\sc structure} 
+to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
+as both evaluate the partition space once. The dataset used in this comparision was an order of magnatude
+larger than the largest used by \citep{Duc&Tur2012} (N = 60, 120, and 240). Any process speed advantages 
+may only been observed with small sample sizes.
 
@@
 greatly increases with both the number of individuals and the number of clusters into 
-which they can be partitioned. The total partition space for any given k can be 
+which they can be partitioned. The total partition space can be 
 defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
-second kind. The likelihood surface may be characterized by multiple peaks and troughs.
+second kind, $k$ is the number of clusters, and $n$ is the number of individuals.
+The likelihood surface spanning all possible partitions
+may be characterized by multiple peaks and troughs.
 The peak or plateau that the  {\sc flock} algorithm finds is entirely dependent on the initial allocation of 
-individuals. It is not surprising that with 2,596 individuals we could have 
+individuals as it has no way to visit areas of the likelihood surface with smaller values.
+ It is not surprising that with 2,596 individuals we could have 
 drastically different plateau sequences with just two clusters and that as \textit{k} increases
-{\sc flock} finds suboptimal allocations.
+{\sc flock} finds suboptimal allocations. If the number of initial allocations does not increase
+as $K$ is increased, the starting allocations become diffusely scattered in the space of all
+possible partitions making it unlikely that they will converge.
 
@@
 Pritchardetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
-to be cautious of interpreting K, no such care in interpreting results 
-is suggested in running {\sc flock}. The authors developed estimation rules based on simulated data 
-sets with different migration regimes and rates, number of loci, and K. 
+to be cautious of interpreting K, much less effort has been put forth examining how the number of individuals
+and thenumber of starting allocations can affect not only the convergence of different runs, but also the 
+inference of $K$. The authors developed estimation rules based on simulated data 
+sets with a modest number of individuals and different migration regimes and rates, number of loci, and K. 
 By relying on identical allocations (mean LLOD values) as a means of supporting 
 a particular \textit{k},  similar but not idential 
-allocations will provide no support to a given \textit{k}. This may not be as big an issue with small datasets,
-however, as sample sizes increase this could have a drastic influence on the estimation of 
-K.
+allocations will provide no support to a given \textit{k}.

### 2014-07-01 tex/main-body-text.tex
Fit Fig 2 into a single column.
Note that I pdfcropped Fig2.pdf to squeak a few more mm out of it.
Changed k to K

@@
 multiple other solutions. For all six groups 
-of runs for \textit{k} = 2 - 4, {\sc flock} resulted in nearly identical clustering of individuals 
+of runs for $K = 2$ to 4, {\sc flock} resulted in nearly identical clustering of individuals 
 (See supplementary materials) none of which differed substantially from results obtain from 
@@
 differed by at most the allocation of 13 individuals. As 
-\textit{k} increases so do the number of solutions
+$K$ increases so do the number of solutions
 that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
@@
 	% NOTE. Pat replaced old Q-val-StrFloc.pdf with a ggplot version 
- \includegraphics[width=25em]{images/Figures-Pat/Fig2.pdf}
-    \caption{Individual \textit{$q_i$} values plotted for each cluster for \textit{k}=5. 
+ \includegraphics[width=\columnwidth]{images/Figures-Pat/Fig2.pdf}
+    \caption{Individual \textit{$q_i$} values plotted for each cluster for $K=5$. 
 {\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
@@
 only the reallocation matrix for the 'best run' is given as output. Increasing the 
-number of iterations did not affect the overall inference of \textit{k}. We 
+number of iterations did not affect the overall inference of $K$. We 
 observed that often during the last iterations individuals would alternate between 
@@
 higher likelihood to reallocate the individual) minimized this, but did not affect the 
-inference of \textit{k} (lower bound estimate of \textit{k} = 2).
+inference of $K$ (lower bound estimate of $K = 2$).
 
@@
 One of the major claimed advantages of {\sc flock} over MCMC methods for solving the 
-K-partition problem is that, 'The very short processing times for each run of {\sc flock} allows for
- comparing partitions from many runs of each \textit{k}'' \citep[][p.~735]{Duc&Tur2012}.
+K-partition problem is that, ``The very short processing times for each run of {\sc flock} allows for
+ comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
 While the authors observed improved process time in their 
 comparisions we did not observe the same advantage in our analysis. In their comparison, the 
-run parameters for {\sc flock} were 50 runs and 20 iterations for each \textit{k} = 1 to K+4. 
+run parameters for {\sc flock} were 50 runs and 20 iterations for each $K$ from 1 to four more than 
+the true value of $K$. 
 These process times were compared to the admixture model with correlated allele 
 frequencies in {\sc structure} with ten iterations of a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior of \textit{k} from 1 to K+3. 
-It seemed reasonable to compare, for each \textit{k}, one iteration of {\sc structure} 
+by a 200,000 sweep sample from the posterior of $K$ from 1 to K+3. 
+It seemed reasonable to compare, for each $K$, one iteration of {\sc structure} 
 to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
@@
  It is not surprising that with 2,596 individuals we could have 
-drastically different plateau sequences with just two clusters and that as \textit{k} increases
+drastically different plateau sequences with just two clusters and that as $K$ increases
 {\sc flock} finds suboptimal allocations. If the number of initial allocations does not increase
@@
 By relying on identical allocations (mean LLOD values) as a means of supporting 
-a particular \textit{k},  similar but not idential 
-allocations will provide no support to a given \textit{k}.
+a particular $K$,  similar but not idential 
+allocations will provide no support to a given $K$.

### 2014-07-03 tex/main-body-text.tex
edited conclusion

@@
 was largest at $K$ = 6 which was further supported by the $q_i$ plots, however,
-the $\Delta K$ method supported a $K$ of 2.
+the $\Delta K$ method supported a $K$ of 3.
 
@@
 \section*{Conclusions}
+The use of genetic markers to identify population structure, specifically
+the number of genetic clusters in a collection $K$ has drastically exanded
+in the past decade to aid in the fields of evolution, epidemiology, and conservation
+genetics. By virtute of this exanding desire to identify $K$ there has been a push to 
+develop methods that can accurately infer $K$ while at the same time be easy to run
+both in terms of proccess time and user interface. Naturally, a program claiming
+``more accurate allocations and more reliable estimates of $K$ [while running]
+much faster than {\sc structure} \citep[][p.~734]{Duc&Tur2012}'' is particularly 
+appealing. Despite these two claims, we have shown that the algorithm implemented 
+in {\sc flock} is equivalent to 
+a restrictured case of the simulated annealing algorthim to find the Bayesian 
+MAP estimate from the marginalized form of the no admixture and non-correlated
+allele frequency model in {\sc structure} with no processing speed advantage for real datasets.
 
-One of the major claimed advantages of {\sc flock} over MCMC methods for solving the 
+It is not surprising then, that because of the similaries in the algortims implemented that 
+we obsered  
+congruent results between the 'best run' from {\sc flock} and {\sc structure} using 
+a real data set from \textit{O. mykiss}. Though we observed very similar results
+between the two programs at low values of $K$, {\sc flock} had a tendency to be inconsistent
+among its 'best runs' at higher values - while still finding a similar solution to {\sc structure}.
+This is most likely a result of the rapidly exanding partition space that must be evaluated as $K$ increases. 
+The total partition space is 
+defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
+second kind, $k$ is the number of clusters, and $n$ is the number of individuals.
+The likelihood surface spanning all possible partitions could
+be characterized by multiple peaks and troughs.
+The peak or plateau that the  {\sc flock} algorithm finds is entirely dependent on the initial allocation of 
+individuals as it has no way to visit areas of the likelihood surface with smaller values.
+If the number of initial allocations does not increase
+as $K$ is increased, the starting allocations become diffusely scattered in the space of all
+possible partitions making it unlikely that they will converge. This coupled with the fact that there is 
+ambiguity in what defines a 'best run' when all plateaus are lenght zero obfuscates any interpretation
+of results from {\sc flock} if no non-zero plateaus are encountered.
+
+One of the major claimed advantages of {\sc flock} over methods for solving the 
 K-partition problem is that, ``The very short processing times for each run of {\sc flock} allows for
@@
 While the authors observed improved process time in their 
-comparisions we did not observe the same advantage in our analysis. In their comparison, the 
+comparisions we did not observe the same advantage in our analysis. This appears to result from two issues:
+(1) the number of individuals in each collection and (2) the lenght of runs for each program in each comparision. 
+In their comparison, the 
 run parameters for {\sc flock} were 50 runs and 20 iterations for each $K$ from 1 to four more than 
@@
 frequencies in {\sc structure} with ten iterations of a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior of $K$ from 1 to K+3. 
+by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the true 
+value of $K$. 
 It seemed reasonable to compare, for each $K$, one iteration of {\sc structure} 
 to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
-as both evaluate the partition space once. The dataset used in this comparision was an order of magnatude
-larger than the largest used by \citep{Duc&Tur2012} (N = 60, 120, and 240). Any process speed advantages 
-may only been observed with small sample sizes.
-
-The partition space over which {\sc flock} evaluates how individuals are clustered 
-greatly increases with both the number of individuals and the number of clusters into 
-which they can be partitioned. The total partition space can be 
-defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
-second kind, $k$ is the number of clusters, and $n$ is the number of individuals.
-The likelihood surface spanning all possible partitions
-may be characterized by multiple peaks and troughs.
-The peak or plateau that the  {\sc flock} algorithm finds is entirely dependent on the initial allocation of 
-individuals as it has no way to visit areas of the likelihood surface with smaller values.
- It is not surprising that with 2,596 individuals we could have 
-drastically different plateau sequences with just two clusters and that as $K$ increases
-{\sc flock} finds suboptimal allocations. If the number of initial allocations does not increase
-as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it unlikely that they will converge.
+as both evaluate the partition space once. We chose to run six iterations for each program
+in order to compare the consistency of results between programs and among runs.
+In addition to changing the number of groups of runs of {\sc flock} to compare,
+the dataset used was an order of magnatude
+larger than the largest used by \citep{Duc&Tur2012} (N = 60, 120, and 240). It may hold that {\sc flock}
+is able to process small datasets faster than {\sc structure}, but with the rapid increase in our ability to economically
+genotype a large number of individuals for a great many number of loci, this advantage may not be relevant to many 
+study designs.
 
@@
 to estimate K may be overly stringent. While there has been ample 
-discourse concering choosing the optimal K in {\sc structure} \citep{Evannoetal2005, 
-Pritchardetal2000, Wap&Gag2006} and \citet{Pritchardetal2000} urges users 
-to be cautious of interpreting K, much less effort has been put forth examining how the number of individuals
-and thenumber of starting allocations can affect not only the convergence of different runs, but also the 
+discourse concering the choice of the optimal K in {\sc structure} \citep{Evannoetal2005, 
+Pritchardetal2000, Wap&Gag2006} much less effort has been put forth examining how the number of individuals
+and the number of starting allocations can affect not only the convergence of different runs, but also the 
 inference of $K$. The authors developed estimation rules based on simulated data 
@@
 a particular $K$,  similar but not idential 
-allocations will provide no support to a given $K$.
+allocations will provide no support to a given $K$. We highly advise, if using {\sc flock}, to
+run multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
+{\sc district} \citep{Rosenberg2004} to visually inspect the 'best run' results.
+
+The development of software to analyze genetic data is ever increasing. In the face of a growing
+number of programs, end users must wade through a seemingly countless number of programs each with
+their own assumptions, strengths, and weaknesses. Lacking a recognition of these properties
+can result in the ineffectual use of both time and resources.
+By providing a different prespective on the algorithm
+implemented in {\sc flock} we hope to increase understanding of its applicability to a different datasets
+while decreasing end user confusion and misuse.

### 2014-07-03 ForPipeline/Flock2Structure.R

@@ -52,18 +52,6 @@ write.csv(x=temp,file=file.path(FOR,paste("k",k," Steelhead_Run",r,"_Defaults-po
 ########################################################################################
 # For the next part I am going to need a way of repeating strings with different indexing
 # For this I'll write a function that takes k as an input and outputs a string
-#temp.string<-function(k){
-#  #need a clever way of repeating stuff inside of a formating loop
-#  #this way I don't have to copy and paste for all k
-#  string<- vector()
-#  for (i in 3:(k+1)){
-#    string[i]<-paste('(formatC(Mpopq.str[,',i, '],width=7,flag=','"+"','))',sep="")
-#  }
-#  string<-string[-(1:2)]
-#  string<-paste(string,collapse=",")
-#  return(string)
-#}
-
 #Eric suggested a vectorized form to replace the loop. 
 temp.string <- function(k) {
   i<-3:(k+1); paste(paste('(formatC(Mpopq.str[,',i, '],width=7,flag=','"+"','))',sep=""), collapse=",")}
@@ -308,7 +296,7 @@ p1 <- ggplot(data=df, aes(x=Flock_q, y=Struct_q, shape=factor(k))) +
   #theme(legend.background = element_rect(fill="gray95", size=.5, linetype="dotted")) +
   labs(y=expression(paste("Structure ", italic(q[i])," values",sep=""), "values",sep=" "),x=expression(paste("Flock ", italic(q[i])," values",sep=""))) +
   geom_abline(intercept=0, slope=1, linetype=2)
-
+p1
 } else {
   p1 <- ggplot(data=df, aes(x=Flock_q, y=Struct_q, shape=factor(k),color=factor(k))) +
     geom_point(alpha=0.5) +
@@ -321,10 +309,97 @@ p1 <- ggplot(data=df, aes(x=Flock_q, y=Struct_q, shape=factor(k))) +
     #theme(legend.background = element_rect(fill="gray95", size=.5, linetype="dotted")) +
     labs(y=expression(paste("Structure ", italic(q[i])," values",sep=""), "values",sep=" "),x=expression(paste("Flock ", italic(q[i])," values",sep=""))) +
[…]

### 2014-07-07 tex/main-body-text.tex
couple quick edits while gel is running

@@
 The use of genetic markers to identify population structure, specifically
-the number of genetic clusters in a collection $K$ has drastically exanded
+the number of genetic clusters in a collection ($K$) has drastically expanded
 in the past decade to aid in the fields of evolution, epidemiology, and conservation
-genetics. By virtute of this exanding desire to identify $K$ there has been a push to 
-develop methods that can accurately infer $K$ while at the same time be easy to run
-both in terms of proccess time and user interface. Naturally, a program claiming
+genetics. By virtute of this expanding desire to identify $K$ there has been a push to 
+develop methods that can accurately infer $K$ while at the same time remain acessable
+to a wide range of end users and run quickly. Naturally, a program claiming
 ``more accurate allocations and more reliable estimates of $K$ [while running]
@@
 in {\sc flock} is equivalent to 
-a restrictured case of the simulated annealing algorthim to find the Bayesian 
+a restricted case of the simulated annealing algorthim to find the Bayesian 
 MAP estimate from the marginalized form of the no admixture and non-correlated
@@
 comparisions we did not observe the same advantage in our analysis. This appears to result from two issues:
-(1) the number of individuals in each collection and (2) the lenght of runs for each program in each comparision. 
-In their comparison, the 
+(1) the number of individuals in each collection and (2) the length of runs for each program in each comparision. 
+In the comparison by \citep{Duc&Tur2012}, the 
 run parameters for {\sc flock} were 50 runs and 20 iterations for each $K$ from 1 to four more than 

### 2014-07-08 tex/main-body-text.tex
Edits to conclusions.

--- 2014-07-01 tex/main-body-text.tex
@@
 was largest at $K$ = 6 which was further supported by the $q_i$ plots, however,
-the $\Delta K$ method supported a $K$ of 2.
-
-\hrule
+the $\Delta K$ method supported a $K$ of 3.
 
@@
 \section*{Conclusions}
+In the past decade, the use of genetic markers to identify population structure and to attempt inference of
+the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, epidemiology, and conservation
+genetics. By virtute of this expanding desire to identify $K$ there has been a push to 
+develop fast and easy-to-use methods that can accurately infer $K$. We show that {\sc flock}, a program billed as providing
+``more accurate allocations and more reliable estimates of $K$ [while running]
+much faster than {\sc structure} \citep[][p.~734]{Duc&Tur2012},'' may not, in fact, represent such 
+a radical departure from {\sc structure} itself.  In particular, we have shown that the algorithm implemented 
+in {\sc flock} is a simulated annealing algorithm for the Bayesian 
+MAP estimate from a marginalized and restricted form of the {\sc structure} model.  
+Given such similarities between the two methods it is not surprising that
+we observed relatively
+congruent results between the ``best run'' from {\sc flock} and {\sc structure} using 
+a real data set from \textit{O. mykiss}.
 
-One of the major claimed advantages of {\sc flock} over MCMC methods for solving the 
-K-partition problem is that, ``The very short processing times for each run of {\sc flock} allows for
+Though we observed very similar results
+between the two programs at low values of $K$, {\sc flock} had a tendency to provide inconsistent solutions
+among its ``best runs'' at higher values, while still finding solutions that were similar to {\sc structure}'s.
+This is likely a consequence of the rapidly increasing size of the search space over possible 
+partitions as $K$ and $n$, the number of individuals, increase. 
+The likelihood surface defined on partitions of this space is likely to be
+be characterized by multiple peaks and troughs.
+The peak or plateau that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
+individuals, as it has no way to visit areas of the surface with smaller likelihood values.
+If the number of initial allocations does not increase
+as $K$ is increased, the starting allocations become diffusely scattered in the space of all
+possible partitions making it unlikely that they will converge. This coupled with the fact that there is 
+ambiguity in what defines a "best run" when all plateaus are of length zero complicates the interpretation
+of results from {\sc flock} if no non-zero plateaus are encountered.
+
+One of the major claimed advantages of {\sc flock} over other methods for genetic clustering
+is that, ``The very short processing times for each run of {\sc flock} allows for
 comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
-While the authors observed improved process time in their 
-comparisions we did not observe the same advantage in our analysis. In their comparison, the 
-run parameters for {\sc flock} were 50 runs and 20 iterations for each $K$ from 1 to four more than 
-the true value of $K$. 
-These process times were compared to the admixture model with correlated allele 
-frequencies in {\sc structure} with ten iterations of a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior of $K$ from 1 to K+3. 
-It seemed reasonable to compare, for each $K$, one iteration of {\sc structure} 
+While the authors observed faster processing times (relative to other methods) in their 
+comparisons, we did not observe the same advantage in our analysis. This appears to result from two issues. First,
+our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  Second, 
+in the comparison by \citet{Duc&Tur2012}, 
+{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ from 1 to four more than 
+the true value of $K$, and computing times for that analysis were compared to times for ten iterations of {\sc structure} using 
+a 50,000 sweep burn-in followed 
+by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the true 
+value of $K$. 
+To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 200,000 sample collection sweeps)
 to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
-as both evaluate the partition space once. The dataset used in this comparision was an order of magnatude
-larger than the largest used by \citep{Duc&Tur2012} (N = 60, 120, and 240). Any process speed advantages 
-may only been observed with small sample sizes.
-
-The partition space over which {\sc flock} evaluates how individuals are clustered 
-greatly increases with both the number of individuals and the number of clusters into 
-which they can be partitioned. The total partition space can be 
-defined by $\sum\limits_{i=1}^k S(n,k)$ where S(n,k) is a Stirling number of the 
-second kind, $k$ is the number of clusters, and $n$ is the number of individuals.
-The likelihood surface spanning all possible partitions
-may be characterized by multiple peaks and troughs.
-The peak or plateau that the  {\sc flock} algorithm finds is entirely dependent on the initial allocation of 
-individuals as it has no way to visit areas of the likelihood surface with smaller values.
- It is not surprising that with 2,596 individuals we could have 
-drastically different plateau sequences with just two clusters and that as $K$ increases
-{\sc flock} finds suboptimal allocations. If the number of initial allocations does not increase
-as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it unlikely that they will converge.
+as both are used to provide a single estimate of the partition of individuals into clusters. One could
+argue that it would be more appropriate to compare the running times of the programs according to the number
+of opportunities that each individual's cluster membership has for random initialization and reallocation.  Under 
+such a perspective, {\sc flock}'s 50 different starting allocations and 20 reallocations should be compared to
+50 separate runs of structure, each of length 20 sweeps, which would surely give {\sc structure}
+an apparent advantage in speed.  Regardless, while it may hold that {\sc flock}
+is able to process small datasets quickly enough to compete with {\sc structure},
+the rapid increase in our ability to economically
+genotype a large number of individuals for a great many number of loci, may render this advantage less relevant to many 
+study designs in the future.
 

### 2014-07-09 tex/main-body-text.tex
Comment out Figure 2.
We hadn't actually referenced it in the main text and I feel
like the paper might be long enough already.  We have it there
in case a referee wants us to put it back in.
Partial edits to conclusion.

@@
 
-\begin{figure}
-\begin{center}
-	% NOTE. Pat replaced old Q-val-StrFloc.pdf with a ggplot version 
- \includegraphics[width=\columnwidth]{images/Figures-Pat/Fig2.pdf}
-    \caption{Individual \textit{$q_i$} values plotted for each cluster for $K=5$. 
-{\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
- attributed to them.}
-    \label{fig:qscatter}
-\end{center}
-\end{figure}
+%\begin{figure}
+%\begin{center}
+% \includegraphics[width=\columnwidth]{images/Figures-Pat/Fig2.pdf}
+%    \caption{Individual \textit{$q_i$} values plotted for each cluster for $K=5$. 
+%{\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
+% attributed to them.}
+%    \label{fig:qscatter}
+%\end{center}
+%\end{figure}
 
@@
 
-One of the major claimed advantages of {\sc flock} over other methods for genetic clustering
+One of the major claimed advantages of {\sc flock} over other methods for genetic clustering (like {\sc structure})
 is that, ``The very short processing times for each run of {\sc flock} allows for
@@
 the rapid increase in our ability to economically
-genotype a large number of individuals for a great many number of loci, may render this advantage less relevant to many 
+genotype a large number of individuals for a great many loci, may render this advantage less relevant to many 
 study designs in the future.
@@
 For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the adhoc rules 
-to estimate K may be overly stringent. While there has been ample 
-discourse concering the choice of the optimal K in {\sc structure} \citep{Evannoetal2005, 
-Pritchardetal2000, Wap&Gag2006} much less effort has been put forth examining how the number of individuals
-and the number of starting allocations can affect not only the convergence of different runs, but also the 
+to estimate $K$ may be overly stringent. While there has been lengthy 
+discourse over the choice of estimators for $K$ in {\sc structure} \citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011} much less effort has been put forth examining how the number of individuals
+and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
 inference of $K$. The authors developed estimation rules based on simulated data 
-sets with a modest number of individuals and different migration regimes and rates, number of loci, and K. 
+sets with a modest number of individuals and different migration regimes and rates, number of loci, and $K$. 
 By relying on identical allocations (mean LLOD values) as a means of supporting 
-a particular $K$,  similar but not idential 
-allocations will provide no support to a given $K$. We highly advise, if using {\sc flock}, to
-run multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
-{\sc district} \citep{Rosenberg2004} to visually inspect the 'best run' results.
+a particular $K$,  similar---but not identical---allocations
+will provide no support to a given $K$. If using {\sc flock}, we advise running
+multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
+{\sc distruct} \citep{Rosenberg2004} to visually inspect the ``best run'' results.
 

### 2014-07-09 tex/main-body-text.tex
Final additions to conclusion and last paragraph.

@@
 
-The development of software to analyze genetic data is ever increasing. In the face of a growing
-number of programs, end users must wade through a seemingly countless number of programs each with
-their own assumptions, strengths, and weaknesses. Lacking a recognition of these properties
-can result in the ineffectual use of both time and resources.
-By providing a different prespective on the algorithm
-implemented in {\sc flock} we hope to increase understanding of its applicability to a different datasets
-while decreasing end user confusion and misuse.
+For many types of genetic analyses, molecular ecologists can choose from 
+amongst a wide range of published methods.  Choosing one of these methods to use can
+be difficult, especially when the methods have been touted as very different.
+We have known colleagues whose solution to this problem is simply to run their data
+through every program available.  This seems an ineffective use of time and resources 
+that could be counteracted by a more transparent accounting of the similarities between
+methods. As one small step toward that,   
+we have provided a different prespective on the algorithm
+implemented in {\sc flock}, showing it to be very similar to {\sc structure}
+using a particular set of options.  We hope this might increase understanding of {\sc flock}'s applicability to different datasets while decreasing end user confusion and misuse.

### 2014-07-09 tex/main-body-text.tex
some final edits to conclusion

@@
 MAP estimate from a marginalized and restricted form of the {\sc structure} model.  
-Given such similarities between the two methods it is not surprising that
+Given such similarities between the two methods, it is not surprising that
 we observed relatively
@@
 in the comparison by \citet{Duc&Tur2012}, 
-{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ from 1 to four more than 
-the true value of $K$, and computing times for that analysis were compared to times for ten iterations of {\sc structure} using 
+{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
+one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of {\sc structure} using 
 a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the true 
+by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the most likely 
 value of $K$.
-To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 200,000 sample collection sweeps)
+To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 150,000 sample collection sweeps)
 to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 

### 2014-07-09 tex/main-body-text.tex
Edits during final read through.

--- 2014-07-09 tex/main-body-text.tex
@@
 was a point estimate of the $Z_i$'s that maximized the posterior probability,
-then one standard approach to finding that maximimum {\em a posteriori}, or MAP, 
+then one standard approach to finding that maximum {\em a posteriori}, or MAP, 
 estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
@@
 membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
-This value is comparable to the $q_i$ values
+These values can be compared to the $q_i$ values
 returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
@@
 subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
-to each cluster) as well as the number of iterations. (In every iteration each individual has the opportunity to be reallocated 
+to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
 to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
@@
 \citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
-strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
-$\geq 6$  is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
-the $K$ for which there is a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
-if multiple plateaus are observed with at least one plateau $\geq 6$, the largest value of $K$ that satisfies 
-this conditon should be taken as the lower bound estimate of $K$. If no plateaus $\geq 6$ are observed, $K$
-is \textit{undecided}.
+strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
+$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
+the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
+if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
+this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
+is declared {\em undecided}.
 
@@
 \subsection*{Results} 
-Run times for both programs were relatively short, albeit non-trivial. All runs of {\sc flock} 
+Run times for both programs were relatively short, albeit not trivial. All runs of {\sc flock} 
 completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
@@
 multiple other solutions. For all six groups 
-of runs for $K = 2$ to 4, {\sc flock} resulted in nearly identical clustering of individuals 
-(See supplementary materials) none of which differed substantially from results obtain from 
-{\sc structure} (Fig 1). Of the 'best runs' for the six groups for $K$ = 2, two  
+of runs with $K = 2$ to 4, {\sc flock} obtained a nearly identical clustering of individuals 
+(See supplementary materials) none of which differed substantially from {\sc structure}'s results
+(Fig.~\ref{fig:qplots}). Of the ``best runs'' for the six groups for $K$ = 2, two  
 runs yielded identical clustering (same mean LLOD score) while all other runs 
 differed by at most the allocation of 13 individuals. As 
-$K$ increases so do the number of solutions
-that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
+$K$ increased so did the number of solutions
+that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed---all intial 
 allocations resulted in solutions that differed. 
-{\sc flock} still produces estimates of the LLOD by individual for the 
+{\sc flock} still produced estimates of the LLOD by individual for the 
 'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
-equal to one. Many of these solutions are, however, suprisingly similar to
- the partions identified by {\sc structure}. Both programs idenfity clusters 
+equal to one. Many of these solutions are, however, surprisingly similar to
+the partitions identified by {\sc structure}. Both programs idenfity clusters 
 which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
@@
 populations to move around the partition space we varied the number of iterations that 
-each run goes through to find the 'optimal' allocation of individuals. Unfortuantely 
-only the reallocation matrix for the 'best run' is given as output. Increasing the 
+each run goes through to find the ``optimal'' allocation of individuals. Unfortuantely 
+only the reallocation matrix for the ``best run'' is given as output. Increasing the 
 number of iterations did not affect the overall inference of $K$. We 
@@
 the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, epidemiology, and conservation
-genetics. By virtute of this expanding desire to identify $K$ there has been a push to 
+genetics. By virtue of this expanding desire to identify $K$ there has been a push to 
 develop fast and easy-to-use methods that can accurately infer $K$. We show that {\sc flock}, a program billed as providing
@@
 between the two programs at low values of $K$, {\sc flock} had a tendency to provide inconsistent solutions
-among its ``best runs'' at higher values, while still finding solutions that were similar to {\sc structure}'s.
+among its ``best runs'' at higher values of $K$, while still finding solutions that were similar to {\sc structure}'s.
 This is likely a consequence of the rapidly increasing size of the search space over possible 
@@
 be characterized by multiple peaks and troughs.
-The peak or plateau that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
+The peak that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
 individuals, as it has no way to visit areas of the surface with smaller likelihood values.
@@
 as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it unlikely that they will converge. This coupled with the fact that there is 
-ambiguity in what defines a "best run" when all plateaus are of length zero complicates the interpretation
+possible partitions making it unlikely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
+ambiguity in what defines a ``best run'' when all plateaus are of length zero complicates the interpretation
 of results from {\sc flock} if no non-zero plateaus are encountered.
@@
 comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
-While the authors observed faster processing times (relative to other methods) in their 
+While \citeauthor{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
 comparisons, we did not observe the same advantage in our analysis. This appears to result from two issues. First,
@@
 a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the true 
+by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
 value of $K$. 
@@
 50 separate runs of structure, each of length 20 sweeps, which would surely give {\sc structure}
-an apparent advantage in speed.  Regardless, while it may hold that {\sc flock}
-is able to process small datasets quickly enough to compete with {\sc structure},
-the rapid increase in our ability to economically
-genotype a large number of individuals for a great many loci, may render this advantage less relevant to many 
-study designs in the future.
+an apparent advantage in speed.  Regardless, we did not find a speed advantage from using {\sc flock}.
+If anything, {\sc structure} was somewhat faster.
 
-For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the adhoc rules 
+For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the {\em ad hoc} rules 
 to estimate $K$ may be overly stringent. While there has been lengthy 
@@
 and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
-inference of $K$. The authors developed estimation rules based on simulated data 
+inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules based on simulated data 
 sets with a modest number of individuals and different migration regimes and rates, number of loci, and $K$. 
@@
 methods. As one small step toward that,   
-we have provided a different prespective on the algorithm
+we have provided a different perspective on the algorithm
 implemented in {\sc flock}, showing it to be very similar to {\sc structure}

### 2014-07-09 tex/main-body-text.tex
Edit figure caption slightly.

@@
     \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
-from {\sc flock} made in the program Distruct.
-Each horizontal plot represents an individual run of the program {\sc structure}, or the 'best run' from a series of runs in {\sc flock}.
+from {\sc flock} made with the program {\sc distruct}.
+Each horizontal plot represents an individual run of the program {\sc structure}, or the ``best run'' from a series of runs in {\sc flock}.
 Within each plot
@@
 from {\sc flock}  are represented by a 
-colored vertical bar.  Results from 
-from the 'best runs' of {\sc flock} showed similar geographic differentiation to 
-\citet{Garzaetal_norcal}.}
+gray-scale vertical bar.
+The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same solution, or how many groups of {\sc flock} runs (out of 6) had the same ``best run.''}
     \label{fig:qplots}

### 2014-07-09 tex/main-body-text.tex

@@
 in the comparison by \citet{Duc&Tur2012}, 
-{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ from 1 to four more than 
-the true value of $K$, and computing times for that analysis were compared to times for ten iterations of {\sc structure} using 
+{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
+one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of {\sc structure} using 
 a 50,000 sweep burn-in followed 

### 2014-08-08 tex/abstract.tex
Substantial edits to shorten introduction, etc.

--- 2014-03-28 tex/abstract.tex
@@
 
-      We show that the program {\sc flock} \citep{Duc&Tur2009} employs 
-a model that is identical to the {\sc structure} 
+      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
+interpreted as an estimation procedure using  
+a model that is essentially identical to the {\sc structure} 
 \citep{Pritchardetal2000} model with no admixture and non-correlated 
 allele frequency priors. The computational routine in {\sc flock} is 
-formally equivalent to searching for the maximum-a-posteriori (MAP) 
+equivalent to searching for the maximum-{\em a-posteriori} (MAP) 
 estimate of this restricted {\sc structure} model via a simulated 
-annealing algorithm with an extreme (and likely suboptimal) cooling 
-schedule.  Understanding {\sc flock} as a special case of the model in 
-{\sc structure} allows a principled perspective on its expected 
-strengths and weaknesses.
+annealing algorithm with an extreme cooling 
+schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
+confirm that {\sc flock} yields results similar to this restricted
+{\sc structure} model, albeit with apparently more variability 
+from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
+{\sc structure} should aid in the understanding of the program's output and behavior.

### 2014-08-08 tex/main-body-text.tex
Substantial edits to shorten introduction, etc.

@@
 \section*{Introduction}
-In most fields, proposed methods that are 
-perceived to be fundamentally novel or new garner more
-attention and accolades---and have a higher chance of 
-publication---than methods that are clearly minor elaborations upon existing work. 
-Not surprisingly, then, in articles and manuscripts, authors may 
-emphasize the differences and downplay the similarities between their 
-work and the published literature.   In molecular ecology, this 
-tendency amongst the creators of statistical methodology can make it 
-difficult for end-users to understand the relationship between 
-different methods.
-
-Of course, new methods almost always build upon pre-existing ones, and 
-much is to be gained by understanding the close relationship between 
-different statistical methods in use in molecular ecology today.  
-Indeed, some authors downplay the similarities between their method and 
-existing ones not because they wish to make their method seem more 
-novel, but rather because they are genuinely unaware of the close ties 
-between their work and existing methods.  Identifying the 
-similarities between new and existing work should help our field
-by 1) reducing end-user confusion about whether 
-it is necessary to analyze data with a (sometimes overwhelming) variety 
-of computer programs; 2) establishing a common language upon which to 
-compare different methods; and 3) providing a principled perspective 
-from which to argue about the expected strengths and weaknesses of any 
-method.
-
 Methods for the unsupervised clustering of genotypes have received 
@@
 The best known example of this class of methods is {\sc structure} 
-\citep{Pritchardetal2000}.  {\sc structure} itself can be described 
-as an elaboration of earlier models used to identify the spawning 
-stock of salmon \citep{Smouseetal1990}, and a variety of other 
+\citep{Pritchardetal2000}, with over 12,000 citations identified on 
+Google Scholar.  A variety of other 
 clustering methods have been developed that are all closely related to 
@@
 BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
-\citep{Coranderetal2004}. An overview of these similarities can be 
+\citep{Coranderetal2004}. An overview of the similarities between these methods can be 
 found in \citet{Anderson2009PGAC}.
 
-Recently, a description of the software {\sc flock} was published, 
-billing the method as a ``non-Bayesian method [that] 
+Recently, several colleagues suggested we try using the software {\sc flock} \citep{Duc&Tur2009},
+for analyzing a number of the data sets from our laboratories.
+{\sc flock} is billed as a ``non-Bayesian method [that] 
 differs substantially from previous 
-clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. A following
-paper asserts that
-\begin{quote}
-``{\sc flock} is very different from  other clustering 
-programs. It does not sample the space of partitions through small 
+clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. As described
+in a later paper, {\sc flock} ``does not sample the space of partitions through small 
 random step walks as in MCMC, and it does not try to optimize some 
@@
 entirely deterministic'' \citep[][p.~736]{Duc&Tur2012}.
-\end{quote}
-These papers describe the rationale behind the {\sc flock} algorithm in
-intuitive terms with analogies to the coalescence of flocks of birds
-or the accretion of snow on a moving snowball.
 
-Here, we provide a different perspective on the {\sc flock} algorithm,
-showing it to be a special, restricted case of the simulated annealing
-algorithm for finding the Bayesian maximum {\em a posteriori}
+In trying to understand the behavior of {\sc flock} we identified 
+a close relationship between {\sc flock}'s algorithm and the program
+{\sc structure}.  In other words, although the algorithm in {\sc flock}
+is, as the authors point out, entirely deterministic, it can be interpreted as a 
+limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
+models.  More specifically, the {\sc flock} 
+algorithm is a special case of the simulated annealing
+algorithm for finding the Bayesian maximum-{\em a-posteriori}
 estimate from a marginalized form of the {\sc structure} model with no
-admixture and non-correlated allele frequencies.  For illustration, we subsequently compare
-the results obtained from {\sc flock} and {\sc structure} on a real data
-set with $>$2,500 individuals.
+admixture and non-correlated allele frequencies.
+
+Being able to interpret {\sc flock} in this fashion should help users
+interpret output from {\sc flock} and predict when it might and might not
+give substantially different results than {\sc structure}. Below we describe
+the correspondence between the programs in more detail, and then, for illustration, compare
+the results obtained from {\sc flock} and {\sc structure} on a large, real data
+set.
 
@@
 \subsection*{{\sc structure}}
-In the {\sc structure} model with no admixture, the unknown ``subpopulation'' that the 
+In the {\sc structure} model with no admixture (which, it should be pointed out, 
+is not the default model in {\sc structure}), the unknown ``subpopulation'' that the 
 $i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, 
@@
 update 
-$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} 
+$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \citep{Hue&And2007} 
 when not
@@
 
-\subsection*{{\sc flock} as a special case of {\sc structure}}
+\subsection*{The {\sc flock} algorithm and {\sc structure}}
 The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
@@
 was a point estimate of the $Z_i$'s that maximized the posterior probability,
-then one standard approach to finding that maximum {\em a posteriori}, or MAP, 
+then one standard approach to finding that maximum-{\em a-posteriori}, or MAP, 
 estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
@@
 ``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
-the algorithm.''  As such, we can expect that it may not provide the best 
-mechanism for searching for the MAP estimate and it may be more susceptible 
+the algorithm.''  As such, we can expect that it may be more susceptible 
 to becoming trapped in local modes of the posterior probability. 
@@
 
-\section*{Performance Comparison with {\sc structure}}
+\section*{Comparison of {\sc structure} and {\sc flock} results}
 
-\subsection*{Analysis of coastal California steelhead} 
 We used a published dataset of 2596 steelhead, \textit{Oncorhynchus mykiss},
@@
 
-\citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
+\citet{Duc&Tur2012} describe a method for using {\sc flock} to estimate, $K$, the number of
 subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
@@
 (See supplementary materials) none of which differed substantially from {\sc structure}'s results
-(Fig.~\ref{fig:qplots}). Of the ``best runs'' for the six groups for $K$ = 2, two  
+(Fig.~\ref{fig:qplots}). Of the `best runs' for the six groups for $K$ = 2, two  
 runs yielded identical clustering (same mean LLOD score) while all other runs 
@@
 from {\sc flock} made with the program {\sc distruct}.
-Each horizontal plot represents an individual run of the program {\sc structure}, or the ``best run'' from a series of runs in {\sc flock}.
+Each horizontal plot represents an individual run of the program {\sc structure}, or the `best run' from a series of runs in {\sc flock}.
 Within each plot
@@
 gray-scale vertical bar.
-The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same solution, or how many groups of {\sc flock} runs (out of 6) had the same ``best run.''}
+The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same `best run.'}
     \label{fig:qplots}
@@
 each run goes through to find the ``optimal'' allocation of individuals. Unfortuantely 
-only the reallocation matrix for the ``best run'' is given as output. Increasing the 
+only the reallocation matrix for the `best run' is given as output. Increasing the 
 number of iterations did not affect the overall inference of $K$. We 
@@
 In the past decade, the use of genetic markers to identify population structure and to attempt inference of
-the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, epidemiology, and conservation
-genetics. By virtue of this expanding desire to identify $K$ there has been a push to 
-develop fast and easy-to-use methods that can accurately infer $K$. We show that {\sc flock}, a program billed as providing
-``more accurate allocations and more reliable estimates of $K$ [while running]
-much faster than {\sc structure} \citep[][p.~734]{Duc&Tur2012},'' may not, in fact, represent such 
-a radical departure from {\sc structure} itself.  In particular, we have shown that the algorithm implemented 
-in {\sc flock} is a simulated annealing algorithm for the Bayesian 
-MAP estimate from a marginalized and restricted form of the {\sc structure} model.  
-Given such similarities between the two methods, it is not surprising that
-we observed relatively
-congruent results between the ``best run'' from {\sc flock} and {\sc structure} using 
-a real data set from \textit{O. mykiss}.
+the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
+epidemiology, and conservation
+genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
+a full-featured program with many options that is widely-used, well-tested, and reasonably well-understood. 
+We've shown here that {\sc flock}, a program that uses a deterministic algorithm, can be regarded
+as providing approximate, maximum-{\em a-posteriori} inference from one of the models implemented in 
+{\sc structure}. Given the similarities between the two methods, we expect that {\sc flock}
+and {\sc structure} (using the no-admixture model with non-correlated allele frequencies) will deliver
+similar results.  We observed such similarity in comparing results from {\sc flock} and {\sc structure} 
+on a real data set from \textit{O. mykiss}
 
-Though we observed very similar results
-between the two programs at low values of $K$, {\sc flock} had a tendency to provide inconsistent solutions
-among its ``best runs'' at higher values of $K$, while still finding solutions that were similar to {\sc structure}'s.
+Though we observed similar results
+between the two programs at low values of $K$, {\sc flock}'s ``best-runs'' solutions became somewhat inconsistent at higher values of $K$, though most solutions were still similar to {\sc structure}'s.
 This is likely a consequence of the rapidly increasing size of the search space over possible 
 partitions as $K$ and $n$, the number of individuals, increase. 
-The likelihood surface defined on partitions of this space is likely to be
-be characterized by multiple peaks and troughs.
-The peak that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
+The likelihood surface defined on partitions of this space will often
+be multimodal---possessing numerous local peaks and troughs.
+The peak that  {\sc flock}'s deterministic algorithm finds is entirely determined by the initial allocation of 
 individuals, as it has no way to visit areas of the surface with smaller likelihood values.
@@
 as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it unlikely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
-ambiguity in what defines a ``best run'' when all plateaus are of length zero complicates the interpretation
+possible partitions making it less likely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
+ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
 of results from {\sc flock} if no non-zero plateaus are encountered.
+Of course, the problem of getting caught in local modes is not unique to {\sc flock}---{\sc structure} itself clearly can wind up sampling predominantly from a fairly isolated part of the space, as evidenced by the different results obtained from different runs of {\sc structure}.
 
-One of the major claimed advantages of {\sc flock} over other methods for genetic clustering (like {\sc structure})
-is that, ``The very short processing times for each run of {\sc flock} allows for
-comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
-While \citeauthor{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
-comparisons, we did not observe the same advantage in our analysis. This appears to result from two issues. First,
-our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  Second, 
-in the comparison by \citet{Duc&Tur2012}, 
+One of the advantages of {\sc flock} over other methods for genetic clustering
+is reported to be the short processing time for each run of {\sc flock}.
+While \citet{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
+comparisons, we did not observe this advantage in our analysis. This appears to result from two issues. First,
+our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  
+Second, in the comparison by \citet{Duc&Tur2012}, 
 {\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
@@
 value of $K$. 
-To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 200,000 sample collection sweeps)
+To us it seemed reasonable to compare, for each $K$, the running times of one iteration of {\sc structure} (50,000 
+burn in and 200,000 sample collection sweeps)
 to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
-as both are used to provide a single estimate of the partition of individuals into clusters. One could
-argue that it would be more appropriate to compare the running times of the programs according to the number
-of opportunities that each individual's cluster membership has for random initialization and reallocation.  Under 
-such a perspective, {\sc flock}'s 50 different starting allocations and 20 reallocations should be compared to
-50 separate runs of structure, each of length 20 sweeps, which would surely give {\sc structure}
-an apparent advantage in speed.  Regardless, we did not find a speed advantage from using {\sc flock}.
-If anything, {\sc structure} was somewhat faster.
+as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
+clusters. In short,
+it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
+constitutes a reasonable comparison with respect to run length and number of runs.
 
-For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the {\em ad hoc} rules 
-to estimate $K$ may be overly stringent. While there has been lengthy 
-discourse over the choice of estimators for $K$ in {\sc structure} \citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011} much less effort has been put forth examining how the number of individuals
+While there has been lengthy 
+discourse over the choice of estimators for $K$ in {\sc structure} 
+\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
+individuals
 and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
-inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules based on simulated data 
-sets with a modest number of individuals and different migration regimes and rates, number of loci, and $K$. 
-By relying on identical allocations (mean LLOD values) as a means of supporting 
-a particular $K$,  similar---but not identical---allocations
+inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
+sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
+By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
+runs as a means of supporting 
+a particular $K$,  similar---but not identical---partitions
 will provide no support to a given $K$. If using {\sc flock}, we advise running
 multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
-{\sc distruct} \citep{Rosenberg2004} to visually inspect the ``best run'' results.
+{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, to determine
+if there are many similar-but-not-identical partitions found amongst {\sc flock}'s replicate 
+solutions.
 
-For many types of genetic analyses, molecular ecologists can choose from 
-amongst a wide range of published methods.  Choosing one of these methods to use can
-be difficult, especially when the methods have been touted as very different.
-We have known colleagues whose solution to this problem is simply to run their data
-through every program available.  This seems an ineffective use of time and resources 
-that could be counteracted by a more transparent accounting of the similarities between
-methods. As one small step toward that,   
-we have provided a different perspective on the algorithm
-implemented in {\sc flock}, showing it to be very similar to {\sc structure}
-using a particular set of options.  We hope this might increase understanding of {\sc flock}'s applicability to different datasets while decreasing end user confusion and misuse.
+We should point out that our comparison was between {\sc flock} and the {\em no-admixture} model in 
+{\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 
+its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
+of individuals in the sample and thus may be a more appropriate model than the no-admixture model
+(or {\sc flock}) when the sample contains individuals that are admixed between populations.  A further advantage
+of the with-admixture model is that it allows inferences for the population of origin of individual alleles at 
+different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 
+corresponds to {\sc structure}'s model with uncorrelated allele frequency priors. \citet{Falushetal2003} found that
+using their ``$F$-model,'' which assumes that allele frequencies between populations are correlated {\em a priori}, 
+allowed {\sc structure} to dissect more subtle population structure than was possible using the prior with
+uncorrelated allele frequencies.  Modifying {\sc flock} so as to reap similar benefits from using a prior
+with correlated allele frequencies seems like it would be an interesting challenge, though difficult
+because {\sc flock}'s formulation effectively integrates the allele frequencies out of the model.
+
+Often, newly-developed algorithms come to be understood within more general contexts and 
+frameworks, and this, in turn, can help guide improvement of those algorithms.  For example, in population
+genetics, the ``Markov recursion'' algorithm of \citet{Gri&Tav1994-AI} was identified by \citet{Felsensteinetal1999} 
+as a special case of importance sampling \citep{Ham&Han1964}, a perspective which allowed 
+\citet{Ste&Don2000} to improve upon the algorithm dramatically.  Here we have identified the {\sc flock}
+algorithm as a limiting case of simulating annealing upon {\sc structure}'s no-admixture model. We hope this will
+help users to understand and interpret {\sc flock}'s behavior and will provide insights that may be useful 
+to {\sc flock}'s authors if they further develop their algorithm.

### 2014-08-08 tex/abstract.tex
Minor edits

@@
 schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
-confirm that {\sc flock} yields results similar to this restricted
-{\sc structure} model, albeit with apparently more variability 
+confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
+model and uncorrelated allele frequency prior, albeit with apparently more variability 
 from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
-{\sc structure} should aid in the understanding of the program's output and behavior.
+{\sc structure} should aid in  understanding the program's output and behavior.

### 2014-08-08 tex/author-title-etc.tex
Substantial edits to shorten introduction, etc.
Minor edits
Wrangled everything to get a single-column submission version, too.

@@
 %%%%% INFORMATION AND  A FEW OTHER THINGS
-\newcommand{\myTitle}{A statistically-grounded interpretation of {\sc flock}}
+\newcommand{\myTitle}{Interpreting the {\sc flock} algorithm from a statistical \\ perspective}
+\title{\myTitle}
 
@@
 % All caps tends to look better
-\newcommand{\myAuthors}{ ERIC C.~ANDERSON$^{*,\dagger,\S}$ and PATRICK D. BARRY$^\ddag$}
+\newcommand{\myAuthors}{ ERIC C.~ANDERSON$^{*,\S}$ and PATRICK D. BARRY$^\dagger$}
+\author{ERIC C.~ANDERSON$^{*}$ and PATRICK D. BARRY$^\dagger$}
 
@@
     110 Shaffer Road,
-    Santa Cruz, CA 95060, USA, $^\dagger$Department of Applied Math and Statistics, University of
-California, Santa Cruz, CA, $^\ddag$University of Alaska, Fairbanks, School of Fisheries and Ocean Science,
+    Santa Cruz, CA 95060, USA, $^\dagger$University of Alaska, Fairbanks, School of Fisheries and Ocean Science,
 17101 Point Lena Loop Rd.,
 Juneau, AK 99801, USA }
+\renewcommand{\AuthorAddresses}{\myAffiliations}
+
+\renewcommand{\KeyWords}{Unsupervised clustering, simulated annealing, software}
+
+\renewcommand{\CorrespondingAuthor}{Eric C. Anderson, Fisheries Ecology Division, Southwest Fisheries Science
+Center, 110 Shaffer Road, Santa Cruz, CA 95060. eric.anderson@noaa.gov}
 
@@
 % even pages)
-\newcommand{\myRunningTitle}{{\sc flock} as a special case of {\sc structure}}
+\newcommand{\myRunningTitle}{A statistical perspective on {\sc flock}}
+\renewcommand{\RunningTitle}{\myRunningTitle}
 

### 2014-08-08 compile-documents.sh

@@ -1,0 +1,28 @@
+
+
+# this is a shell script that will compile the documents 
+# associated with this manuscript.
+# this should be run in the ./tex directory
+
+if [ $(basename $PWD) != "tex" ]; then
+	echo "Looks like you are not running $0 from the right directory.
+Your working directory must be the \"tex\" directory within the
+\"flock-comment\" directory.  Aborting..." > /dev/stderr
+exit 1
+fi
+
+
+# first, LaTeX the two column version of the manuscript.
+latexmk -pdf  flock-comment-main.tex  || echo "latexmk returned error compiling flock-comment-main.tex"
+
+# then LaTeX the one-column submission format version
+latexmk -pdf  flock-comment-one-column-main.tex  || echo "latexmk returned error compiling flock-comment-one-column-main.tex"
+
+
+echo; echo; echo
+echo "Check it out! If everything worked, then you have two
+compiled documents ready to go:
+   flock-comment-one-column-main.pdf   <--  version for submission
+   flock-comment-main.pdf              <--  two column version for easy reading
+   
+"

### 2014-08-08 tex/abstract.tex
Incorporate Kristen's comments on manuscript.

@@
 confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
-model and uncorrelated allele frequency prior, albeit with apparently more variability 
+model and uncorrelated allele frequency prior with more variability 
 from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 

### 2014-08-08 tex/main-body-text.tex
Incorporate Kristen's comments on manuscript.

@@
 
-Recently, several colleagues suggested we try using the software {\sc flock} \citep{Duc&Tur2009},
-for analyzing a number of the data sets from our laboratories.
-{\sc flock} is billed as a ``non-Bayesian method [that] 
+The recently introduced software {\sc flock} \citep{Duc&Tur2009} has
+been described
+as a ``non-Bayesian method [that] 
 differs substantially from previous 
-clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. As described
-in a later paper, {\sc flock} ``does not sample the space of partitions through small 
+clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. As reported
+in a later paper, ``{\sc flock} is very different from  other clustering 
+programs. It does not sample the space of partitions through small 
 random step walks as in MCMC, and it does not try to optimize some 
@@
 
-In trying to understand the behavior of {\sc flock} we identified 
-a close relationship between {\sc flock}'s algorithm and the program
-{\sc structure}.  In other words, although the algorithm in {\sc flock}
+However, here we identify 
+the close relationship between {\sc flock}'s algorithm and the program
+{\sc structure}.  Although the algorithm in {\sc flock}
 is, as the authors point out, entirely deterministic, it can be interpreted as a 
@@
 \begin{center}
-	 % NOTE. Eric made Flock-Fig1.jpg as a Preview because Flock-Fig1.pdf
-	 % ate up too many resources while editing the document.  For final 
-	 % production, it should be changed back to Flock-Fig1.pdf
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1-GreyScale.pdf} % Flock-Fig1.jpg}  
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}   % Use this if we want it grayscale -> Flock-Fig1-GreyScale.pdf  
     \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
@@
 genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
-a full-featured program with many options that is widely-used, well-tested, and reasonably well-understood. 
-We've shown here that {\sc flock}, a program that uses a deterministic algorithm, can be regarded
-as providing approximate, maximum-{\em a-posteriori} inference from one of the models implemented in 
-{\sc structure}. Given the similarities between the two methods, we expect that {\sc flock}
-and {\sc structure} (using the no-admixture model with non-correlated allele frequencies) will deliver
-similar results.  We observed such similarity in comparing results from {\sc flock} and {\sc structure} 
-on a real data set from \textit{O. mykiss}
+a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
+We've shown here that {\sc flock} is essentially equivalent to the program {\sc structure} using the no-admixture model with non-correlated allele frequencies.
+In comparing results from {\sc flock} and {\sc structure}
+on a real data set from {\em O. mykiss} we find that the two programs give similar results, only {\sc flock}'s results at higher values of $K$ appear to be more variable.
 
 Though we observed similar results
-between the two programs at low values of $K$, {\sc flock}'s ``best-runs'' solutions became somewhat inconsistent at higher values of $K$, though most solutions were still similar to {\sc structure}'s.
+between the two programs at low values of $K$, {\sc flock}'s `best-run' solutions became somewhat inconsistent at higher values of $K$.
 This is likely a consequence of the rapidly increasing size of the search space over possible 
@@
 The peak that  {\sc flock}'s deterministic algorithm finds is entirely determined by the initial allocation of 
-individuals, as it has no way to visit areas of the surface with smaller likelihood values.
+individuals, as it is a deterministic algorithm.
 If the number of initial allocations does not increase
@@
 ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
-of results from {\sc flock} if no non-zero plateaus are encountered.
-Of course, the problem of getting caught in local modes is not unique to {\sc flock}---{\sc structure} itself clearly can wind up sampling predominantly from a fairly isolated part of the space, as evidenced by the different results obtained from different runs of {\sc structure}.
+of results from {\sc flock} if no non-zero plateaus are encountered. The problem of getting 
+caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
+different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.
 
@@
 as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
-clusters. In short,
+clusters. Though
 it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
-constitutes a reasonable comparison with respect to run length and number of runs.
+constitutes a reasonable comparison with respect to run length and number of runs, {\sc flock} did not carry a distinct run-time advantage over {\sc structure} in our analyses.
 
@@
 a particular $K$,  similar---but not identical---partitions
-will provide no support to a given $K$. If using {\sc flock}, we advise running
+will provide no support to a given $K$. Given {\sc flock}'s tendency to converge to 
+different solutions with large $K$ and $n$, it seems that 
+this method for estimating $K$ is unlikely to be reliable or workable in large, complex
+problems.  One possible patch for this problem would be running
 multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
-{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, to determine
-if there are many similar-but-not-identical partitions found amongst {\sc flock}'s replicate 
-solutions.
+{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, and then 
+determine plateaus amongst {\sc flock}'s replicate solutions not by identical LLOD, but by visual inspection of 
+partitions that
+are very similar but are perhaps not identical; however such an approach would be labor-intensive
+and would rely on arbitrary, difficult-to-reproduce decisions.  It might be better to use {\sc structure} in such cases.
+Estimating $K$ is a hard problem---especially in real populations that seldom conform to the assumptions of genetic clustering models---and should be undertaken with caution, regardless of the method 
+used.
 
@@
 its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
-of individuals in the sample and thus may be a more appropriate model than the no-admixture model
-(or {\sc flock}) when the sample contains individuals that are admixed between populations.  A further advantage
+of individuals in the sample and thus may be a more appropriate model than the no-admixture model,
+or {\sc flock}, when the sample contains individuals that are admixed between populations.  A further advantage
 of the with-admixture model is that it allows inferences for the population of origin of individual alleles at 

### 2014-08-08 tex/main-body-text.tex
Minor edit

@@
 
-We should point out that our comparison was between {\sc flock} and the {\em no-admixture} model in 
+We made our comparison between {\sc flock} and the {\em no-admixture} model in 
 {\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 

### 2014-08-09 tex/main-body-text.tex
Even more minor edits

@@
 In comparing results from {\sc flock} and {\sc structure}
-on a real data set from {\em O. mykiss} we find that the two programs give similar results, only {\sc flock}'s results at higher values of $K$ appear to be more variable.
+on a real data set from {\em O. mykiss} we find that the two programs give similar results, only, at values of
+$K>4$, {\sc flock}'s results appear to be more variable.
 
@@
 it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
-constitutes a reasonable comparison with respect to run length and number of runs, {\sc flock} did not carry a distinct run-time advantage over {\sc structure} in our analyses.
+constitutes a reasonable comparison with respect to run length and number of runs, our results indicate that {\sc flock} did not carry a distinct run-time advantage over {\sc structure}.
 
@@
 are very similar but are perhaps not identical; however such an approach would be labor-intensive
-and would rely on arbitrary, difficult-to-reproduce decisions.  It might be better to use {\sc structure} in such cases.
-Estimating $K$ is a hard problem---especially in real populations that seldom conform to the assumptions of genetic clustering models---and should be undertaken with caution, regardless of the method 
-used.
+and would rely on arbitrary, difficult-to-reproduce decisions.  It probably better to use {\sc structure} in such cases, keeping in mind, of course, that
+estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
+used. This is especially true in real populations that do not conform to the assumptions of genetic clustering models.
 
@@
 \citet{Ste&Don2000} to improve upon the algorithm dramatically.  Here we have identified the {\sc flock}
-algorithm as a limiting case of simulating annealing upon {\sc structure}'s no-admixture model. We hope this will
+algorithm as a limiting case of simulating annealing upon {\sc structure}'s no-admixture model with uncorrelated
+allele frequencies. We hope this will
 help users to understand and interpret {\sc flock}'s behavior and will provide insights that may be useful 

### 2014-08-13 tex/main-body-text.tex
Made final edits to manuscript and readme.

@@
 \subsection*{{\sc flock}}
-Here we translate the recipe given by \citeauthor{Duc&Tur2009} for the 
+Here we translate the procedure given by \citeauthor{Duc&Tur2009} for the 
 {\sc flock} algorithm into a specification in terms of the variables
@@
 from {\sc flock}  are represented by a 
-gray-scale vertical bar.
-The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same `best run.'}
+%gray-scale
+vertical bar.
+The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same visually indistinguishable `best run.'}
     \label{fig:qplots}
@@
 between the two programs at low values of $K$, {\sc flock}'s `best-run' solutions became somewhat inconsistent at higher values of $K$.
+This observation was consistent with \citet{Duc&Tur2012}'s comparison of the 
+two programs using simulated data (Fig. 4 in \citet{Duc&Tur2012})
+for $K=4 \& 8, L=10$. 
 This is likely a consequence of the rapidly increasing size of the search space over possible 
@@
 as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it less likely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
+possible partitions making it less likely that many (if any at all) will converge to the same local mode. 
+This coupled with the fact that there is 
 ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
@@
 different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.
+By increasing the number of loci used in their simulations \citet{Duc&Tur2012} showed, surprisingly, that 
+{\sc flock} did converge on $K=8$, but held no advantage in inference of $K$ over {\sc structure}. We suspect that this 
+resulted from the confined search space defined
+by the small number of individuals (n=240) used in the analysis. 
+%They simulated data using 500 individuals per population but only used 30 of them in the analysis.
+% if you have more loci than individuals per population with high Fst between groups it does great!
 
@@
 {\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
-one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of {\sc structure} using 
+one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of 
+the admixture and correlated allele frequency model in {\sc structure} using 
 a 50,000 sweep burn-in followed 
@@
 help users to understand and interpret {\sc flock}'s behavior and will provide insights that may be useful 
-to {\sc flock}'s authors if they further develop their algorithm.
+in the potential development of the {\sc flock} algorithm.

### 2014-08-15 tex/main-body-text.tex
Minor edits and tweaks.

@@
 \subsection*{{\sc flock}}
-Here we translate the procedure given by \citeauthor{Duc&Tur2009} for the 
+Here we translate the procedure given by \citet{Duc&Tur2009} for the 
 {\sc flock} algorithm into a specification in terms of the variables
@@
 a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
-We've shown here that {\sc flock} is essentially equivalent to the program {\sc structure} using the no-admixture model with non-correlated allele frequencies.
+We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model with non-correlated allele frequencies within the program {\sc structure}.
 In comparing results from {\sc flock} and {\sc structure}
@@
 This observation was consistent with \citet{Duc&Tur2012}'s comparison of the 
-two programs using simulated data (Fig. 4 in \citet{Duc&Tur2012})
-for $K=4 \& 8, L=10$. 
+two programs using simulated data 
+for $K=4$ \& 8, $L=10$ (Fig. 4 in \citet{Duc&Tur2012}). 
 This is likely a consequence of the rapidly increasing size of the search space over possible 
@@
 different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.
-By increasing the number of loci used in their simulations \citet{Duc&Tur2012} showed, surprisingly, that 
-{\sc flock} did converge on $K=8$, but held no advantage in inference of $K$ over {\sc structure}. We suspect that this 
-resulted from the confined search space defined
-by the small number of individuals (n=240) used in the analysis. 
+By increasing the number of loci used in their simulations, \citet{Duc&Tur2012} showed  that 
+{\sc flock} was able to converge in a simulated data set with $K=8$, but held no advantage in inference of $K$ over {\sc structure}.
+That simulated data set, with $n=240$ was considerably smaller and less complex than the real data investigated here.
 %They simulated data using 500 individuals per population but only used 30 of them in the analysis.
@@
 are very similar but are perhaps not identical; however such an approach would be labor-intensive
-and would rely on arbitrary, difficult-to-reproduce decisions.  It probably better to use {\sc structure} in such cases, keeping in mind, of course, that
+and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} in such cases, keeping in mind, of course, that
 estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
-used. This is especially true in real populations that do not conform to the assumptions of genetic clustering models.
+used. This is especially true in real populations which do not conform to the assumptions of genetic clustering models.
 
@@
 or {\sc flock}, when the sample contains individuals that are admixed between populations.  A further advantage
-of the with-admixture model is that it allows inferences for the population of origin of individual alleles at 
+of the with-admixture model is that it allows inference of the population of origin of individual alleles at 
 different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 

### 2014-08-20 tex/flock-comment-letter-to-editor.tex
add a letter to the editor.

\documentclass[11pt]{letter}
\usepackage{graphicx}

\textwidth = 6.5 in
\textheight = 10 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = -.75 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.2in
\parindent = 0.0in

\date{}

\address{\mbox{}\vspace*{-1.0in}
\mbox{}\hfill
\includegraphics*{./images/NMFS_letterhead.pdf}\\}

\signature{Eric C. Anderson \\ \mbox{} \\ eric.anderson@noaa.gov\\ ph.~831-420-3983}

\begin{document}

\begin{letter}{ \today

Editor \\
Molecular Ecology Resources\\
}

\opening{Dear Editor:}
This letter accompanies our submission of, ``Interpreting the {\sc flock} algorithm from a statistical perspective,'' by Eric C. Anderson and Patrick D. Barry.  In this short comment we show that the method employed in the software {\sc flock} (published in {\em Molecular Ecology Resources} in 2009) can be identified as a special case of simulated annealing upon one of the models implemented in the software {\sc structure}.

In two publications {\sc flock}'s authors have been somewhat outspoken in asserting that {\sc flock} is very different from {\sc structure}. Indeed, we embarked upon this project because numerous colleagues were recommending we try the program {\sc flock} in cases where {\sc structure} did not reveal much population structure.  Upon a close reading of the mathematical details of {\sc flock}'s algorithm, however, it was clear that {\sc flock} is very similar to {\sc structure}.  We describe this from a mathematical perspective and then perform analyses on a real data set to confirm it.  These analyses additionally expose some very real problems suffered by the {\sc flock} approach to estimating $K$, the number of clusters.  While we feel our findings would easily substantiate a position that there were few, if any, reasons to use {\sc flock} instead of {\sc structure}, we don't assert that point strongly in the paper. Rather, in consideration of the work that has been done to develop {\sc flock}, we try to adopt a neutral tone, stating our mathematical and simulation findings as plainly as possible and allowing readers to make up their own minds.  Of course, if the editors felt that a stronger stance were appropriate, we could provide that, as well.

We suspect that {\sc flock} is being recommended to many others in our field as, ``very different from other clustering programs," which is something of a shame given that {\sc structure} is a firmly established and well-tested analytical tool in our field.  Consequently, we find our paper very relevant to the readership of {\em Molecular Ecology Resources}.

A draft of this paper was sent to Julie Turgeon (the corresponding author on the original {\sc flock} paper) a little over one month ago, inviting her and her coauthor to provide comments and/or recommend any changes, but she has chosen to respond after submission.

\closing{Best regards,}
\setlength{\topmargin}{0in}
\textheight = 9 in
\end{letter}

 \end{document}

### 2014-08-20 tex/main-body-text.tex
Add acknowledgments and fix "deterministic" statements.
Closes #2 and #3.

@@
 {\sc structure}.  Although the algorithm in {\sc flock}
-is, as the authors point out, entirely deterministic, it can be interpreted as a 
+is, as the authors point out, deterministic (apart from a random initial allocation of 
+individuals to cluster), it can be interpreted as a 
 limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
@@
 be multimodal---possessing numerous local peaks and troughs.
-The peak that  {\sc flock}'s deterministic algorithm finds is entirely determined by the initial allocation of 
+The peak that  {\sc flock}'s algorithm finds is entirely determined by the initial, random, allocation of 
 individuals, as it is a deterministic algorithm.
@@
 help users to understand and interpret {\sc flock}'s behavior and will provide insights that may be useful 
-in the potential development of the {\sc flock} algorithm.
+in the potential development of {\sc flock} or other clustering algorithms.
+
+\section*{Acknowledgments}
+We are grateful to Robin Waples for helpful comments on a draft of this paper.

### 2014-08-21 tex/main-body-text.tex
Add Kristen to acknowledments
caught one thing on final read through

@@
 $K$ increased so did the number of solutions
-that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed---all intial 
+that {\sc flock} found (Fig 1). For all groups of runs for $K > 2$, no plateaus were observed---all intial 
 allocations resulted in solutions that differed. 
@@
 one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
-for a $K\geq 2$, the plateau record varied greatly among each
+for a $K > 2$, the plateau record varied greatly among each
 group for $K = 2$ (plateau sequences: Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
@@
 \section*{Acknowledgments}
-We are grateful to Robin Waples for helpful comments on a draft of this paper.
+We are grateful to Kristen Ruegg and Robin Waples for helpful comments on a draft of this paper.

### 2014-08-21 tex/abstract.tex
Fixed some mathematical sections then submitted.
On a last reading I found some bone-headed errors in
the mathematical section --- I left the \ell subscript
on things when it shouldn't have been.  So, I cleaned
that up and fixed it.

Then included referees in the cover letter and fixed a
few typos and submitted it.

This is the version that was submitted.

@@
 allele frequency priors. The computational routine in {\sc flock} is 
-equivalent to searching for the maximum-{\em a-posteriori} (MAP) 
+equivalent to searching for the maximum-{\em a-posteriori}
 estimate of this restricted {\sc structure} model via a simulated 

### 2014-08-21 tex/flock-comment-letter-to-editor.tex
Fixed some mathematical sections then submitted.
On a last reading I found some bone-headed errors in
the mathematical section --- I left the \ell subscript
on things when it shouldn't have been.  So, I cleaned
that up and fixed it.

Then included referees in the cover letter and fixed a
few typos and submitted it.

This is the version that was submitted.

@@
 We describe this from a mathematical perspective and then perform analyses on a real data set to confirm it.
-These analyses additionally expose some very real problems suffered by the {\sc flock} approach to estimating $K$, the number of clusters.
+These analyses additionally expose some real problems that make it difficult to apply the {\sc flock} approach to estimating $K$, the number of clusters.
 While we feel our findings would easily substantiate a position that there were few, if any, reasons to use {\sc flock} instead of {\sc structure}, we don't assert that point strongly in the paper.
-Rather, in consideration of the work that has been done to develop {\sc flock}, we try to adopt a neutral tone, stating our mathematical and simulation findings as plainly as possible and allowing readers to make up their own minds.
-Of course, if the editors felt that a stronger stance were appropriate, we could provide that, as well.
+Rather, in consideration of the work that has been done to develop {\sc flock} and in the spirit of us all being molecular ecologists working together to understand the tools we use, we try to adopt a neutral tone, stating our mathematical and simulation findings as plainly as possible and leaving it to the reader to make up his or her own mind about which programs to use.
+Of course, if the editors felt that a stronger stance were appropriate, we could provide that in a revised draft.
 
@@
 A draft of this paper was sent to Julie Turgeon (the corresponding author on the original {\sc flock} paper) a little over one month ago, inviting her and her coauthor to provide comments and/or recommend any changes, but she has chosen to respond after submission.
+
+We have given some thought to who would be good referees for this paper.
+Given the technical and terse nature of our comparison between {\sc flock} and {\sc structure}, it would be advantageous to enlist referees who are very familiar with the model underlying {\sc structure}.
+Obvious choices would be Jonathan Pritchard, Matthew Stephens, and John Novembre.
+However, since these are very busy investigators running large human statistical-genetic labs, other possibilities would be Melissa Hubisz, Bruce Rannala, Greg Wilson (author of BayesAss), Philip Andolfatto, John Hulsenbeck, Jerome Pella, or Michelle Masuda.
+Specifically in the molecular ecology realm, we think Oscar Gaggiotti would make a very good referee.
+
+Daniel Falush would be the most appropriate editor for this manuscript because he has contributed to the development of {\sc structure} and is extremely familiar with the underlying model.
 

### 2014-08-21 tex/main-body-text.tex
Fixed some mathematical sections then submitted.
On a last reading I found some bone-headed errors in
the mathematical section --- I left the \ell subscript
on things when it shouldn't have been.  So, I cleaned
that up and fixed it.

Then included referees in the cover letter and fixed a
few typos and submitted it.

This is the version that was submitted.

@@
 BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
-\citep{Coranderetal2004}. An overview of the similarities between these methods can be 
+\citep{Coranderetal2004}. An overview of the similarities between several of these methods can be 
 found in \citet{Anderson2009PGAC}.
@@
 is, as the authors point out, deterministic (apart from a random initial allocation of 
-individuals to cluster), it can be interpreted as a 
+individuals to clusters), it can be interpreted as a 
 limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
@@
 (\ie $Z_i=k$), depends on the 
-cluster memberships and genotypes  of all the remaining individuals,
+cluster memberships and genotypes at locus $\ell$ of all the remaining individuals,
 which we denote by  $Z_{(-i)}$ and $Y_{(-i)\ell}$, respectively.
@@
 found in individuals {\em other than individual $i$} that currently
-belong to cluster $k$. Thus, calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
+belong to cluster $k$. Thus, calculating the product of (\ref{eq:cdm}) across all the loci
+($\ell = 1,\ldots,L$) gives the full conditional multilocus genotype probability 
+for individual $i$:
+\begin{eqnarray}
+\lefteqn{P(Y_{i}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)}) =}  \label{eq:cdm-multi-prob} \\
+& &  \prod_{\ell=1}^L P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell}). \nonumber
+\end{eqnarray}
+Computing (\ref{eq:cdm-multi-prob}) for 
+for each value of $Z_i=k \in 
 \{1,\ldots,K\}$ 
@@
 \begin{equation}
-P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})~~,~~k=1,\ldots,K.
+P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})~~,~~k=1,\ldots,K.
 \label{eq:fc}
@@
 be reallocated to whichever cluster gives the highest value to 
-$P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$, which is exactly the probability
-defined in (\ref{eq:cdm}).  
+$P(Y_{i}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)})$, which is 
+exactly the probability
+defined in (\ref{eq:cdm-multi-prob}).  
 If \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
@@
 on maximizing a likelihood that, although it may not be formally identical to
-that in (\ref{eq:cdm}), is nearly identical to it.
+the probability defined in (\ref{eq:cdm-multi-prob}), is nearly identical to it.
 
@@
 {\sc structure} model without admixture assumes a uniform prior over 
-the $K$ different clusters, $P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$
-is exactly proportional to $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$, which 
+the $K$ different clusters, $P(Y_{i}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)})$
+is exactly proportional to $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})$, which 
 we have shown is the
@@
 \[
-P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
+P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})
 \]
@@
 \begin{equation}
-\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell}),
+\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)}),
 \end{equation}
-\ie the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$
+\ie the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})$
 (or, as stated above, a nearly identical function of $k$, depending on how
@@
 \[
-\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})\biggr]^\beta
+\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)})\biggr]^\beta
 \]
@@
 $K$ increased so did the number of solutions
-that {\sc flock} found (Fig 1). For all groups of runs for $K > 2$, no plateaus were observed---all intial 
+that {\sc flock} found (Fig 1). For all groups of runs for $K > 2$, no plateaus were observed---all initial 
 allocations resulted in solutions that differed. 
@@
 The peak that  {\sc flock}'s algorithm finds is entirely determined by the initial, random, allocation of 
-individuals, as it is a deterministic algorithm.
+individuals, as the algorithm proceeds deterministically after the initial random allocation.
 If the number of initial allocations does not increase

### 2014-10-29 tex/main-body-text.tex

@@
 \section*{Comparison of {\sc structure} and {\sc flock} results}
-
-We used a published dataset of 2596 steelhead, \textit{Oncorhynchus mykiss},
-genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal} to compare the programs
-{\sc flock} and {\sc structure}.  {\sc flock} 
-was run six times using the default setting for all 
+Data sets with known $K$ were simulated
+(code provided on GitHub repository https://github.com/eriqande/flock-comment)
+and evaluated in the programs {\sc flock} and {\sc structure} 
+for the known value.The first dataset simulated was composed of five populations of 
+sizes 500, 500, 800, 1100, and 1100 comprised of individuals 
+with 15 loci and an average pairwise $F_{ST} = 0.03$ between populations. 
+The second and third dataset were simulated with population sizes
+250, 250, 400, 550, and 550 and average pairwise $F_{ST} = 0.02 and 0.03$
+respectively. {\sc flock} was run nine times using the default setting for all 
 run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
-LLOD threshold = 0) for values of $K$ from 2 to 6. The stopping point of 
-$K=6$ was chosen after four consecutive values of $K$
-yielded no plateau (see below). For each value of $K$ from 2 to 6, the no-admixture, 
-uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times with 
-a 50,000 sweep burn-in followed by a 150,000 sweep sample from the posterior. 
+LLOD threshold = 0) for the known value of $K$. The no-admixture, 
+uncorrelated-allele-frequency model implemented in {\sc structure} was run nine times with 
+a 5,000 sweep burn-in followed by a 20,000 sweep sample from the posterior. 
+% Looking back some of the runs were under different conditions, not huge differences. 
+% But 1000 burn-in and 5,000 sweep look about the same
 All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
@@
 normalized likelihood values (the exponentiated log-likelihood value for an individual's 
-membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
+membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
+longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
 These values can be compared to the $q_i$ values
@@
 of individuals. 
-\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
-strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
+\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$.\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
 $\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
@@
 \subsection*{Results} 
-Run times for both programs were relatively short, albeit not trivial. All runs of {\sc flock} 
-completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
-All runs of {\sc structure} were completed in 
-486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
-The processing time of each algorithm increased as the number or 
-putative clusters increases. Increasing
-$K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
-for {\sc structure} increased by a factor of 1.6.
+Run times for both programs were relatively short, albeit in the case of {\sc flock} 
+non-trivial. {\sc flock} averaged 67.3 minutes
+for each run of the first simulated dataset, while each group of 
+9 iterations for {\sc structure} averaged 18 minutes. CALCULATE
+TIME FOR OTHER SIMULATIONS
 
-{\sc flock} and {\sc structure} provide similar results in 
-terms of how individuals are allocated among runs at low values of $K$, 
-while at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
-multiple other solutions. For all six groups 
-of runs with $K = 2$ to 4, {\sc flock} obtained a nearly identical clustering of individuals 
-(See supplementary materials) none of which differed substantially from {\sc structure}'s results
-(Fig.~\ref{fig:qplots}). Of the `best runs' for the six groups for $K$ = 2, two  
-runs yielded identical clustering (same mean LLOD score) while all other runs 
-differed by at most the allocation of 13 individuals. As 
-$K$ increased so did the number of solutions
-that {\sc flock} found (Fig 1). For all groups of runs for $K > 2$, no plateaus were observed---all initial 
-allocations resulted in solutions that differed. 
-{\sc flock} still produced estimates of the LLOD by individual for the 
-'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
-equal to one. Many of these solutions are, however, surprisingly similar to
-the partitions identified by {\sc structure}. Both programs idenfity clusters 
-which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
-although less reliably so in the case of {\sc flock}.
+{\sc flock} and {\sc structure} provide nearly identical results in 
+terms of how individuals are allocated among runs for 
+all simulations (Fig.~\ref{fig:Sims}). As $F_{ST}$ decreases 
+between populations the solutions that {\sc flock} are more variable. 
+The average Euclidean distance between the known $q_i$ value and 
+the infered value from each program increases as the genetic 
+differentiation among populations decreases. {\sc flock} had an 
+smaller mean Euclidean distance over all runs for simulations 
+1 and 2, but had much larger variation in simulation 3 (Fig.~\ref{fig:ED}).
+%calculate Euclidean distance between flock and structure values.
+
+There was very little variation in the 'best-run' returned by {\sc flock}. 
+From the nine different sets of runs for the first simulated dataset, 
+three groups of runs ended with 
+identical clusters (runs 2,8 \& 9, runs 1 \& 3, and runs 4 \& 7). 
+Of the other 31 pairwise comparisions that were not identical
+15 differed by the placement of 1 indvidual, 11 differed by the placement
+of 2 individuals, and 5 differed by the placement of 3 individuals.
 
 Despite the similar allocation of individuals into $K$ clusters, the 
-plateau analysis to estimate $K$ in {\sc flock} leads to different conclusions among groups of runs, as well as 
-one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
-for a $K > 2$, the plateau record varied greatly among each
-group for $K = 2$ (plateau sequences: Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
-Run4 = 2,3,4,2,4,3; Run5 = 2,2,5,5,3; and; Run6 = 2,2,2,2,2,3,3). The plateau analysis 
-procedure suggested by \citet{Duc&Tur2012} leads to
-either an estimate of K $\geq$ 2 (Run1) or {\em undecided} as a consequence of no plateau 
-lengths being greater than 6. The ($\ln P(D)$) from {\sc structure}
-was largest at $K$ = 6 which was further supported by the $q_i$ plots, however,
-the $\Delta K$ method supported a $K$ of 3.
+plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$. 
+For the first simulated dataset plateaus of varioius lengths were observed, 
+but none greater than 5 while no plateaus were observed for either second or 
+third simulated datasets.
 
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}   % Use this if we want it grayscale -> Flock-Fig1-GreyScale.pdf  
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Simulations_Fig1.pdf}   % Use this if we want it color (I'll make greyscale later)  
     \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
-from {\sc flock} made with the program {\sc distruct}.
-Each horizontal plot represents an individual run of the program {\sc structure}, or the `best run' from a series of runs in {\sc flock}.
+of {\sc flock} made with the program {\sc distruct}.
+Each horizontal plot represents one of nine individual runs of the program {\sc structure}, or the `best run' from a series of nine runs in {\sc flock} at $K=5$.
 Within each plot
@@
 %gray-scale
-vertical bar.
-The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same visually indistinguishable `best run.'}
-    \label{fig:qplots}
+vertical bar. Simulation 1 was composed of populations of 
+size 500, 500, 800, 1100, and 1100 with an average pairwise $F_{ST} = 0.03$ between populations. 
+The second and third dataset were simulated with population sizes
+250, 250, 400, 550, and 550 and average pairwise $F_{ST} =$ 0.02 and 0.03
+respectively.}
+    \label{fig:Sims}
 \end{center}
@@
 
-%\begin{figure}
-%\begin{center}
-% \includegraphics[width=\columnwidth]{images/Figures-Pat/Fig2.pdf}
-%    \caption{Individual \textit{$q_i$} values plotted for each cluster for $K=5$. 
-%{\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
-% attributed to them.}
-%    \label{fig:qscatter}
-%\end{center}
-%\end{figure}
-
-To investigate the ability of the arrangement of individuals within the reference 
-populations to move around the partition space we varied the number of iterations that 
-each run goes through to find the ``optimal'' allocation of individuals. Unfortuantely 
-only the reallocation matrix for the `best run' is given as output. Increasing the 
-number of iterations did not affect the overall inference of $K$. We 
-observed that often during the last iterations individuals would alternate between 
-reference groups. Increasing the log likelihood 
-difference threshold for reallocation to 0.18 (requiring at least a 1.5 times 
-higher likelihood to reallocate the individual) minimized this, but did not affect the 
-inference of $K$ (lower bound estimate of $K = 2$).
+\begin{figure*}
+\begin{center}
+    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/EucliDist.pdf}   % Use this if we want it color (I'll make greyscale later)  
+    \caption{Eucledian distance between true $q_i$ value and inferred value from {\sc structure} and {\sc flock} .}
+    \label{fig:ED}
+\end{center}
+\end{figure*}
 
@@
 In comparing results from {\sc flock} and {\sc structure}
-on a real data set from {\em O. mykiss} we find that the two programs give similar results, only, at values of
-$K>4$, {\sc flock}'s results appear to be more variable.
+on simulated datasets with varying levels of differnetiation among populations
+we find that the two programs give similar results.
 

### 2014-10-29 tex/main-body-text.tex
changed a figure after looking at distance between qi values from programs
corrected minor typo for Fst value

@@
 The second and third dataset were simulated with population sizes
-250, 250, 400, 550, and 550 and average pairwise $F_{ST} = 0.02 and 0.03$
+250, 250, 400, 550, and 550 and average pairwise $F_{ST} = 0.02 and 0.01$
 respectively. {\sc flock} was run nine times using the default setting for all 
@@
 The second and third dataset were simulated with population sizes
-250, 250, 400, 550, and 550 and average pairwise $F_{ST} =$ 0.02 and 0.03
+250, 250, 400, 550, and 550 and average pairwise $F_{ST} =$ 0.02 and 0.01
 respectively.}
@@
 \begin{center}
-    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/EucliDist.pdf}   % Use this if we want it color (I'll make greyscale later)  
+    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/ED_fig2.pdf}   
+% HIDEOUS GRAPH, we will talk and I might make one with actual distributions! 
     \caption{Eucledian distance between true $q_i$ value and inferred value from {\sc structure} and {\sc flock} .}

### 2014-12-02 tex/main-body-text.tex
Changed results, with new figs, cleaned figure directory out

--- 2014-10-29 tex/main-body-text.tex
@@
 \section*{Comparison of {\sc structure} and {\sc flock} results}
-Data sets with known $K$ were simulated
-(code provided on GitHub repository https://github.com/eriqande/flock-comment)
+Data sets were simulated (code provided on GitHub repository 
+https://github.com/eriqande/flock-comment)
 and evaluated in the programs {\sc flock} and {\sc structure} 
-for the known value.The first dataset simulated was composed of five populations of 
-sizes 500, 500, 800, 1100, and 1100 comprised of individuals 
-with 15 loci and an average pairwise $F_{ST} = 0.03$ between populations. 
-The second and third dataset were simulated with population sizes
-250, 250, 400, 550, and 550 and average pairwise $F_{ST} = 0.02 and 0.03$
-respectively. {\sc flock} was run nine times using the default setting for all 
+for the known value ($K=5$). Each dataset was composed of five populations of 
+sizes 250, 250, 400, 550, and 550 comprised of individuals 
+with 15 loci and an average pairwise $F_{ST} = 0.03, 0.02, 0.01, and 0.005$ among
+populations. {\sc flock} was run nine times at $K=5$ with the default setting for all 
 run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
-LLOD threshold = 0) for the known value of $K$. The no-admixture, 
-uncorrelated-allele-frequency model implemented in {\sc structure} was run nine times with 
-a 5,000 sweep burn-in followed by a 20,000 sweep sample from the posterior. 
-% Looking back some of the runs were under different conditions, not huge differences. 
-% But 1000 burn-in and 5,000 sweep look about the same
-All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
+LLOD threshold = 0). The default model in {\sc structure}, the no-admixture and correlated
+allele frequency (NA-C) model, the no-admixture and non-correlated allle frequency (NA-NC) model, 
+and the no-admixture and non-correlated allele frequency (NA-NC) model 
+were run nine times with a 5,000 sweep burn-in followed by a 20,000 sweep sample 
+from the posterior. All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
 of RAM running 64-bit Windows 7.
@@
 of individuals. 
-\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$.\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
+\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
+\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
 $\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
@@
 
+%This section may get cut depending on how much we want to talk about estimating K.
 It is now common practice to use the approximate marginal likelihood ($\ln P(D)$) from {\sc structure}
@@
 Run times for both programs were relatively short, albeit in the case of {\sc flock} 
-non-trivial. {\sc flock} averaged 67.3 minutes
+non-trivial. {\sc flock} averaged 32.8 minutes
 for each run of the first simulated dataset, while each group of 
-9 iterations for {\sc structure} averaged 18 minutes. CALCULATE
-TIME FOR OTHER SIMULATIONS
+9 iterations for {\sc structure} averaged 14.7, 15.3 and 18.7 minutes
+for the NA-NC, NA-C, and A-C models respectively.
 
@@
 all simulations (Fig.~\ref{fig:Sims}). As $F_{ST}$ decreases 
-between populations the solutions that {\sc flock} are more variable. 
-The average Euclidean distance between the known $q_i$ value and 
-the infered value from each program increases as the genetic 
-differentiation among populations decreases. {\sc flock} had an 
-smaller mean Euclidean distance over all runs for simulations 
-1 and 2, but had much larger variation in simulation 3 (Fig.~\ref{fig:ED}).
-%calculate Euclidean distance between flock and structure values.
+between populations the solutions that {\sc flock} found were more variable.
+Of the nine runs of {\sc flock} for the first simulated dataset, between 0 and 22
+differences between the final arrangement of indivuals within the ($K=5$)
+clusters. Of the nine runs, four pairs of runs found the 
+same solution (runs 1 \& 2, 5 \& 6, 3 \& 7, and 4\&9). Run 8, which 
+had no plateaus, was the most different with an average of 19.86 differences.
+For the third simulated dataset, the number 
+of differences ranged from 482 to 911.  {\sc structure} showed less variation for 
+each set of simulations. Solutions for the first simulated dataset differed by the
+placement of between 0 and 4 individuals and between 5 and 17 for the third 
+dataset.
+%Create a figure comparing all of the values. All the text is somewhat hard to read
 
-There was very little variation in the 'best-run' returned by {\sc flock}. 
-From the nine different sets of runs for the first simulated dataset, 
-three groups of runs ended with 
-identical clusters (runs 2,8 \& 9, runs 1 \& 3, and runs 4 \& 7). 
-Of the other 31 pairwise comparisions that were not identical
-15 differed by the placement of 1 indvidual, 11 differed by the placement
-of 2 individuals, and 5 differed by the placement of 3 individuals.
+Plateau lengths greater than 0 were only obsered for the first simulated dataset. For each
+of the nine runs of {\sc flock} the plateau lengths varied from 0 to 4. In no analysis 
+did we observe a plateau lenght greater than 6. Despite the similar allocation of individuals into $K$ clusters, the 
+plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$.
 
-Despite the similar allocation of individuals into $K$ clusters, the 
-plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$. 
-For the first simulated dataset plateaus of varioius lengths were observed, 
 but none greater than 5 while no plateaus were observed for either second or 
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Simulations_Fig1.pdf}   % Use this if we want it color (I'll make greyscale later)  
+    \includegraphics[width=\textwidth]{images/Figures-Pat/Fig_AllSim.pdf}   % Use this if we want it color (I'll make greyscale later)  
     \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
@@
 %gray-scale
-vertical bar. Simulation 1 was composed of populations of 
-size 500, 500, 800, 1100, and 1100 with an average pairwise $F_{ST} = 0.03$ between populations. 
-The second and third dataset were simulated with population sizes
-250, 250, 400, 550, and 550 and average pairwise $F_{ST} =$ 0.02 and 0.03
-respectively.}
+vertical bar. Each dataset is comprised of five populations of sizes 550, 550, 400, 250, and 250 individuals. The 
+average pairwise $F_{ST}$ among populations was 0.03, 0.02, 0.01, and 0.005 for simulations 1, 2, 3, and 4 respecitively. 
+}
     \label{fig:Sims}
@@
 \begin{center}
-    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/EucliDist.pdf}   % Use this if we want it color (I'll make greyscale later)  
-    \caption{Eucledian distance between true $q_i$ value and inferred value from {\sc structure} and {\sc flock} .}
-    \label{fig:ED}
+    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/Loss.pdf}
+    \caption{Comparision of zero-One loss for the admixture and correlated allele frequency (A-C) model, 
+the no-admixture and correlated allele frequency (NA-C) model, the no-admixture and non-correlated 
+allle frequency (NA-NC) model, the no-admixture and non-correlated allele frequency (NA-NC) 
+model in {\sc structure} and the {\sc flock} model for each of the four simulated datasets.}
+    \label{fig:Loss}
 \end{center}

### 2015-01-10 tex/main-body-text.tex
Got behind in commits, added script to simulate data, run flockture and structure and then do some comps

@@
 
-\section*{Comparison of {\sc structure} and {\sc flock} results}
-Data sets were simulated (code provided on GitHub repository 
-https://github.com/eriqande/flock-comment)
-and evaluated in the programs {\sc flock} and {\sc structure} 
-for the known value ($K=5$). Each dataset was composed of five populations of 
-sizes 250, 250, 400, 550, and 550 comprised of individuals 
-with 15 loci and an average pairwise $F_{ST} = 0.03, 0.02, 0.01, and 0.005$ among
-populations. {\sc flock} was run nine times at $K=5$ with the default setting for all 
-run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
-LLOD threshold = 0). The default model in {\sc structure}, the no-admixture and correlated
-allele frequency (NA-C) model, the no-admixture and non-correlated allle frequency (NA-NC) model, 
-and the no-admixture and non-correlated allele frequency (NA-NC) model 
-were run nine times with a 5,000 sweep burn-in followed by a 20,000 sweep sample 
-from the posterior. All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
-of RAM running 64-bit Windows 7.
+\section*{Comparison of {\sc structure} and {\sc flock}}
+The implementation of {\sc flock} as a VBA procedure (macro) in Microsoft Excel precludes
+a rigorous evaluation of its behavior due to the inability to efficiently batch process. Consequenty, 
+we adapted the source code from {\sc structure} as describe above to create {\sc flockture}
+(code provided on GitHub repository https://github.com/eriqande/flockture) in order to accurately
+assess the strengths and weaknesses of the {\sc flock} algorithm. First, using simulated 
+data we demonstrated that {\sc flock} and {\sc flockture} are essentially equivalent programs.
+Second, we simulated and analyzed datasets with {\sc flockture} and three models in {\sc structure} 
+(the admixture and correlated allele frequency (A-C) model, 
+the no-admixture and correlated allele frequency (NA-C) model, 
+and the no-admixture and non-correlated allele frequency (NA-NC) model) to help better understand
+under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .
 
@@
 These values can be compared to the $q_i$ values
-returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
-The program {\sc clumpp} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
-while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
-
-\citet{Duc&Tur2012} describe a method for using {\sc flock} to estimate, $K$, the number of
+returned by {\sc structure}, which, in the no-admixture case, are posterior 
+probabilities of cluster membership.\citet{Duc&Tur2012} describe a method 
+for using {\sc flock} to estimate, $K$, the number of
 subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
-to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
-to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
+to each cluster) as well as the number of iterations. (In each iteration every individual 
+has the opportunity to be reallocated 
+to a different cluster.) At the conclusion of each run, the log-likelihood value for 
+membership to all $K$ clusters is computed
 and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
@@
 The plateau record is a list of how many runs converged to the same allocation
-of individuals. 
-\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
+of individuals. \citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
 \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
@@
 
-%This section may get cut depending on how much we want to talk about estimating K.
-It is now common practice to use the approximate marginal likelihood ($\ln P(D)$) from {\sc structure}
-or a derived quantity such as $\Delta K$ \citep{Evannoetal2005} to try to estimate $K$.
-While we take the stance that estimates of $K$ made with {\em any} unsupervised clustering algorithm
-from data on real (non-idealized) populations should always be interpreted and used
-cautiously, it is nonetheless instructive to compare \citet{Duc&Tur2012}'s method for
-estimating $K$ to results obtained using $\ln P(D)$ and $\Delta K$, and we do so, below.
+To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
+were simulated using a coalescent based program (code provided on GitHub repository 
+https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
+were comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
+and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
+populations. Both programs were run nine times at $K=5$ with the default setting for all 
+run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
+LLOD threshold = 0). We judged {\sc flockture} as an accurate representation of {\sc flock} based on the
+following criteria: (1) Clustering of individuals into corrected population as defined by a zero-one loss function, 
+(2) individual $q_i$ values, and (3) plateau records. All runs were performed on a JNCS D685 with an i7 3.40GHz 
+processor and 8GB of RAM running 64-bit Windows 7.
 
-\subsection*{Results} 
+Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
+we used  {\sc flockture} to analyze a many more datasets that encompassed a broader 
+range of population differentiation. Fourteen datasets were simulated as above (five populations 
+comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes)
+with average pairwise $F_{ST}$ values ranging between 0.11 and 0.005. 
+ {\sc flockture} was run with the default run parameters and each run of 
+{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
+from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
+and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
+was used to relabel clusters among different runs,
+while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+
+\subsection*{{\sc flock} and {\sc flockture}} 
+{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
+simulated datasets. The zero-one loss across all datasets had eqivalent results Fig.~\ref{fig:FvFloss}.
+When genetic differentiation is high both programs correctly apportion the majority of individuals and 
+as genetic differentiation degrades among populations both programs show corresponding declines in 
+assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
+substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
+values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
+for {\sc flock}. In an attempt to understand the deviations being observed we took a subset of the 
+first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
+{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
+individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
+with the same starting clusters, but with the reference population labels switched. While the final allocation
+of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
+between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
+be identical, but should have displayed equivalent behavior. Plateaus were only observed for the dataset 
+with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
+the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
+the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
+variability that we would expect from randomizing the initial grouping of individuals to 
+initiate the algorithms.
+
 Run times for both programs were relatively short, albeit in the case of {\sc flock} 
-non-trivial. {\sc flock} averaged 32.8 minutes
-for each run of the first simulated dataset, while each group of 
-9 iterations for {\sc structure} averaged 14.7, 15.3 and 18.7 minutes
-for the NA-NC, NA-C, and A-C models respectively.
-
-{\sc flock} and {\sc structure} provide nearly identical results in 
-terms of how individuals are allocated among runs for 
-all simulations (Fig.~\ref{fig:Sims}). As $F_{ST}$ decreases 
-between populations the solutions that {\sc flock} found were more variable.
-Of the nine runs of {\sc flock} for the first simulated dataset, between 0 and 22
-differences between the final arrangement of indivuals within the ($K=5$)
-clusters. Of the nine runs, four pairs of runs found the 
-same solution (runs 1 \& 2, 5 \& 6, 3 \& 7, and 4\&9). Run 8, which 
-had no plateaus, was the most different with an average of 19.86 differences.
-For the third simulated dataset, the number 
-of differences ranged from 482 to 911.  {\sc structure} showed less variation for 
-each set of simulations. Solutions for the first simulated dataset differed by the
-placement of between 0 and 4 individuals and between 5 and 17 for the third 
-dataset.
-%Create a figure comparing all of the values. All the text is somewhat hard to read
-
-Plateau lengths greater than 0 were only obsered for the first simulated dataset. For each
-of the nine runs of {\sc flock} the plateau lengths varied from 0 to 4. In no analysis 
-did we observe a plateau lenght greater than 6. Despite the similar allocation of individuals into $K$ clusters, the 
-plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$.
-
-but none greater than 5 while no plateaus were observed for either second or 
-third simulated datasets.
+non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
+over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
+general increase as population structure became more subtle, however {\sc flock} did not 
+demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
+% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
+% at later iterations leading to increased process times. Seems reasonable.
 
 \begin{figure*}
-\begin{center}
-    \includegraphics[width=\textwidth]{images/Figures-Pat/Fig_AllSim.pdf}   % Use this if we want it color (I'll make greyscale later)  
-    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
-of {\sc flock} made with the program {\sc distruct}.
-Each horizontal plot represents one of nine individual runs of the program {\sc structure}, or the `best run' from a series of nine runs in {\sc flock} at $K=5$.
-Within each plot
-$q_i$ values from {\sc structure} and the normalized likelihood values 
-from {\sc flock}  are represented by a 
-%gray-scale
-vertical bar. Each dataset is comprised of five populations of sizes 550, 550, 400, 250, and 250 individuals. The 
-average pairwise $F_{ST}$ among populations was 0.03, 0.02, 0.01, and 0.005 for simulations 1, 2, 3, and 4 respecitively. 
-}
-    \label{fig:Sims}
-\end{center}
+\centering
+\begin{minipage}{.45\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
+  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
+of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
+  \label{fig:FvFloss}
+\end{minipage}%
+\hfill
+\begin{minipage}{.45\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
+  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of dataset 1 
+with average pairwise $F_{ST} = $ 0.03. }
+  \label{fig:FvFqi}
+\end{minipage}
 \end{figure*}
 
+\subsection*{{\sc flockture} and {\sc structure}} 
+As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
+well differentiated populations. Both programs have less than 1\% assignment error for most runs 
+at values of $F_{ST}  \geq$ 0.0603 (Fig.~\ref{fig:FvSloss}). Infrequently, the 
+NA-NC and NA-C models in {\sc structure} grouped two populations as one and 
+assign individuals from another to two separate populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
+disappears as population differentiation declines. As the subtlety of genetic structure 
+increased loss for both programs increases. 
+Performance of the two programs as $F_{ST}$ approaches 0.0238 is relatively similar
+characterized by the A-C model in {\sc structure} having the largest loss, the NA-C and NA-NC models performing
+relatively similar and {\sc flockture} fluctuating around the results of the other three models. The scale of these 
+differences is extremely small - on the order of 10 individuals being misassigned. As genetic 
+differentiation continues to decline {\sc flockture}  showed consistently better performance 
+at levels of $F_{ST}$ between 0.0238 to 0.0229.  Again, these differences are on the scale of roughly 20 individuals
+or 1\%. In simulation 12 we observed the variation in {\sc flockture} results increase drastically (Fig.~\ref{fig:FvSloss}),
+and for all other simulated datasets {\sc structure} had a lower assignment error.
+
+The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau quickly 
+diminished as the amount of differentiation among populations decreased. For simulations 1-4 a single plateau
+greater than 6 was observed in every run. For simulation 5 only a single run yielded a plateau record 
+with a single plateau greater than six. For simulations 6 to 11 plateau records of various lengths were observed
+but all plateaus were less than 6. No plateaus were observed for datasets 12 through 20. Despite the lack of 
+plateaus observed for these datasets graphs of the $q_i$ values were suggestive of population structure within the 
+data (Fig.~\ref{fig:FvSdistruct}  Simulation 13). {\sc district} plots of the normalized likelihood values from
+{\sc flockture} were consistent and suggestive of population structure within the dataset until simulation 17
+after which the majority of the 'best runs' were substantially different in how they clustered individuals.
+
  \begin{figure*}
-\begin{center}
-    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/Loss.pdf}
-    \caption{Comparision of zero-One loss for the admixture and correlated allele frequency (A-C) model, 
-the no-admixture and correlated allele frequency (NA-C) model, the no-admixture and non-correlated 
-allle frequency (NA-NC) model, the no-admixture and non-correlated allele frequency (NA-NC) 
-model in {\sc structure} and the {\sc flock} model for each of the four simulated datasets.}
-    \label{fig:Loss}
-\end{center}
+\centering
+  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FvSloss1.pdf}
+  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
+  frequency (A-C) model, 
+the no-admixture and correlated allele frequency (NA-C) model, 
+and the no-admixture and non-correlated allele frequency (NA-NC) model in {\sc structure} and {\sc flockture}. 
+Average pairwise $F_{ST}$ 
+values between populations is listed above each graph.}
+  \label{fig:FvSloss}
+\end{figure*} % These graphs need to be redone so that text is legible and corrected for the NA-C NOT the A-NC 
+%model in structure. I probably had a transcription error in R. Check code to be sure.
+
+ \begin{figure*}
+\centering
+  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}
+  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
+  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
+ runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
+  frequency (A-C) model, 
+the no-admixture and correlated allele frequency (NA-C) model, 
+and the no-admixture and non-correlated allele frequency (NA-NC) model. Each simulation is comprised of
+ five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
+ indicated above each plot.}
+  \label{fig:FvSdistruct}
 \end{figure*}
+
+% haven't touched these yet...
 
@@
 a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
-We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model with non-correlated allele frequencies within the program {\sc structure}.
-In comparing results from {\sc flock} and {\sc structure}
+We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
+with non-correlated allele frequencies within the program {\sc structure}. In comparing results from {\sc flock} and {\sc structure} 
 on simulated datasets with varying levels of differnetiation among populations
@@
 By increasing the number of loci used in their simulations, \citet{Duc&Tur2012} showed  that 
-{\sc flock} was able to converge in a simulated data set with $K=8$, but held no advantage in inference of $K$ over {\sc structure}.
-That simulated data set, with $n=240$ was considerably smaller and less complex than the real data investigated here.
+{\sc flock} was able to converge in a simulated data set with $K=8$, but held no advantage in inference of $K$ 
+over {\sc structure}. That simulated data set, with $n=240$ was considerably smaller and less complex than the real data investigated here.
 %They simulated data using 500 individuals per population but only used 30 of them in the analysis.
@@
 it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
-constitutes a reasonable comparison with respect to run length and number of runs, our results indicate that {\sc flock} did not carry a distinct run-time advantage over {\sc structure}.
+constitutes a reasonable comparison with respect to run length and number of runs, our results indicate 
+that {\sc flock} did not carry a distinct run-time advantage over {\sc structure}.
 

### 2015-02-06 tex/main-body-text.tex
couple new figures. troubleshooting reproducability on another computer

@@
 \section*{Comparison of {\sc structure} and {\sc flock}}
-The implementation of {\sc flock} as a VBA procedure (macro) in Microsoft Excel precludes
-a rigorous evaluation of its behavior due to the inability to efficiently batch process. Consequenty, 
-we adapted the source code from {\sc structure} as describe above to create {\sc flockture}
-(code provided on GitHub repository https://github.com/eriqande/flockture) in order to accurately
-assess the strengths and weaknesses of the {\sc flock} algorithm. First, using simulated 
-data we demonstrated that {\sc flock} and {\sc flockture} are essentially equivalent programs.
-Second, we simulated and analyzed datasets with {\sc flockture} and three models in {\sc structure} 
-(the admixture and correlated allele frequency (A-C) model, 
-the no-admixture and correlated allele frequency (NA-C) model, 
-and the no-admixture and non-correlated allele frequency (NA-NC) model) to help better understand
+{\sc flock}, initially designed to map admixture in hybrid populations 
+\citep{Duc&Tur2009}, was recently proposed as an alternative to 
+{\sc structure} to estimate the number of populations from a collection with 
+unsampled source populations \citep{Duc&Tur20123}. In doing so, 
+{\sc flock} relies on individuals with like genotypes clustering together 
+through an iterative process. It stands to reason then, a reasonable way 
+to gauge how well the algorithm accomplishes the main objective 
+clustering of individuals into correct groups is to simulate data of known 
+population structure with varying levels of differentiation and 
+evaluate how well it does assuming the correct number of populations. 
+The implementation of {\sc flock} as a VBA procedure (macro) in 
+Microsoft Excel precludes a rigorous evaluation of its behavior due 
+to the inability to efficiently batch process. Consequenty, we adapted 
+the source code from {\sc structure} as describe above to create {\sc flockture}
+(code provided on GitHub repository https://github.com/eriqande/flockture) 
+in order to accurately assess the strengths and weaknesses of the {\sc flock} 
+algorithm. First, using simulated data we demonstrated that {\sc flock} 
+and {\sc flockture} are essentially equivalent programs.
+Second, we simulated and analyzed datasets with {\sc flockture} 
+and three models in {\sc structure} (the admixture and correlated 
+allele frequency (A-C) model, the no-admixture and correlated 
+allele frequency (NA-C) model, and the no-admixture and non-correlated 
+allele frequency (NA-NC) model) to help better understand
 under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .
@@
 returned by {\sc structure}, which, in the no-admixture case, are posterior 
-probabilities of cluster membership.\citet{Duc&Tur2012} describe a method 
+probabilities of cluster membership. \citet{Duc&Tur2012} describe a method 
 for using {\sc flock} to estimate, $K$, the number of
@@
 Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
-we used  {\sc flockture} to analyze a many more datasets that encompassed a broader 
-range of population differentiation. Fourteen datasets were simulated as above (five populations 
-comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes)
-with average pairwise $F_{ST}$ values ranging between 0.11 and 0.005. 
+we used  {\sc flockture} to analyze 48 datasets for two marker types 
+(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
+range of population differentiation. The 96 datasets were simulated as above (five populations 
+comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the micro satellites
+and 96 loci for the SNPs)
+with average pairwise $F_{ST}$ values ranging between 0.005 and 0.040. 
  {\sc flockture} was run with the default run parameters and each run of 
@@
 
-Run times for both programs were relatively short, albeit in the case of {\sc flock} 
-non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
-over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
-general increase as population structure became more subtle, however {\sc flock} did not 
-demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
-% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
-% at later iterations leading to increased process times. Seems reasonable.
-
 \begin{figure*}
@@
 \end{figure*}
+
+Run times for both programs were relatively short, albeit in the case of {\sc flock} 
+non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
+over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
+general increase as population structure became more subtle, however {\sc flock} did not 
+demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
+% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
+% at later iterations leading to increased process times. Seems reasonable.
 
@@
 \centering
-  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FvSloss1.pdf}
+\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
   \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
-  frequency (A-C) model, 
-the no-admixture and correlated allele frequency (NA-C) model, 
-and the no-admixture and non-correlated allele frequency (NA-NC) model in {\sc structure} and {\sc flockture}. 
-Average pairwise $F_{ST}$ 
-values between populations is listed above each graph.}
-  \label{fig:FvSloss}
-\end{figure*} % These graphs need to be redone so that text is legible and corrected for the NA-C NOT the A-NC 
-%model in structure. I probably had a transcription error in R. Check code to be sure.
+  frequency (A.C) model, 
+the no-admixture and correlated allele frequency (NA.C) model, 
+and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
+  \label{fig:uSatLoss}
+\end{figure*}
+
+ \begin{figure*}
+\centering
+  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
+  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
+  frequency (A.C) model, 
+the no-admixture and correlated allele frequency (NA.C) model, 
+and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
+  \label{fig:SNPloss}
+\end{figure*}
 

### 2015-02-06 FLOCKTRUEvSTRUCTUREnew.R

@@ -3,7 +3,6 @@
 
 #set wd
 system('find /Users -name "flock-comment" -print 2>/dev/null')
-setwd("/Users/patdbarry/flock-comment")
 wd<-getwd()
 
 # We are going to need a few folders to do all this
@@ -22,17 +21,20 @@ StructureAreaDIR<-readLines('StructureAreaDIR.txt')
 
 
 ##### Set up the parameters to run simulations and run flockture and structure ####
-MGrate<-c(10)#c(5,7,10,15,20,24,28,30,32,36,40,60,90,150)#The values of pop differentiation is parameterized by migration rate M = 4 Ne m?
+MGrate<-c(20,24,28,30,34,36,40,60,90,150)#c(5,7,10,15,20,24,28,30,32,36,40,60,90,150)#The values of pop differentiation is parameterized by migration rate M = 4 Ne m?
 DatNum<-length(MGrate)#how many datasets do we have?
 Reps<-9 #how many reps do we want to run for each program
 Nloci<-15
 Npops<- 5
 N<-2000
+marker=1 #1 for usats 2 for SNPs for simulation purposes
 
 #### Simulate some datasets and store the input files in Dat* folder #####
 #     sim_data.sh simulates datasets
 sapply((1:DatNum),function(x) system(paste('mkdir ',paste('SimDat',x,sep=""),sep="")))
 
+if(marker==1){
+#### FOR uSATs####
 #might need a loop to do the following
 # move MGrate into sim_data_MIG_RATE.sh
 # loops over our MGrate vector to create simulated datasets and stores them in 
@@ -60,7 +62,32 @@ for(x in 1:DatNum){
   path<-lapply(1:length(files2mov),function (y) paste('mv ',files2mov[y],' ',wd,'/SimDat',x,sep=""))
   lapply(1:length(files2mov),function(y) system(path[[y]][1]))
 }
+}#do for uSats
 
+if(marker==2){
+  for(x in 1:DatNum){ 
+    setwd(paste(flockcommentDIR,"/simdata",sep=""))
+    path<-paste("sed 's#^M=.*#M=",MGrate[x],"#' sim_data_snps_MIG_RATE.sh > sim_data_snps_MIG_RATE_DAT.sh",sep="")
[…]

### 2015-02-08 FLOCKTRUEvSTRUCTUREnew.R
Added all the stuff to reproduce results. just need a readme

@@ -21,28 +21,45 @@ StructureAreaDIR<-readLines('StructureAreaDIR.txt')
 
 
 ##### Set up the parameters to run simulations and run flockture and structure ####
-MGrate<-c(20,24,28,30,34,36,40,60,90,150)#c(5,7,10,15,20,24,28,30,32,36,40,60,90,150)#The values of pop differentiation is parameterized by migration rate M = 4 Ne m?
-DatNum<-length(MGrate)#how many datasets do we have?
+Seedset<-3
 Reps<-9 #how many reps do we want to run for each program
-Nloci<-15
 Npops<- 5
 N<-2000
 marker=1 #1 for usats 2 for SNPs for simulation purposes
+if(marker==1){
+  MGrate<-c(20,24,28,30,31,32,36,37,37.5,38,40,50,60,75,90,125)
+}
+if(marker==2){
+  MGrate<-c(20,24,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
+}
+DatNum<-length(MGrate)#how many datasets do we have?
+qi_indLoss<-F #set true if you want graphs of qi values and loss by simulation
 
+##########################################################################
 #### Simulate some datasets and store the input files in Dat* folder #####
-#     sim_data.sh simulates datasets
-sapply((1:DatNum),function(x) system(paste('mkdir ',paste('SimDat',x,sep=""),sep="")))
+
+setwd(flockcommentDIR)
+sapply((1:(DatNum*Seedset)),function(x) system(paste('mkdir ',paste('SimDat',x,sep=""),sep="")))
 
 if(marker==1){
 #### FOR uSATs####
-#might need a loop to do the following
 # move MGrate into sim_data_MIG_RATE.sh
 # loops over our MGrate vector to create simulated datasets and stores them in 
 # folders named SimDatX where X is the number of the simulation
 # details on the simulation are stored in SimDiets to see the pairwise Fst Values etc.
-for(x in 1:DatNum){ 
-  setwd(paste(flockcommentDIR,"/simdata",sep=""))
[…]

### 2015-02-08 simdata/sim_data_snps_MIG_RATE.sh

@@ -8,9 +8,9 @@
 # sample sizes from the different populations will be
 SIZES="250 250 400 550 550"
 REPS=1
-NUMLOC=15
-M=150
-MUT=2.0  # scaled mutation rate
+NUMLOC=96
+M=25 
+S=50  # scaled mutation rate
 ms="./mac_progs/ms"
 ms2geno="./mac_progs/ms2geno"
 gsi_sim="./mac_progs/gsi_sim"
@@ -21,8 +21,8 @@ rm -f BaseFile_*.txt struct_intput*.txt
 
 echo "SEEDS1" > seedms   # set the seeds for ms
 echo "SEEDS2" > ms2geno_seeds # set seeds for ms2geno
-$ms $((2 * 2000)) $((REPS * NUMLOC)) -t $MUT -I 5 $(echo $SIZES | awk '{for(i=1;i<=NF;i++) printf("%d ",2*$i);}')  $M | \
-   $ms2geno -l $NUMLOC -b $SIZES -m 0 0 0 0 0 -u .10 2
+$ms $((2 * 2000)) $((REPS * NUMLOC)) -s $S -I 5 $(echo $SIZES | awk '{for(i=1;i<=NF;i++) printf("%d ",2*$i);}')  $M | \
+   $ms2geno -l $NUMLOC -b $SIZES -m 0 0 0 0 0 -s 1 10 4 -s 2 10 4 -s 3 10 4 -s 4 10 4 -s 4 10 4 
 
 
 echo "

### 2015-02-09 FLOCKTRUEvSTRUCTURE.R
New script that works with the sgl_pipe repo. Took down some outdated files

@@ -1,5 +1,13 @@
 #### FLOCKTURE vs STRUCTURE ####
+#load some useful libraries
+library(ggplot2)
+library(stringr)
+library(plyr)
+library(reshape2)
+library(miscFuncs)
 #Overall Goal - Run both programs a number of times ~9 on various levels of population differentiation
+
+marker=1 #### RUN WITH MICROSATELLITES (=1) OR SNPS (=2) 
 
 #set wd
 system('find /Users -name "flock-comment" -print 2>/dev/null')
@@ -15,22 +23,24 @@ flockcommentDIR<-readLines('flock-commentDIR.txt')
 #flockture - this will be where we simulate all of our datasets
 system('find /Users -name "flockture" -print 2>/dev/null -quit > flocktureDIR.txt') # this should always be the first result
 flocktureDIR<-readLines('flocktureDIR.txt')
-#StructureArea - this is where clump_and_distruct live
-system('find /Users -name "StructureArea" -print 2>/dev/null > StructureAreaDIR.txt')
-StructureAreaDIR<-readLines('StructureAreaDIR.txt')
+#slg_pipe - this is where clump_and_distruct live
+system('find /Users -name "slg_pipe" -print 2>/dev/null > slg_pipeDIR.txt')
+slg_pipeDIR<-readLines('slg_pipeDIR.txt')
+#Structure - this is where structure lives
+system('find /Users -name "structure" -print 2>/dev/null > StructureDIR.txt')
+StructureDIR<-readLines('StructureDIR.txt')
 
-
+########## PART 1 #################################################################
 ##### Set up the parameters to run simulations and run flockture and structure ####
 Seedset<-3
 Reps<-9 #how many reps do we want to run for each program
 Npops<- 5
 N<-2000
-marker=1 #1 for usats 2 for SNPs for simulation purposes
 if(marker==1){
-  MGrate<-c(20,24,28,30,31,32,36,37,37.5,38,40,50,60,75,90,125)
+  MGrate<-c(20,24,28,30)#,31,32,36,37,37.5,38,40,50,60,75,90,125)
 }
 if(marker==2){
-  MGrate<-c(20,24,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
+  MGrate<-c(20,24)#,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
 }
 DatNum<-length(MGrate)#how many datasets do we have?
[…]

### 2015-02-10 tex/main-body-text.tex

@@
 \citep{Duc&Tur2009}, was recently proposed as an alternative to 
-{\sc structure} to estimate the number of populations from a collection with 
-unsampled source populations \citep{Duc&Tur20123}. In doing so, 
+{\sc structure} for estimating the number of populations from a collection with 
+unsampled source populations \citep{Duc&Tur2012}. In doing so, 
 {\sc flock} relies on individuals with like genotypes clustering together 
@@
 to gauge how well the algorithm accomplishes the main objective 
-clustering of individuals into correct groups is to simulate data of known 
+of clustering individuals into their correct groups is to simulate data of known 
 population structure with varying levels of differentiation and 
@@
 https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
-were comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
+were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
 and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
@@
 range of population differentiation. The 96 datasets were simulated as above (five populations 
-comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the micro satellites
+composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
 and 96 loci for the SNPs)
-with average pairwise $F_{ST}$ values ranging between 0.005 and 0.040. 
+with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
  {\sc flockture} was run with the default run parameters and each run of 
@@
 {\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
-simulated datasets. The zero-one loss across all datasets had eqivalent results Fig.~\ref{fig:FvFloss}.
+simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
 When genetic differentiation is high both programs correctly apportion the majority of individuals and 
@@
 values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
-for {\sc flock}. In an attempt to understand the deviations being observed we took a subset of the 
+for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the deviations being observed we took a subset of the 
 first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
@@
 between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
-be identical, but should have displayed equivalent behavior. Plateaus were only observed for the dataset 
+be identical, but should have displayed equivalent behavior.
+
+Plateaus were only observed for the dataset 
 with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
@@
 As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
-well differentiated populations. Both programs have less than 1\% assignment error for most runs 
-at values of $F_{ST}  \geq$ 0.0603 (Fig.~\ref{fig:FvSloss}). Infrequently, the 
-NA-NC and NA-C models in {\sc structure} grouped two populations as one and 
-assign individuals from another to two separate populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
-disappears as population differentiation declines. As the subtlety of genetic structure 
-increased loss for both programs increases. 
-Performance of the two programs as $F_{ST}$ approaches 0.0238 is relatively similar
-characterized by the A-C model in {\sc structure} having the largest loss, the NA-C and NA-NC models performing
-relatively similar and {\sc flockture} fluctuating around the results of the other three models. The scale of these 
-differences is extremely small - on the order of 10 individuals being misassigned. As genetic 
-differentiation continues to decline {\sc flockture}  showed consistently better performance 
-at levels of $F_{ST}$ between 0.0238 to 0.0229.  Again, these differences are on the scale of roughly 20 individuals
-or 1\%. In simulation 12 we observed the variation in {\sc flockture} results increase drastically (Fig.~\ref{fig:FvSloss}),
-and for all other simulated datasets {\sc structure} had a lower assignment error.
+well differentiated populations. Infrequently, at higher levels of population differentiation the 
+NA-NC and NA-C models in {\sc structure} could be trapped in a local mode resulting in the 
+assignment of individuals from two groups into one while individuals from another 
+were separated into two populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
+(only present in the microsatellite datasets) disappears as population differentiation 
+declines. As the subtlety of genetic structure 
+increased loss for both programs increases and differences among markers emerged.
+For the microsatellite dataset we observed that both programs
+produce roughtly equivalent results at higher levels of population differentiation. At intermediate 
+values, the variance of the loss incurred by each model increases and certain models do marginally
+better in assignment. No model does consistently better than the others, until $F_{ST}$ drops 
+below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
+variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
+ {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
+values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
+higher proportion of individuals.
 

### 2015-02-10 FLOCKTRUEvSTRUCTURE.R
Found a formatting error in the last step, it would only display the simulations for the first set of seeds

@@ -37,10 +37,10 @@ Reps<-9 #how many reps do we want to run for each program
 Npops<- 5
 N<-2000
 if(marker==1){
-  MGrate<-c(20,24,28,30)#,31,32,36,37,37.5,38,40,50,60,75,90,125)
+  MGrate<-c(20,24,28,30,31,32,36,37,37.5,38,40,50,60,75,90,125)
 }
 if(marker==2){
-  MGrate<-c(20,24)#,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
+  MGrate<-c(20,24,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
 }
 DatNum<-length(MGrate)#how many datasets do we have?
 qi_indLoss<-F #set true if you want graphs of qi values and loss by simulation
@@ -538,8 +538,8 @@ ggsave(plot=p1,filename=paste(flockcommentDIR,'/SimDat',ds,'/QiCompRep',j,'SimDa
 
 ################################################################
 ################# Make final figure ############################
-SumMat<-matrix(data=NA,ncol=3,nrow=Reps*4*DatNum)
-for (i in 1:DatNum){
+SumMat<-matrix(data=NA,ncol=3,nrow=Reps*4*(DatNum*Seedset))
+for (i in 1:(DatNum*Seedset)){
 Fst<-round(unlist(rep(read.table(paste(wd,'/SimDat',i,'/AvgFst.txt',sep="")),Reps*4)),6)
 LossMat<-read.csv(paste(wd,'/SimDat',i,'/ZeroOneLoss_SimDat',i,'.csv',sep=""),sep=",",colClasses=c('character',rep('numeric',4)))
 LossMelt<-melt(LossMat)
@@ -568,7 +568,11 @@ P1<-ggplot(SumMat2,aes(x=Fst ,y=Loss,color=Model,fill=Model))  +
   theme(axis.text.x = element_text(angle = 25, hjust = 1))+
   labs(x=expression('F'['ST']))
 
-ggsave(plot=P1,filename=paste(wd,'/Loss',mark,'_Color.pdf',sep=''),dpi=300,width=6,height=6,units='in') 
-ggsave(plot=P1+scale_colour_grey(start=.8,end=0),filename=paste(wd,'/Loss',mark,'_BW.pdf',sep=''),dpi=300,width=6,height=6,units='in')
[…]

### 2015-02-10 FLOCKTRUEvSTRUCTURE.R
Minor initial changes.

@@ -1,55 +1,67 @@
-#### FLOCKTURE vs STRUCTURE ####
-#load some useful libraries
+#### FLOCKTURE vs STRUCTURE
+
+
+#### load some useful libraries ####
 library(ggplot2)
 library(stringr)
 library(plyr)
 library(reshape2)
 library(miscFuncs)
-#Overall Goal - Run both programs a number of times ~9 on various levels of population differentiation
 
-marker=1 #### RUN WITH MICROSATELLITES (=1) OR SNPS (=2) 
 
-#set wd
-system('find /Users -name "flock-comment" -print 2>/dev/null')
-wd<-getwd()
+#### Top-level settings and paths and stuff  ####
 
-# We are going to need a few folders to do all this
-# I wrote in some system commands to find relavent folders for people that might install 
-# directories in different places. 
+### Set paths
+# Your R working directory should be the flock-comment directory
+# and flockture and slg_pipe should be directories one level above.
+# If you are using RStudio and opened the flock-comment project you should
+# be good to go.
 
-#flock-comment - this will be our main working directory
-system('find /Users -name "flock-comment" -print 2>/dev/null >flock-commentDIR.txt')
-flockcommentDIR<-readLines('flock-commentDIR.txt')
-#flockture - this will be where we simulate all of our datasets
-system('find /Users -name "flockture" -print 2>/dev/null -quit > flocktureDIR.txt') # this should always be the first result
-flocktureDIR<-readLines('flocktureDIR.txt')
-#slg_pipe - this is where clump_and_distruct live
-system('find /Users -name "slg_pipe" -print 2>/dev/null > slg_pipeDIR.txt')
-slg_pipeDIR<-readLines('slg_pipeDIR.txt')
+# now, we want to get absolute paths to a lot of the directories we will be using:
+wd<-getwd()  # where you started...
+flockcommentDIR <- normalizePath("../flock-comment") # main working directory
+flocktureDIR <- normalizePath("../flockture") #  this will be where we simulate all of our datasets
+slg_pipeDIR <- normalizePath("../slg_pipe") # this is where clump and distruct live
+
+## Crazy stuff.  Will have to sort this out later.   Don't know what it should be.
 #Structure - this is where structure lives
-system('find /Users -name "structure" -print 2>/dev/null > StructureDIR.txt')
[…]

### 2015-02-10 FLOCKTRUEvSTRUCTURE.R
Cleaning up microsat data simulation
Tweaks and cleaning to:

1. set seeds -- DONE
2. rewrite script with different mig setting: replace with postional pars in script -- DONE
3. run simulation -- DONE

@@ -64,44 +64,51 @@ if(any(lapply(paste("SimDat", 1:(DatNum*Seedset), sep = ""), dir.create) == FALS
 }
 
 if(marker==1){
-#### FOR uSATs####
-# move MGrate into sim_data_MIG_RATE.sh
-# loops over our MGrate vector to create simulated datasets and stores them in 
-# folders named SimDatX where X is the number of the simulation
-# details on the simulation are stored in SimDiets to see the pairwise Fst Values etc.
-Nloci<-15
-setwd(paste(flockcommentDIR,"/simdata",sep=""))
-seedsms<-readLines('uSat_seedsms.txt')
-seedsms2geno<-readLines('uSat_seedsms2geno.txt')
-index<-seq(from=1,to=DatNum*Seedset,by=DatNum)
-for (s in 1:Seedset){
-  path<-paste("sed 's/SEEDS1/",seedsms[s],"/' sim_data_MIG_RATE.sh > sim_data_MIG_RATE_DAT1.sh",sep="")
-  system(path)
-  path<-paste("sed 's/SEEDS2/",seedsms2geno[s],"/' sim_data_MIG_RATE_DAT1.sh > sim_data_MIG_RATE_DAT2.sh",sep="")
-  system(path)
+  # ECA says: this does:
+  # 1. set seeds
+  # 2. rewrite script with different mig setting: replace with postional pars in script
+  # 3. run simulation
+  # 4. compute Average pairwise Fst
+  # 5. move files to SimDataX folder
+  #### FOR uSATs####
+  # move MGrate into sim_data_MIG_RATE.sh
+  # loops over our MGrate vector to create simulated datasets and stores them in 
+  # folders named SimDatX where X is the number of the simulation
+  # details on the simulation are stored in SimDets to see the pairwise Fst Values etc.
+  Nloci<-15
   
-  for(x in 1:DatNum){ 
-  path<-paste("sed 's#^M=.*#M=",MGrate[x],"#' sim_data_MIG_RATE_DAT2.sh > sim_data_MIG_RATE_DAT.sh",sep="")
-  system(path)
-  system('chmod 755 sim_data_MIG_RATE_DAT.sh')
-  # run the shell command to generate baseline and structure file
-  system('./sim_data_MIG_RATE_DAT.sh > SimDets.txt')
+  setwd(file.path(flockcommentDIR, "simdata"))
   
-  #what is the avg. pariwise FST?
[…]

### 2015-02-10 tex/main-body-text.tex
Added new results from SNP datasets. Some minor additions to the Rscript to make a table of plateau records from FLOCKTURE

@@
 For the microsatellite dataset we observed that both programs
-produce roughtly equivalent results at higher levels of population differentiation. At intermediate 
+produce roughly equivalent results at higher levels of population differentiation. At intermediate 
 values, the variance of the loss incurred by each model increases and certain models do marginally
-better in assignment. No model does consistently better than the others, until $F_{ST}$ drops 
+better in assignment. No model did consistently better than others, until $F_{ST}$ drops 
 below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
@@
 
-The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau quickly 
-diminished as the amount of differentiation among populations decreased. For simulations 1-4 a single plateau
-greater than 6 was observed in every run. For simulation 5 only a single run yielded a plateau record 
-with a single plateau greater than six. For simulations 6 to 11 plateau records of various lengths were observed
-but all plateaus were less than 6. No plateaus were observed for datasets 12 through 20. Despite the lack of 
-plateaus observed for these datasets graphs of the $q_i$ values were suggestive of population structure within the 
-data (Fig.~\ref{fig:FvSdistruct}  Simulation 13). {\sc district} plots of the normalized likelihood values from
-{\sc flockture} were consistent and suggestive of population structure within the dataset until simulation 17
-after which the majority of the 'best runs' were substantially different in how they clustered individuals.
+The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau 
+diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
+only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. No plateau record had
+a length of six or greater at any $F_{ST}$ value greater than 0.0337. Results from the SNP datasets 
+were concordant. No runs of {\sc flockture} resulted in a single plateau length of greater than six and
+only $F_{ST}$ values greater 0.0323 had plateau records with a plateau greater than six. Despite the lack of 
+plateaus observed for many of the simulated datasets graphs of the $q_i$ values were suggestive 
+of population structure within the 
+data (Fig.~\ref{fig:FvSdistruct}  Simulation 13).
 

### 2015-02-10 FLOCKTRUEvSTRUCTURE.R

@@ -573,6 +573,29 @@ ggsave(plot=P1+scale_colour_grey(start=.8,end=0),filename=paste(wd,'/Loss',mark,
 #########################################################
 #########################################################
 
+#Plateau Table
+setwd(wd)
+Plata<-matrix(data=NA,ncol=Reps+1,nrow=(DatNum*Seedset))
 #Lets Take a look at the Plateaus
 
-Plata<-
+for (d in 1:(DatNum*Seedset)){
+  for (r in 1:Reps){
+  Fst<-scan(file=paste(wd,'/SimDat',d,'/AvgFst.txt',sep=""))
+  Plata[d,1]<-Fst
+  plateau<-scan(file=paste(wd,'/SimDat',d,'/FlockturePlateaus.csv',sep=""),what="character",nlines=1,skip=(r-1))
+  plateau<-paste(unlist(str_split(plateau,","))[which(unlist(str_split(plateau,","))>1)],collapse=",") 
+  Plata[d,r+1]<-plateau
+  }
+}
+colnames(Plata)<-c('Fst',paste('Rep',seq(1,9,1),sep=''))
+Plata<-Plata[order(Plata[,1]),]
+sink(file=paste('LatexPlateauTable_',mark,'.txt',sep=''),append=F)
+print(xtable(Plata), include.rownames=FALSE)
+sink()
+
+system(paste("sed '/INSERTFILEHERE/r LatexPlateauTable_",mark,".txt' <Table2.tex >PlateuTable_",mark,"INT.tex",sep=""))
+system(paste("sed 's/INSERTFILEHERE//' <PlateuTable_",mark,"INT.tex >PlateuTable_",mark,".tex",sep=""))
+system("rm *INT.tex")            
+
+##### clean up all the misc files sitting around
+system("rm *DIR.txt")

### 2015-02-10 FLOCKTRUEvSTRUCTURE.R
added supplementary tables for the plateau records and updated Rscript to create them

@@ -584,6 +584,7 @@ for (d in 1:(DatNum*Seedset)){
   Plata[d,1]<-Fst
   plateau<-scan(file=paste(wd,'/SimDat',d,'/FlockturePlateaus.csv',sep=""),what="character",nlines=1,skip=(r-1))
   plateau<-paste(unlist(str_split(plateau,","))[which(unlist(str_split(plateau,","))>1)],collapse=",") 
+  if(is.na(as.numeric(unlist(str_split(plateau,","))[1]))){plateau<-"1"}
   Plata[d,r+1]<-plateau
   }
 }

### 2015-02-11 FLOCKTRUEvSTRUCTURE.R
Working through the simulation script.
Got part way through structure and then hung.

I think it was an R studio barf.

@@ -7,26 +7,24 @@ library(stringr)
 library(plyr)
 library(reshape2)
 library(miscFuncs)
+library(dplyr)  # for the %>% operator
 
 
-#### Top-level settings and paths and stuff  ####
+##### Top-level settings and paths and stuff  ####
 
 ### Set paths
 # Your R working directory should be the flock-comment directory
-# and flockture and slg_pipe should be directories one level above.
+# and flockture and slg_pipe should be directories that sit beside flock-comment
+# one directory level above.
 # If you are using RStudio and opened the flock-comment project you should
 # be good to go.
 
 # now, we want to get absolute paths to a lot of the directories we will be using:
-wd<-getwd()  # where you started...
+wd<-getwd()  # where you started...in case we need to use it later...
 flockcommentDIR <- normalizePath("../flock-comment") # main working directory
 flocktureDIR <- normalizePath("../flockture") #  this will be where we simulate all of our datasets
 slg_pipeDIR <- normalizePath("../slg_pipe") # this is where clump and distruct live
-
-## Crazy stuff.  Will have to sort this out later.   Don't know what it should be.
-#Structure - this is where structure lives
-#system('find /Users -name "structure" -print 2>/dev/null > StructureDIR.txt')
-#StructureDIR<-readLines('StructureDIR.txt')
+StructBinaryPath <- normalizePath("../slg_pipe/inputs/ForStructure/structure2.3.4")
 
 
 ### Set genetic marker type to simulate
@@ -35,15 +33,15 @@ marker=1 # RUN WITH MICROSATELLITES (=1) OR SNPS (=2)
 ### Other options to set
 qi_indLoss <- F # set true if you want graphs of qi values and loss by simulation
 
-#### PART 1: Simulate data sets ####
+##### PART 1: Simulate data sets ####
 
-### Set up the parameters controlling the simulations ####
+### Set up the parameters controlling the simulations 
 Seedset <- 3
 Reps <- 9 # how many reps do we want to run for each program
 Npops <- 5
 N <- 2000
 
-# set up migration rates to use 
+## set up migration rates to use 
 if(marker == 1){
[…]

### 2015-02-11 tex/main-body-text.tex
edited conclusion

@@
 for using {\sc flock} to estimate, $K$, the number of
-subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
+subpopulations. The user must specify both the number of runs 
+(different initial allocations of individuals 
 to each cluster) as well as the number of iterations. (In each iteration every individual 
@@
 membership to all $K$ clusters is computed
-and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
-and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
-have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
-{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
-The plateau record is a list of how many runs converged to the same allocation
-of individuals. \citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
-\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
-$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
-the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
-if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
-this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
+and the log likelihood difference (LLOD), the difference between the 
+cluster with the highest likelihood 
+and second highest likelihood, is calculated for each individual. {\sc flock} 
+identifies runs that have converged to the same solution by detecting 
+that they have the same mean LLOD score, and, in 
+{\sc flock} parlance, a series of runs that have converged to the same 
+solution are termed a ``plateau.'' The plateau record is a list of 
+how many runs converged to the same allocation
+of individuals. \citet{Duc&Tur2012} developed two stopping 
+rules that they use to estimate $K$. \citet{Duc&Tur2012} advocate 
+running  {\sc flock} until either a single plateau of
+$\geq 6$ runs is reached or four consecutive $K$ values yield no 
+plateaus. If the first stopping condition is satisfied, the $K$ for which 
+there is a single plateau of $\geq 6$ runs is the point estimate for $K$. 
+Alternatively, if multiple plateaus are observed with at least one plateau 
+of $\geq 6$ runs, the largest value of $K$ that satisfies 
+this conditon should be taken as the lower bound estimate of $K$. 
+If no plateaus of $\geq 6$ runs are observed, $K$
 is declared {\em undecided}.
@@
 run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
-LLOD threshold = 0). We judged {\sc flockture} as an accurate representation of {\sc flock} based on the
-following criteria: (1) Clustering of individuals into corrected population as defined by a zero-one loss function, 
-(2) individual $q_i$ values, and (3) plateau records. All runs were performed on a JNCS D685 with an i7 3.40GHz 
-processor and 8GB of RAM running 64-bit Windows 7.
+LLOD threshold = 0). We judged {\sc flockture} as an accurate 
+representation of {\sc flock} based on the
+following criteria: (1) Clustering of individuals into corrected population as
+ defined by a zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
+All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
+RAM running 64-bit Windows 7.
 
-Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
-we used  {\sc flockture} to analyze 48 datasets for two marker types 
+Once we were confident that no major differences in inference existed between 
+{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for two marker types 
 (microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
@@
 values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
-for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the deviations being observed we took a subset of the 
+for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the 
+deviations being observed we took a subset of the 
 first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
@@
   \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
-  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of dataset 1 
-with average pairwise $F_{ST} = $ 0.03. }
+  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of simulated
+data with average pairwise $F_{ST} = $ 0.03. }
   \label{fig:FvFqi}
@@
 well differentiated populations. Infrequently, at higher levels of population differentiation the 
-NA-NC and NA-C models in {\sc structure} could be trapped in a local mode resulting in the 
-assignment of individuals from two groups into one while individuals from another 
-were separated into two populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
-(only present in the microsatellite datasets) disappears as population differentiation 
+NA-NC and NA-C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
+resulting from the assignment of individuals from two groups into a single cluster while 
+individuals from another group were separated into two clusters
+with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
+(only present in a single microsatellite dataset) disappears as population differentiation 
 declines. As the subtlety of genetic structure 
-increased loss for both programs increases and differences among markers emerged.
+increased loss for both programs increases and differences in the performance of
+each model with different marker types emerged.
 For the microsatellite dataset we observed that both programs
@@
 
-The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau 
+The number and length of plateaus {\sc flockture} encountered quickly 
 diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
-only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. No plateau record had
-a length of six or greater at any $F_{ST}$ value greater than 0.0337. Results from the SNP datasets 
-were concordant. No runs of {\sc flockture} resulted in a single plateau length of greater than six and
-only $F_{ST}$ values greater 0.0323 had plateau records with a plateau greater than six. Despite the lack of 
-plateaus observed for many of the simulated datasets graphs of the $q_i$ values were suggestive 
-of population structure within the 
-data (Fig.~\ref{fig:FvSdistruct}  Simulation 13).
+only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
+replicate runs yielded plateau records varying in length from two to eight all of which had at 
+least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
+contained at least one plateau of length six or greater. No plateaus were observed for any run when 
+$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
+were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
+Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
+plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
+graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
+likelihood were suggestive of population structure.
+
+Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
+structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
+(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was indistinguisable
+and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
+and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
+similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
+As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
+values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.
 
@@
 \includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
-  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
-  frequency (A.C) model, 
-the no-admixture and correlated allele frequency (NA.C) model, 
-and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
+  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
+  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
+and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
+datasets were simulated with varying levels of population differentation. A coalescent based program
+was used to generate 15 loci microsatellite genotypes for 2000 individuals comprising five populations.}
   \label{fig:uSatLoss}
@@
   \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
-  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
-  frequency (A.C) model, 
-the no-admixture and correlated allele frequency (NA.C) model, 
-and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
+  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
+  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
+and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
+datasets were simulated with varying levels of population differentation. A coalescent based program
+was used to generate 96 loci single nucleotide polymorphism (SNP) genotypes for 2000 
+individuals comprising five populations.}
   \label{fig:SNPloss}
@@
 
-% haven't touched these yet...
-
 \section*{Conclusions}
@@
 a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
-We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
-with non-correlated allele frequencies within the program {\sc structure}. In comparing results from {\sc flock} and {\sc structure} 
+Estimating $K$ is a difficult task, and because it is often based on individual assignment we 
+evaluated these programs on their ability to correctly assign indivdiuals at the true value
+of $K$. We have shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
+with non-correlated allele frequencies within the program {\sc structure}. 
+We adapted the source code of {\sc structure} to create {\sc flockture} and show that
+ {\sc flock} and {\sc flockture} are essentially equivalent programs. 
+In comparing results from {\sc flockture} and {\sc structure} 
 on simulated datasets with varying levels of differnetiation among populations
@@
 Though we observed similar results
-between the two programs at low values of $K$, {\sc flock}'s `best-run' solutions became somewhat inconsistent at higher values of $K$.
-This observation was consistent with \citet{Duc&Tur2012}'s comparison of the 
-two programs using simulated data 
-for $K=4$ \& 8, $L=10$ (Fig. 4 in \citet{Duc&Tur2012}). 
-This is likely a consequence of the rapidly increasing size of the search space over possible 
-partitions as $K$ and $n$, the number of individuals, increase. 
+between the two programs at large values of $F_{ST}$, 
+{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
+and became somewhat inconsistent at lower values of $F_{ST}$.
+This is likely a consequence of the low population divergence and also
+the large size of the search space over possible partitions. 
 The likelihood surface defined on partitions of this space will often
 be multimodal---possessing numerous local peaks and troughs.
-The peak that  {\sc flock}'s algorithm finds is entirely determined by the initial, random, allocation of 
+The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
 individuals, as the algorithm proceeds deterministically after the initial random allocation.
-If the number of initial allocations does not increase
-as $K$ is increased, the starting allocations become diffusely scattered in the space of all
-possible partitions making it less likely that many (if any at all) will converge to the same local mode. 
-This coupled with the fact that there is 
-ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
-of results from {\sc flock} if no non-zero plateaus are encountered. The problem of getting 
+By having a large number of individuals
+that can be arranged into 5 groups defines a large partion space.
+Decreasing $F_{ST}$ leads to more homogenous 
+populations whose multilocus genotypes have a smaller difference in 
+likelihood of assigning to any given cluster. This could act by increasing the number of
+local peaks, leading {\sc flockture} settle on many different solutions. The problem of getting 
 caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
-different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.
-By increasing the number of loci used in their simulations, \citet{Duc&Tur2012} showed  that 
-{\sc flock} was able to converge in a simulated data set with $K=8$, but held no advantage in inference of $K$ 
-over {\sc structure}. That simulated data set, with $n=240$ was considerably smaller and less complex than the real data investigated here.
-%They simulated data using 500 individuals per population but only used 30 of them in the analysis.
-% if you have more loci than individuals per population with high Fst between groups it does great!
+different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.
 
@@
 is reported to be the short processing time for each run of {\sc flock}.
-While \citet{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
-comparisons, we did not observe this advantage in our analysis. This appears to result from two issues. First,
-our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  
-Second, in the comparison by \citet{Duc&Tur2012}, 
-{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
-one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of 
-the admixture and correlated allele frequency model in {\sc structure} using 
-a 50,000 sweep burn-in followed 
-by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
-value of $K$. 
-To us it seemed reasonable to compare, for each $K$, the running times of one iteration of {\sc structure} (50,000 
-burn in and 200,000 sample collection sweeps)
-to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
-as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
-clusters. Though
-it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
-constitutes a reasonable comparison with respect to run length and number of runs, our results indicate 
-that {\sc flock} did not carry a distinct run-time advantage over {\sc structure}.
+While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
+their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
+$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
+were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
+a 50,000 sweep burn-in followed by a 200,000 sweep sample 
+from the posterior for values of $K$ from 1 to three more than the true 
+value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
+constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
+{\sc flock} were to be implemented in a coding language rather than a VBA routine in microsoft excel
+it would be approximately 200 times faster.
 
@@
 a particular $K$,  similar---but not identical---partitions
-will provide no support to a given $K$. Given {\sc flock}'s tendency to converge to 
-different solutions with large $K$ and $n$, it seems that 
-this method for estimating $K$ is unlikely to be reliable or workable in large, complex
-problems.  One possible patch for this problem would be running
+will provide no support to a given $K$. We have shown that even at intermediate
+values of $F_{ST}$ {\sc flockture} has a tendency to converge to 
+different solutions where no plateaus are encountered. Despite the lack of
+plateaus plots of the normalized likelihood values demonstrate that
+{\sc flockture} still produces valuable information. One possible patch for this problem would be running
 multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
@@
 are very similar but are perhaps not identical; however such an approach would be labor-intensive
-and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} in such cases, keeping in mind, of course, that
+and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} 
+in such cases, keeping in mind, of course, that
 estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
 used. This is especially true in real populations which do not conform to the assumptions of genetic clustering models.
-
-We made our comparison between {\sc flock} and the {\em no-admixture} model in 
-{\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 
-its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
-of individuals in the sample and thus may be a more appropriate model than the no-admixture model,
-or {\sc flock}, when the sample contains individuals that are admixed between populations.  A further advantage
-of the with-admixture model is that it allows inference of the population of origin of individual alleles at 
-different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 
-corresponds to {\sc structure}'s model with uncorrelated allele frequency priors. \citet{Falushetal2003} found that
-using their ``$F$-model,'' which assumes that allele frequencies between populations are correlated {\em a priori}, 
-allowed {\sc structure} to dissect more subtle population structure than was possible using the prior with
-uncorrelated allele frequencies.  Modifying {\sc flock} so as to reap similar benefits from using a prior
-with correlated allele frequencies seems like it would be an interesting challenge, though difficult
-because {\sc flock}'s formulation effectively integrates the allele frequencies out of the model.
 
@@
 We are grateful to Kristen Ruegg and Robin Waples for helpful comments on a draft of this paper.
+P. Barry was supported by a fellowship from the Rasmuson Foundation.

### 2015-02-12 FLOCKTRUEvSTRUCTURE.R
Make progress messages. Redirect stucture output.
Closes #4

@@ -180,6 +180,7 @@ for (Ds in 1:(DatNum*Seedset)){
   Flockture.Runtime<-vector()
   FlocktureReps<-Reps #how many times are we running flockture
   for (P in 1:FlocktureReps){
+    message("Doing Flockture run on SimDat ", Ds, " out of ", DatNum * Seedset, "    Rep ", P, " of ", FlocktureReps)
     # run flockture on it and grab the results out
     ptm<-proc.time()
     catch <- run_flockture_bin(D, K = 5, iter = 20, reps = 50)
@@ -261,7 +262,7 @@ if(marker==2){
 main_param<-paste(flockcommentDIR,'/mainparams_',sep="") #note this has an _ because it will change depending on the model that is run
 ex_param<-paste(flockcommentDIR,'/extraparams',sep="")#for these models we don't have any extraparams so we can use the same blank file
 
-# need to run 3 model AdmixCorr, NonAdmixCorr, NonAdmixNonCorr - might want to make this user specified
+# need to run 3 models:  AdmixCorr, NonAdmixCorr, NonAdmixNonCorr - might want to make this user specified
 # seach SimDat file Rep times
 for (SDt in 1:(DatNum*Seedset)){ # goes into different SimDat folders
   # lets make some vectors to store some processing times
@@ -269,17 +270,22 @@ for (SDt in 1:(DatNum*Seedset)){ # goes into different SimDat folders
   NoAdmixCorrRuntime<-vector()
   NoAdmixNonCorrRuntime<-vector()
   for (R in 1:Reps){
+    message(paste("Running Structure:  SimDat", SDt, "of", DatNum*Seedset, "   Rep Number", R, "of", Reps))
+    
     #AdmixCorr
+    message("     AdmixCorr")
     ptm<-proc.time()
[…]

### 2015-03-06 DistructPlot_tex/Fig_AllSim.tex
Added new figures, edited each section, made all figures in color and greyscale

\documentclass[varwidth=true, border={80, 10}]{standalone}
\usepackage{graphicx} 
\usepackage{amssymb} 
\usepackage{epstopdf} 
\usepackage{amsfonts} 
\usepackage{natbib} 
\usepackage{subfigure} 
\usepackage{pdfsync} 
\usepackage{xspace}

\textwidth = 6.5 in 
\textheight = 8 in 
\oddsidemargin = -0.90 in 
\evensidemargin = 0.0 in 
\topmargin = 0.0 in 
\headheight = 0.0 in 
\headsep = 0.0 in 
\parskip = 0.2in 
\parindent = 0.0in

\newlength{\nameraise} 
\setlength{\nameraise}{.07in}

\newlength{\nameoverhang} 
\setlength{\nameoverhang}{0.0in}

\begin{document} 
 A. $F_{ST}$ = \input{../SimDat17/AvgFst.txt}\\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r001.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE A.C (9/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r011.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C (8/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r010.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C  (1/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r020.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (7/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r022.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (1/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r027.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (1/9)} \\
\includegraphics{../SimDat17/final_pdf/BB_ds_Clumped_NoLabel_k005r028.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (9/9)} \\
\noindent\rule[.5cm]{8.2in}{0.4pt}\\
\vspace{-2.5em}\\
B. $F_{ST}$ = \input{../SimDat38/AvgFst.txt}\\
\includegraphics{../SimDat38/final_pdf/BB_ds_Clumped_NoLabel_k005r001.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE A.C (9/9)} \\
\includegraphics{../SimDat38/final_pdf/BB_ds_Clumped_NoLabel_k005r010.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C (9/9)} \\
\includegraphics{../SimDat38/final_pdf/BB_ds_Clumped_NoLabel_k005r019.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (9/9)} \\
\includegraphics{../SimDat38/final_pdf/BB_ds_Clumped_NoLabel_k005r028.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (9/9)} \\
\noindent\rule[.5cm]{8.2in}{0.4pt}\\
\vspace{-2.5em}\\
C. $F_{ST}$ = \input{../SimDat10/AvgFst.txt}\\
\includegraphics{../SimDat10/final_pdf/BB_ds_Clumped_NoLabel_k005r001.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE A.C (9/9)} \\
\includegraphics{../SimDat10/final_pdf/BB_ds_Clumped_NoLabel_k005r010.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C (9/9)} \\
\includegraphics{../SimDat10/final_pdf/BB_ds_Clumped_NoLabel_k005r019.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (9/9)} \\
\includegraphics{../SimDat10/final_pdf/BB_ds_Clumped_NoLabel_k005r028.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (9/9)} \\
\noindent\rule[.5cm]{8.2in}{0.4pt}\\
\vspace{-2.5em}\\
D. $F_{ST}$ = \input{../SimDat13/AvgFst.txt}\\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r001.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE A.C (9/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r010.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C (9/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r019.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (9/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r029.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (9/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r028.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (1/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r031.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (1/9)} \\
\includegraphics{../SimDat13/final_pdf/BB_ds_Clumped_NoLabel_k005r032.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (1/9)} \\
\noindent\rule[.5cm]{8.2in}{0.4pt}\\
\vspace{-2.5em}\\
E. $F_{ST}$ = \input{../SimDat31/AvgFst.txt}\\
\includegraphics{../SimDat31/final_pdf/BB_ds_Clumped_NoLabel_k005r001.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE A.C (9/9)} \\
\includegraphics{../SimDat31/final_pdf/BB_ds_Clumped_NoLabel_k005r010.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.C (9/9)} \\
\includegraphics{../SimDat31/final_pdf/BB_ds_Clumped_NoLabel_k005r019.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{STRUCTURE NA.NC  (9/9)} \\
\includegraphics{../SimDat31/final_pdf/BB_ds_Clumped_NoLabel_k005r028.pdf}~\hspace*{-\nameoverhang}\raisebox{\nameraise}{FLOCK (1/9)*} \\

\end{document}

### 2015-03-06 tex/abstract.tex
Added new figures, edited each section, made all figures in color and greyscale

@@
 annealing algorithm with an extreme cooling 
-schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
-confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
-model and uncorrelated allele frequency prior with more variability 
-from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
+schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  We 
+demonstrate the similarities between the two programs in a two step approach. First,
+we modified the source code of  {\sc structure} to resemble {\sc flock} producing
+the program {\sc flockture}. With simulated data we confirm similar behavior between 
+{\sc flock}  and {\sc flockture} by assing the programs' performance
+to assign individuals back to their correct population with a 
+0-1 loss function (ie. a loss of 1 is incurred if an individual is assinged 
+to an incorrect group). Second, we simulated many datasets under varying 
+levels of population differentiation for both microsatllite and SNP genoyptes. We show that
+{\sc flockture} yields results similar to {\sc structure} although with much larger 
+variability from run to run. {\sc flockture} did perform better than {\sc structure} 
+when genotypes were composed of SNPs and differentiation was moderate 
+($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} had a lower loss
+for both marker types. Interpreting {\sc flock}'s algorithm as a special case of the model in 
 {\sc structure} should aid in understanding the program's output and behavior.

### 2015-03-06 tex/main-body-text.tex
Added new figures, edited each section, made all figures in color and greyscale

@@
 give substantially different results than {\sc structure}. Below we describe
-the correspondence between the programs in more detail, and then, for illustration, compare
-the results obtained from {\sc flock} and {\sc structure} on a large, real data
-set.
+the correspondence between the programs in more detail, and then, for illustration,
+create the progam {\sc flockture} by adapting the code of {\sc structure} to 
+use a simulated annealing approach. We then compare results from large simulated 
+datasets with varying degrees of population differentiation.
 
@@
 population structure with varying levels of differentiation and 
-evaluate how well it does assuming the correct number of populations. 
+evaluate how well it does assuming the correct number of populations. The 
+algorithme performance can then be measured with a zero-one 
+loss function (where a loss of 1 is incurred if an individual is misassigned).
+
 The implementation of {\sc flock} as a VBA procedure (macro) in 
@@
 and three models in {\sc structure} (the admixture and correlated 
-allele frequency (A-C) model, the no-admixture and correlated 
-allele frequency (NA-C) model, and the no-admixture and non-correlated 
-allele frequency (NA-NC) model) to help better understand
+allele frequency (A.C) model, the no-admixture and correlated 
+allele frequency (NA.C) model, and the no-admixture and non-correlated 
+allele frequency (NA.NC) model) to help better understand
 under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .
@@
 following criteria: (1) Clustering of individuals into corrected population as
- defined by a zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
+ defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
 All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
@@
 general increase as population structure became more subtle, however {\sc flock} did not 
-demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
-% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
-% at later iterations leading to increased process times. Seems reasonable.
+demonstrate a linear increase.
 
@@
 well differentiated populations. Infrequently, at higher levels of population differentiation the 
-NA-NC and NA-C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
+NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
 resulting from the assignment of individuals from two groups into a single cluster while 
@@
 with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
-(only present in a single microsatellite dataset) disappears as population differentiation 
-declines. As the subtlety of genetic structure 
-increased loss for both programs increases and differences in the performance of
+(only present in a single microsatellite dataset) disappeared as population differentiation 
+declined. As the subtlety of genetic structure 
+increased loss for both programs increased and differences in the performance of
 each model with different marker types emerged.
@@
 produce roughly equivalent results at higher levels of population differentiation. At intermediate 
-values, the variance of the loss incurred by each model increases and certain models do marginally
-better in assignment. No model did consistently better than others, until $F_{ST}$ drops 
+values, the variance of the loss incurred by each model increased and certain models did marginally
+better in assignment. No model did consistently better than others, until $F_{ST}$ droped 
 below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
@@
 only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
-replicate runs yielded plateau records varying in length from two to eight all of which had at 
+replicate runs yielded plateau records varying in length from two to eight, all of which had at 
 least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
-contained at least one plateau of length six or greater. No plateaus were observed for any run when 
+contained a plateau of length six or greater. No plateaus were observed for any run when 
 $F_{ST}$ was larger than 0.02187. Results from the SNP datasets
@@
 structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
-(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was indistinguisable
+(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguisable
 and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
@@
 \centering
-\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
+\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatLoss.pdf}%uSatloss_BW.pdf is a greyscale version
   \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
@@
 and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
-datasets were simulated with varying levels of population differentation. A coalescent based program
-was used to generate 15 loci microsatellite genotypes for 2000 individuals comprising five populations.}
+datasets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
+5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
+of both programs. }
   \label{fig:uSatLoss}
@@
 \centering
-  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
+  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%SNPloss_BW.pdf is a greyscale version
   \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
@@
 and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
-datasets were simulated with varying levels of population differentation. A coalescent based program
-was used to generate 96 loci single nucleotide polymorphism (SNP) genotypes for 2000 
-individuals comprising five populations.}
+datasets, composed of 2,000 individuals with 96 SNP genotypes comprising 
+5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
+of both programs. }
   \label{fig:SNPloss}
@@
 \centering
-  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}
+  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
   \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
@@
  runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
-  frequency (A-C) model, 
-the no-admixture and correlated allele frequency (NA-C) model, 
-and the no-admixture and non-correlated allele frequency (NA-NC) model. Each simulation is comprised of
+  frequency (A.C) model, 
+the no-admixture and correlated allele frequency (NA.C) model, 
+and the no-admixture and non-correlated allele frequency (NA.NC) model. Each simulation is comprised of
  five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
@@
 likelihood of assigning to any given cluster. This could act by increasing the number of
-local peaks, leading {\sc flockture} settle on many different solutions. The problem of getting 
+local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
 caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
 different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.
+In species characterized by large subpopulations and low divergence, as is the case with many 
+marine fish, the use of {\sc flock} should probably be avoided.
 
@@
 constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
-{\sc flock} were to be implemented in a coding language rather than a VBA routine in microsoft excel
+{\sc flock} were to be implemented in a coding language rather than a VBA routine in Microsoft Excel
 it would be approximately 200 times faster.

### 2015-03-06 FLOCKTRUEvSTRUCTURE.R

@@ -51,18 +51,18 @@ qi_indLoss <- F # set true if you want graphs of qi values and loss by simulatio
 ##### PART 1: Simulate data sets ####
 
 ### Set up the parameters controlling the simulations 
-Seedset <-  1 # 3
+Seedset <-  3
 ### ECA: gotta set reps back after testing
-Reps <- 1  #9 # how many reps do we want to run for each program
+Reps <- 9 # how many reps do we want to run for each program
 Npops <- 5
 N <- 2000
 
 ## set up migration rates to use 
 if(marker == 1) {
   ## ECA got to set MGrate back after testing
-  MGrate <- c(20, 24)   #c(20,24,28,30)#,31,32,36,37,37.5,38,40,50,60,75,90,125)
+  MGrate <- c(20,24,28,30,31,32,36,37,37.5,38,40,50,60,75,90,125)
 } else if(marker == 2) {
-  MGrate<-c(20,24) #,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
+  MGrate<-c(20,24,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)
 } else {
   stop("Unrecognized value for variable marker")
 }
@@ -632,8 +632,8 @@ P1<-ggplot(SumMat2,aes(x=Fst ,y=Loss,color=Model,fill=Model))  +
   theme(axis.text.x = element_text(angle = 25, hjust = 1))+
   labs(x=expression('F'['ST']))
 
-ggsave(plot=P1,filename=paste(flockcommentDIR,'/Loss',mark,'_Color.pdf',sep=''),dpi=300,width=6,height=6,units='in') 
-ggsave(plot=P1+scale_colour_grey(start=.8,end=0),filename=paste(flockcommentDIR,'/Loss',mark,'_BW.pdf',sep=''),dpi=300,width=6,height=6,units='in')
+ggsave(plot=P1,filename=paste(flockcommentDIR,'/Loss',mark,'_Color.pdf',sep=''),dpi=300,width=11,height=8.5,units='in') 
+ggsave(plot=P1+scale_colour_grey(start=.8,end=0),filename=paste(flockcommentDIR,'/Loss',mark,'_BW.pdf',sep=''),dpi=300,width=11,height=8.5,units='in')
 
 #Plateau Table
 setwd(flockcommentDIR)

### 2015-03-11 tex/abstract.tex
Eric edited the manuscript.  Pretty close to ready to go now.

--- 2014-08-21 tex/abstract.tex
@@
       We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
-interpreted as an estimation procedure using  
-a model that is essentially identical to the {\sc structure} 
+interpreted as an estimation procedure based on 
+a model essentially identical to the {\sc structure} 
 \citep{Pritchardetal2000} model with no admixture and non-correlated 
-allele frequency priors. The computational routine in {\sc flock} is 
-equivalent to searching for the maximum-{\em a-posteriori}
-estimate of this restricted {\sc structure} model via a simulated 
-annealing algorithm with an extreme cooling 
-schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
-confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
-model and uncorrelated allele frequency prior with more variability 
-from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
+allele frequency priors. Rather than using MCMC, the {\sc flock} algorithm 
+searches for the maximum-{\em a-posteriori}
+estimate of this {\sc structure} model via a simulated 
+annealing algorithm with a rapid cooling 
+schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  We 
+demonstrate the similarities between the two programs in a two step approach. First,
+to enable rapid batch processing of many simulated data sets,
+we modified the source code of  {\sc structure} to use the {\sc flock} algorithm, producing
+the program {\sc flockture}. With simulated data we confirmed that results obtained with
+{\sc flock}  and {\sc flockture} and very similar (though flockture is some 200 times faster). 
+Second, we simulated multiple large datasets under varying 
+levels of population differentiation for both microsatellite and SNP genotypes. We analyzed them
+with {\sc flockture} and {\sc structure} and assessed each program on its ability to correctly cluster
+individuals to their correct subpopulation.  We show that
+{\sc flockture} yields results similar to {\sc structure} although with greater 
+variability from run to run. {\sc flockture} did perform better than {\sc structure} 
+when genotypes were composed of SNPs and differentiation was moderate 
+($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
+for both marker types. Interpreting {\sc flock}'s algorithm as a special case of the model in 
 {\sc structure} should aid in understanding the program's output and behavior.

### 2015-03-11 tex/main-body-text.tex
Eric edited the manuscript.  Pretty close to ready to go now.

@@
 
-However, here we identify 
-the close relationship between {\sc flock}'s algorithm and the program
+Given the above sort of emphasis on the distinctiveness of the {\sc flock}
+algorithm, it would not be surprising if most readers incorrectly believed {\sc flock}
+to be based upon a completely different model than the well-known program
+{\sc structure}. Here we remedy such confusion with a simple derivation
+that identifies 
+the close relationship between the {\sc flock} algorithm and the program
 {\sc structure}.  Although the algorithm in {\sc flock}
@@
 estimate from a marginalized form of the {\sc structure} model with no
-admixture and non-correlated allele frequencies.
+admixture and non-correlated allele frequencies.  Thus, while the models underlying
+{\sc structure} and {\sc flock} are nearly indistinguishable, {\sc structure}'s inference
+relies on MCMC samples from the posterior distribution, while
+{\sc flock}'s inference involves finding a local maximum in the posterior surface.
 
@@
 the correspondence between the programs in more detail, and then, for illustration,
+and to aid in application to simualted data, we
 create the progam {\sc flockture} by adapting the code of {\sc structure} to 
-use a simulated annealing approach. We then compare results from large simulated 
-datasets with varying degrees of population differentiation.
+use a simulated annealing approach. We then compare results of the two
+programs from large 
+datasets simulated with varying degrees of population differentiation.
 
@@
 
-The first step of the flock algorithm is initialization, during which 
+The first step of the {\sc flock} algorithm is initialization, during which 
 the program randomly allocates each of the $N$ individuals to one of 
@@
 software {\sc geneclass2}  \citep{Piryetal2004}.  If \citeauthor{Duc&Tur2009}
-use the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
+used the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
 be reallocated to whichever cluster gives the highest value to 
@@
 defined in (\ref{eq:cdm-multi-prob}).  
-If \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
-formulation of \citet{Paetkauetal1995}, then their reallocations are based
+However, it turns out that \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
+formulation of \citet{Paetkauetal1995}---though replacing $0.01$ with $1/(2n)$ where 
+$n$ is the number of individuals curently allocated to cluster (Pierre Duchesne, pers. comm.)---so 
+their reallocations are based
 on maximizing a likelihood that, although it may not be formally identical to
@@
 The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
-$Z$'s in proportion to their posterior probability.  If, all that was desired
+$Z$'s in proportion to their posterior probability.  If all that was desired
 was a point estimate of the $Z_i$'s that maximized the posterior probability,
@@
 ``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
-the algorithm.''  As such, we can expect that it may be more susceptible 
-to becoming trapped in local modes of the posterior probability. 
-In the following section we show that {\sc structure} and {\sc flock} find similar
-solutions in the analysis of a large microsatellite data set, though, as expected,
-{\sc flock} seems to find the same solution less consistently than does {\sc structure}.
+the algorithm.''  As such, we should not be surprised if it is prone 
+to becoming trapped in local modes of the posterior probability, though
+the practical implications of that are not yet clear.
 
@@
 {\sc structure} for estimating the number of populations from a collection with 
-unsampled source populations \citep{Duc&Tur2012}. In doing so, 
-{\sc flock} relies on individuals with like genotypes clustering together 
-through an iterative process. It stands to reason then, a reasonable way 
-to gauge how well the algorithm accomplishes the main objective 
-of clustering individuals into their correct groups is to simulate data of known 
+unsampled source populations \citep{Duc&Tur2012}. Whatever the ultimate goal
+of its application,  
+{\sc flock} relies on clustering individuals with like genotypes together  
+through an iterative process. Thus, a reasonable way 
+to gauge how well the algorithm functions is to simulate data of known 
 population structure with varying levels of differentiation and 
-evaluate how well it does assuming the correct number of populations. The 
-algorithme performance can then be measured with a zero-one 
-loss function (where a loss of 1 is incurred if an individual is misassigned).
+evaluate how well it does in recovering that structure when the data are
+analyzed assuming
+the number of subpopulations is exactly the number of subpopulations in the simulation. The 
+algorithm performance can then be measured with a zero-one 
+loss function (where a loss of 1 is incurred if an individual is group with the wrong cluster).
 
 The implementation of {\sc flock} as a VBA procedure (macro) in 
-Microsoft Excel precludes a rigorous evaluation of its behavior due 
-to the inability to efficiently batch process. Consequenty, we adapted 
-the source code from {\sc structure} as describe above to create {\sc flockture}
+Microsoft Excel retards rigorous evaluation of its behavior due 
+the its slow execution speed and 
+the inability to efficiently batch-process multiple data sets. Consequenty, we adapted 
+the source code from {\sc structure} as described above to create {\sc flockture}
 (code provided on GitHub repository https://github.com/eriqande/flockture) 
 in order to accurately assess the strengths and weaknesses of the {\sc flock} 
-algorithm. First, using simulated data we demonstrated that {\sc flock} 
+algorithm. First, using simulated data we demonstrate that {\sc flock} 
 and {\sc flockture} are essentially equivalent programs.
-Second, we simulated and analyzed datasets with {\sc flockture} 
+Second, we simulate and analyze datasets with {\sc flockture} 
 and three models in {\sc structure} (the admixture and correlated 
@@
 allele frequency (NA.NC) model) to help better understand
-under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .
+under what conditions, if any, running {\sc flock} may be preferable to {\sc structure}.
 
@@
 membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
-longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
+longest plateau---refered to as the `best run' by \citep{Duc&Tur2012} (see below).
 These values can be compared to the $q_i$ values
 returned by {\sc structure}, which, in the no-admixture case, are posterior 
-probabilities of cluster membership. \citet{Duc&Tur2012} describe a method 
-for using {\sc flock} to estimate, $K$, the number of
-subpopulations. The user must specify both the number of runs 
+probabilities of cluster membership. In {\sc flock} the user must specify both the number of runs 
 (different initial allocations of individuals 
-to each cluster) as well as the number of iterations. (In each iteration every individual 
+to each cluster) as well as the number of iterations (in each iteration every individual 
 has the opportunity to be reallocated 
-to a different cluster.) At the conclusion of each run, the log-likelihood value for 
+to a different cluster).  At the conclusion of each run, the log-likelihood value for 
 membership to all $K$ clusters is computed
@@
 how many runs converged to the same allocation
-of individuals. \citet{Duc&Tur2012} developed two stopping 
-rules that they use to estimate $K$. \citet{Duc&Tur2012} advocate 
-running  {\sc flock} until either a single plateau of
-$\geq 6$ runs is reached or four consecutive $K$ values yield no 
-plateaus. If the first stopping condition is satisfied, the $K$ for which 
-there is a single plateau of $\geq 6$ runs is the point estimate for $K$. 
-Alternatively, if multiple plateaus are observed with at least one plateau 
-of $\geq 6$ runs, the largest value of $K$ that satisfies 
-this conditon should be taken as the lower bound estimate of $K$. 
-If no plateaus of $\geq 6$ runs are observed, $K$
-is declared {\em undecided}.
+of individuals.
 
-To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
+To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
 were simulated using a coalescent based program (code provided on GitHub repository 
-https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
-were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
-and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
-populations. Both programs were run nine times at $K=5$ with the default setting for all 
+https://github.com/eriqande/flock-comment/simdata). The 5 populations
+were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
+Each of the four data sets was simulated with a different 
+average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
+Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
 run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
@@
 Once we were confident that no major differences in inference existed between 
-{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for two marker types 
+{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
 (microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
@@
  {\sc flockture} was run with the default run parameters and each run of 
-{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
+{\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
 from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
@@
 was used to relabel clusters among different runs,
-while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+while the program {\sc distruct} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
 
@@
 simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
-When genetic differentiation is high both programs correctly apportion the majority of individuals and 
-as genetic differentiation degrades among populations both programs show corresponding declines in 
-assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
-substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
-values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
-for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the 
-deviations being observed we took a subset of the 
+When genetic differentiation is high both programs correctly cluster the majority of individuals and 
+as genetic differentiation declines among populations both programs show a corresponding deterioration in 
+clustering ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
+substantially among the simulated datasets.
+
+We attempted for some time to develop {\sc flockture} to yield
+completely identical results to those from {\sc flock}; however we did not
+persist in those endeavors when we realized that {\sc flock} itself does
+not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the 
+we took a subset of the 
 first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
-{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
+{\sc flock} with all individuals originating from population 1 being placed in reference population 1 and all 
 individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
 with the same starting clusters, but with the reference population labels switched. While the final allocation
-of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
-between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
-be identical, but should have displayed equivalent behavior.
+of individuals matched, the mean LLOD was slightly different (5.045009 vs. 5.028084), indicating it would
+be futile to write {\sc flockture} to yield formally identical results to {\sc flock}---{\sc flock} itself
+doesn't yield identical results when it should. Therefore, comparisons
+between {\sc flock} and {\sc flockture} were limited to random initial starting points and so will not 
+be identical, but do display equivalent behavior.
 
@@
 
-\begin{figure*}
-\centering
-\begin{minipage}{.45\textwidth}
-  \centering
+\begin{figure}
+\begin{center}
   \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
-  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
-of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
+  \caption{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate 
+  runs of simulated datasets comprised of five populations with varying
+  levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
   \label{fig:FvFloss}
-\end{minipage}%
-\hfill
-\begin{minipage}{.45\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
-  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of simulated
-data with average pairwise $F_{ST} = $ 0.03. }
-  \label{fig:FvFqi}
-\end{minipage}
-\end{figure*}
+\end{center}
+\end{figure}
 
-Run times for both programs were relatively short, albeit in the case of {\sc flock} 
-non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
-over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
-general increase as population structure became more subtle, however {\sc flock} did not 
-demonstrate a linear increase.
+{\sc flockure} ran much more quickly than {\sc flock}.  {\sc flock} 
+averaged 1966.67 ($\pm 64.32$) seconds per run,
+over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds---a 200-fold improvement in 
+speed.
 
@@
 individuals from another group were separated into two clusters
-with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
-(only present in a single microsatellite dataset) disappeared as population differentiation 
-declined. As the subtlety of genetic structure 
+with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct} A). This phenomenon 
+(only present in a single microsatellite dataset) is an instance of {\sc structure}'s
+Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
 increased loss for both programs increased and differences in the performance of
@@
 values, the variance of the loss incurred by each model increased and certain models did marginally
-better in assignment. No model did consistently better than others, until $F_{ST}$ droped 
+better in assignment. No model did consistently better than others, until $F_{ST}$ dropped 
 below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
 variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
- {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
-values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
+{\sc flockture} did, however, have consistently smaller loss than  {\sc structure} at high to intermediate 
+values of $F_{ST}$. Once $F_{ST}$ fell below 0.016, however, {\sc structure} began correctly assigning a 
 higher proportion of individuals.
@@
 The number and length of plateaus {\sc flockture} encountered quickly 
-diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
-only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
-replicate runs yielded plateau records varying in length from two to eight, all of which had at 
-least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
-contained a plateau of length six or greater. No plateaus were observed for any run when 
-$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
-were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
+diminished as the amount of differentiation among populations decreased. For the microsatellite datasets,
+when $F_{ST}$ was less than 0.0337, no plateau record
+contained a plateau of length six or greater. No plateaus were observed 
+(i.e. no two runs of {\sc flockture} gave exactly the same solution) when 
+$F_{ST}$ was smaller than 0.02187. Results from the SNP datasets
+were concordant.
 Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
-plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
+plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets, 
 graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
@@
 structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
-(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguisable
-and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
+(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguishable
+and results were highly concordant with those of {\sc structure}. As population differentiation declines 
 and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
-similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
-As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
+clusters similar to those from {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
+As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases, the normalized likelihood
 values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.
@@
 a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists.
-Estimating $K$ is a difficult task, and because it is often based on individual assignment we 
-evaluated these programs on their ability to correctly assign indivdiuals at the true value
-of $K$. We have shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
-with non-correlated allele frequencies within the program {\sc structure}. 
+
+We have shown here that the model incorporated by {\sc flock} is analytically equivalent to {\sc structure}'s 
+no-admixture model 
+with non-correlated allele frequencies.
 We adapted the source code of {\sc structure} to create {\sc flockture} and show that
- {\sc flock} and {\sc flockture} are essentially equivalent programs. 
+{\sc flock} and {\sc flockture}  perform equivalently. 
 In comparing results from {\sc flockture} and {\sc structure} 
-on simulated datasets with varying levels of differnetiation among populations
+on simulated datasets with varying levels of differentiation among populations
 we find that the two programs give similar results.
@@
 individuals, as the algorithm proceeds deterministically after the initial random allocation.
-By having a large number of individuals
-that can be arranged into 5 groups defines a large partion space.
 Decreasing $F_{ST}$ leads to more homogenous 
 populations whose multilocus genotypes have a smaller difference in 
-likelihood of assigning to any given cluster. This could act by increasing the number of
+likelihood of assigning to any given cluster. This likely increases the number of
 local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
 caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
-different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.
-In species characterized by large subpopulations and low divergence, as is the case with many 
-marine fish, the use of {\sc flock} should probably be avoided.
+different solutions in complex spaces; however our simulationssuggest this is a more  
+prevalent problem for {\sc flockture}.
+{\sc flock} is not best-suited to species characterized by large
+subpopulations and low divergence, as is the case with many 
+marine fish.
 
@@
 constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
-{\sc flock} were to be implemented in a coding language rather than a VBA routine in Microsoft Excel
-it would be approximately 200 times faster.
+{\sc flock} were to be implemented using good coding practices in a fast language, rather than as
+a VBA routine in Microsoft Excel,
+it would be approximately 200 times faster. This is not an inconsequential difference and is one reason
+we used {\sc flockture} for a large-scale simulation assessment of the {\sc flock} algorithm.
+
+Many researchers wish to estimate $K$, the number of subpopulations, from genetic data.  This is
+a difficult task, statistically, because estimating $K$ is a problem in model selection, and hence can
+be exquisitely sensitive to the priors on nuisance parameters as well as to departures from the underlying
+model (e.g. clinal rather than demic structure). For these reasons, it seems inadvisable to us to
+put too much stock in the inference of $K$ from any program with a simple model of population structure which
+is almost undoubtedly violated in reality. Accordingly, we have based our comparison of 
+{\sc flockture} and {\sc structure} on their ability to correctly cluster individuals when the true value
+of $K$ is known. It should always be kept in mind that real data may depart in unexpected ways from
+simulated data.
 
@@
 different solutions where no plateaus are encountered. Despite the lack of
-plateaus plots of the normalized likelihood values demonstrate that
-{\sc flockture} still produces valuable information. One possible patch for this problem would be running
-multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
-{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, and then 
-determine plateaus amongst {\sc flock}'s replicate solutions not by identical LLOD, but by visual inspection of 
-partitions that
-are very similar but are perhaps not identical; however such an approach would be labor-intensive
-and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} 
-in such cases, keeping in mind, of course, that
-estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
-used. This is especially true in real populations which do not conform to the assumptions of genetic clustering models.
+plateaus, plots of the normalized likelihood values demonstrate that
+{\sc flockture} still produces valuable information, and it might be somewhat
+restrictive to insist that {\sc flock}'s clustering of individuals is useful
+only for the most likely $K$, and only if a long plateau has been
+observed (Pierre Duchesne, pers. comm).
 

### 2015-03-12 tex/response_to_refs_letter.tex
Addressing reviewer comments.

\documentclass[11pt]{letter}
\usepackage{graphicx}
\usepackage{color}

\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.2in
\parindent = 0.0in

\date{}

\address{\mbox{}\vspace*{-1.0in}
\mbox{}\hfill
\includegraphics*{/Users/eriq/Documents/work/nonprj/NMFS_FORMS/NMFS_letterhead.pdf}\\}

\signature{Eric C. Anderson \\ \mbox{} \\ eric.anderson@noaa.gov\\ ph.~831-420-3983}

\newcommand{\reply}[1]{\begin{quotation}\small\sl\textcolor{blue}{#1}\end{quotation}}

\begin{document}

\begin{letter}{ \today

Dr. Nolan Kane\\
News and Views Editor\\
Molecular Ecology Resources \\
}

\opening{Dear Dr.~Kane:}

\reply{Some notes:  using flockture was mandatory because there is no other way to do a reproducible anaysis
with {\sc flock}.  Until people start distributing scriptable, reproducible tools, that is what we 
had to do.}

{\bf Editor Comments to Author:}

This paper, once revised, has the potential to shed light on some of the key similarities and differences among STRUCTURE and FLOCK, both of which are useful and important programs for population genetic analysis.  In that, I agree heartily with reviewer 2. However, reviewer 1 raises many important points that must be dealt with if this is to become acceptable. In particular, it would be important to broaden the scope of the article beyond a single dataset.  There are numerous well-analyzed datasets available these days.  For instance, take:

Recommendations for utilizing and reporting population genetic analyses: the reproducibility of genetic clustering using the program STRUCTURE
KJ Gilbert, RL Andrew, DG Bock, ...
Molecular Ecology 21 (20), 4925-4930

for some approaches that we used to assess repeatability of analytical procedures in quite a number of Structure datasets.  It would not be necessary to go quite as in-depth as in that article, but I do agree with reviewer 1 that a single dataset is insufficient.  Additionally it would be very helpful to use either simulated datasets or other datasets where we have a good idea of what the 'true' answer should be (e.g. some independent prior knowledge of some kind, or at the least a very thorough existing analysis to give reasonable expectations).

Reviewer 1 raises numerous other important points which will improve this paper substantially when dealt with, as does Reviewer 2 (though this reviewer's concerns tend to be much more minor). Despite the large number of suggestions between the reviewers, is should be quite possible to make these changes relatively easily - with the availability of so many existing, well-analyzed datasets it should be relatively easy to run at least a few more through FLOCK and STRUCTURE for a broader set of comparisons.  Alternately, it might make sense to do a fuller analysis of many datasets and re-submit as a full paper. Either would be quite useful to the Molecular Ecology and MER readership.

I look forward to a revised manuscript.

Best regards,
Dr. Nolan Kane
News and Views Editor, Molecular Ecology Resources

{\bf Referee Number 2 Comments:}

Editorial suggestions on Manuscript ID: MER-14-0343 ``Interpreting the FLOCK algorithm from a statistical perspective''
Page and Line numbers are from the pdf file.
\begin{itemize}

\item Page 2, Abstract. Line 18. See ``...no-admxixture...'' Spelling
\reply{fixed}
\item P. 2, Line 19. ``uncorrelated allele frequency prior albeit with more variability...''
\reply{done}
\item P. 3, Line 24. ``The best known example of this class ...''
\reply{done}
\item P. 3, Lines 44-45. ``Being able to interpret Recognizing FLOCK's in this fashion underpinnings should help users ... when it might and might not give substantially...''
\reply{superb edit. done}
\item P. 5. Line 68. ``...to randomly drawn values from the prior, ...''
\reply{done}
\item P. 5, Lines 70-71. ``...at iteration t = 0, 1, 2, ..., T,
1. each ...
\reply{done}
\item P. 6, Line 74. ``for all k = 1,..., K, implying the belief that the subpopulations are equally
represented in the sample before observing the genotypes present.
\reply{great. done.}
\item Page 7, Lines 107-108. ``...the program randomly allocates or each of partitions the N
individuals into (approximately) one of K equal-sized clusters.
\reply{done}
\item Page 7, Lines 109- ...each individual is given the chance to be reallocated (i.e., moved to a different cluster). This is done on the basis of maximum likelihood: non-randomly assigned to the cluster for which its multilocus likelihood is maximal: as Duchesne \& Turgeon say ...''
\reply{This is a great edit. we've incorporated it.}
\item Page 8, Line 138. ``If, all that was desired...'' Delete comma. 
\reply{done}
\item Page 9, Line 155. ``... we can expect that it may be more is susceptible to ...''
\reply{this line got edited out in other changes}
\item Page 9, Line158. ``...the same solution among repeat applications to the same data set less consistently than does STRUCTURE.''
\reply{This line was removed/changed in a previous edit}
\item Page 10, Line 161. ``...to compare the programs FLOCK and STRUCTURE.'' [add] The criteria for comparison include amount of computer time used, estimates of K, and the allocation of individuals among clusters.'' Now move Lines 177 -- 182 to immediately follow, i.e., ``For FLOCK, the user must specify ... is calculated for each individual.'' Continue with Lines 161 -168, i.e., ``FLOCK was run six times ... running 64-bit Windows 7.'' You might also want to explain the choice of 50,000 sweep burn-in and 150,000 sweep sample.
\reply{These lines were completely overhauled given the changes in our simulation assessment}
\item Page 10. Lines 169 `` 175. No change except Page 10, Line 175. ``DISTRICT'' spelling
\reply{done}
\item Page 10, Lines 176 `` 198. Omit Lines 177- 182 that have been moved above. Page 11, Line 197. ``...for estimating K with FLOCK to results obtained from
STRUCTURE using ln P(D) ...''
\reply{eliminated already}
\item Page 12, Lines 209 -- 228. Figure 1 is referenced which uses lower case k, while this text
uses the upper case K.
\reply{eliminated already}
\item Page 12, Line 217. ``Both programs idenfity ...'' Spelling
\reply{done}
\item Page 12, Lines 230 `` 233. ``To investigate the stability of the partition arrangement of individuals within from the reference populations to move around the partition space, we varied ... Unfortuantely Spelling only the ...is given as FLOCK output.''
\reply{done}
\item Page 13, Line 234. ``...during the last later iterations of runs, individuals ...''
\reply{eliminated already}
\item Page 14, Lines 247--248. ``...give similar results, only although at values of K > 4, ...
appear to be more variable among random starting partitions''.
\reply{eliminated already}
\item Page 14, Lines 260--262. ``This, coupled with ...of length zero, complicates ... from
FLOCK. if no non zero plateaus are encountered.
\reply{eliminated already}
\item Page 15, Lines 270--271. `` ... in their comparisons, we did not observe this advantage,
probably for in our analysis. This appears to result from two issues reasons. ``
\reply{the whole discussion of speed was completely revamped once we implemented {\sc flockture}.}
\item Page 16, Lines 302--303. ``This is especially true in real populations and the actual conditions under which samples are drawn which may not conform to the assumptions of genetic clustering models.''
\reply{eliminated already}
\item Page 16, Lines 304--307. ``We made our comparison between FLOCK and the no- admixture model in STRUCTURE which is not the instead of its default with-admixture model. used in STRUCTURE. Rather, by default, STRUCTURE uses its with admixture model. Such a The default model explicitly tries to account for the admixed origin of individuals in the sample and thus may be is more appropriate model than the no admixture model, or FLOCK, when the sample contains such individuals. that are admixed between populations.''
\reply{eliminated already}
\item Page 17, Line 318. ``... with correlated allele frequencies seems like it would seem an interesting challenge, ...''
\reply{eliminated already}
\item Page 22 List of Figures. Lines 386 - 393. Program DISTRUCT plots of cluster membership probabilities of steelhead individuals (qi values from STRUCTURE and the normalized likelihood values from FLOCK). Individual fish are grouped left to right into their geographic samples, and each horizontal plot represents a run of the program
STRUCTURE, or the ``best run'' from a series of runs in FLOCK. Within horizontal plots, each steelhead individual is represented by a vertical bar composed of colored segments of lengths proportional to the various cluster membership probabilities. The numbers in parentheses ... same visually indistinguishable ``best run.''
\reply{This is an excellent, concise description of those plots. We thank the referee for 
providing it, and we have used it.}
\end{itemize}

{\bf Referee Number 1 comments}

I have several important concerns over this paper:

\reply{I'll give brief replies to each of the following here as an outline of how we
changed the paper, and will give more complete details in the following sections}
\begin{enumerate}
\item The conceptual framework is inappropriate since FLOCK is not an MCMC algorithm
\reply{It is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
recognizing it as a special case of well known statistical methods is appropriate.}
\item There is only one dataset which is way too little to come to conclusions
\reply{We agree with the referee. We have undertaken a simulation study across a range of different $F_\mathrm{ST}$ values.}
\item There are basic flaws in the interpretation of the outputs of FLOCK, leading to invalid
comparisons relative to K values other than the estimate
\reply{We have eliminated issues of what can and can't be said about allocation of individuals  at a
certain value of $K$ by doing all the simulations and the analyses at a known value of $K$.}
\item Comparisons of estimates of K cannot be performed with sufficient rigor
\reply{We agree with the referee here.  In fact, we weren't entirely comfortably addressing the issue of estimating
$K$ in the previous version of the manuscript, lest it appear that we were implicitly encouraging it. 
Therefore, we now focus entirely on allocation of individuals.}
\item The speed comparisons are based on an ill-chosen processing unit
\reply{We implemented the {\sc flock} algorithm by modifying the C source code for {\sc structure} and
adding an R wrapper.  This allows a more direct comparison of the amount of computation involved.  If 
{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  But since it
runs 200 times slower than it should for what it is doing, there doesn't seem much speed advantage in 
the data sets we ran.  We report that now in the paper in a way that should be easily understood. }
\end{enumerate}

I now describe each of those concerns in more detail.

{\bf 1) Conceptual framework}

It must be emphasized that FLOCK is not an MCMC based algorithm. It has no target function, no transition probabilities and does not move in small measured probabilistic steps. No amount of mathematical analysis will ever change this.

\reply{I don't believe that we said {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
 case of a simulated annealing algorithm.}

In fact, similar iterative methods have been around for a very long time (e.g. Newton's method for finding roots), much before the first MCMC algorithms were designed (1950?s). This family of methods should be analysed within a truly pertinent conceptual framework (attractors, basins of attraction, fixed points, orbits, etc). The book A First Course in Discrete Dynamical Systems (RA Holmgren, 1994) provides an excellent introduction to this topic. 
\reply{The referee is correct that iterative methods have been around for a long time.
In fact, rather than viewing it as special case of simulated annealing, one could just as well
note that {\sc flock} is a type of gradient descent-algorithm attempting to maximize the
likelihood of the data given a configuration of clusters.
Regardless of what name is given to it, since it is effectively an algorithm
that (intentionally or not) yields solutions with 
high (and possibly maximal) likelihood, we have difficulty understanding
how the referee can honestly declare that, for purposes of analysis of the algorithm,
the conceptual framework of dynamical systems is any more ``truly pertinent'' than
the conceptual framework of statistics.   }

The current popularity of MCMC methods in the field of genetic analysis does not justify trying to cast every current algorithm into an MCMC framework.

\reply{Again, we are not so much casting it in an MCMC framework as we are in a 
statistical/probabilistic framework.  Since it is a problem in statistical inference
it seems to us reasonable to understand it in terms of the probabilities involved. Those
probabilities are, for the most part, identical to those in the program {\sc structure} because
the underlying model is the same in the two programs.  Regarding casting the algorithm in
comparison to {\sc structure}, it makes sense to us to try to understand and explain the {\sc flock}
algorithm in those terms since {\sc structure} is so well known and because the models
underlying it and {\sc flock} are essentially identical.  }

For example:
\begin{itemize}
\item Line 156: it is suggested that Flock can be trapped in a local mode, this is incorrect because Flock does NOT search for maximum values, either local or else
\reply{We thank the referee for pointing out our erroneous language here.  We should not have used
the word ``trapped.''  We have updated that sentence to read, ``As such, we should not be surprised
if it is prone 
to converging to different parts of the posterior probability surface on runs started
from different initial conditions.}

\item Line 254: it is suggested that Flock is searching over all possible partitions. Flock does not sample nor search the surface or space of all partitions.
\reply{We would not have wanted it to sound like we thought {\sc flock} was searching
each and every possible
partition, since this would be infeasible in most problems, so we thank the referee for pointing out that our reference to a space of all partitions might
have confused readers in this way.  We have changed that section appropriately.  However, on whether
or not {\sc flock} samples from the space of partitions, we respectfully submit that the referee is
incorrect when he states that, ``Flock does not sample nor search the surface or space of all partitions.'' 
Consider this: 1) for each run, the initial starting partition in {\sc flock} is chosen (or you could say,
sampled) randomly from amongst
possible partitions with roughly the same number of individuals; 2) the starting partition uniquely
determines the partition that the algorithm will converge to since it is deterministic, 3) different
starting partitions can lead to different final solutions/partitions.  How is it then that each
run in {\sc flock} is {\em not} a sample from the partitions?  }
\item Line 234: ?individuals would alternate between reference groups?: this, precisely, is an orbit
\reply{We removed discussions of individuals alternating between reference groups.}
\end{itemize}

{\bf 2) Data}

Performance comparisons between programs based on a single dataset are of little value. The fact that it is large and real does not compensate. For this purpose, one would expect that a large set of simulations spanning several parameter values with or without several real datasets would be used instead of a single empirical dataset. The advantage of simulations is, of course, that the true genetic structure of the data (essentially K) is known and so the relative performances may be compared on a solid basis. Here, unless I missed something, there is no independent (of the two programs) source of knowing with a reasonable degree of certainty the actual number of clusters represented within this collection of genotypes. In fact, there appears to be several candidate values for K with this particular dataset (see below).

\reply{We agree whole-heartedly and undertake a numerous simulations of data like the real
steelhead data in the previous submission.  Having done this, and analyzed the results,
we can now address the comments below.}
Several (over)generalizations are based on this unique dataset:
\begin{itemize}
\item (p.9 ) FLOCK seems to find the same solution less consistently than does STRUCTURE
\reply{From our simulations using {\sc flockture} this appears to be true.  One need only look
at the variability present in the purple boxplots in Figures~2 and~3 to see that there is more
variability between {\sc flockture} analyses (50 runs of 20 iterations each) on the same data set 
than between {\sc structure} analyses. }
\item (p.14) The problem of getting caught in local modes is not unique to FLOCK...; however our results suggest it is a bigger problem for FLOCK
\reply{The simulation results bear this out, but we have reworded it to, ``This likely 
increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions in 
each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.''
}
\item (p.16) Given FLOCK'S tendency to converge to different solutions with large K and n...
\reply{See above.  Note that we don't vary $K$ now so we don't call it a problem ``with large $K$'' any
longer.}
\item (p.16) It is probably better to use STRUCTURE in such cases...
\reply{We don't say this, but we do point out that {\sc flock} is not best-suited to high-gene-flow
species with very low $F_\mathrm{ST}$.}
\end{itemize}

{\bf 3) Interpretation of FLOCK outputs (purpose of FLOCK, best run, stochasticity of plateau lengths)
Purpose of FLOCK}

I agree with the following statement (p. 11):
``While we take the stance that estimates of K made with any unsupervised clustering algorithm from data on real (non-idealized) populations should always be interpreted and used cautiously...''
I am fully aware that we are dealing with probabilistic statements here, not certainties. However, I am also convinced that the search for the number of genetic units (K) is a central issue. This is precisely what Flock intends to do, and it is also the main task that STRUCTURE was built for.

\reply{We disagree with the referee that estimating $K$ was the ``main task that {\sc structure} was
built for.''  One of us had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
{\sc structure} paper was published. At that time, the members of the Donnelly lab were not eagerly using
{\sc structure} to estimate $K$ from different samples; they were using {\sc structure} to correct
for population admixture when doing association tests in human genetics---our understanding
is {\em that} was the task that structure was original built for.  Furthermore, when I contacted
Jonathan a few years later to ask him some details about how $Ln P(D)$ was calculated by {\sc structure},
he pointed out to me (while getting his notes out of his filing cabinet) that the method for estimating
$K$ was added into the paper to satisfy a referee who had requested it, and that Matthew Stephens 
(one of the co-authors) was not ever extremely pleased to have it in there.  Finally, I teach an annual
course with Matthew Stephens, and each year he explains to the students why estimating $K$ is so
difficult and why estimates of $K$ from {\sc structure} or any other procedure should be 
regarded with circumspection.  }

Thus, the basic purpose of FLOCK is to estimate K and if the estimate is not undecided, to obtain a reasonably accurate partition of the collection of genotypes. Consistency across different values of K is not intended. By considering outputs for any value of K, the authors make invalid comparisons. Comparisons among clustering solutions for different values of K provided by STRUCTURE are commonly reported in the literature, and the successive clustering solutions may at times make biological sense. That practice, however, is not recommended in FLOCK and therefore cannot serve as basis for evaluating its results outside the estimated value of K.

\reply{We do not share the referee's enthusiasm for estimating $K$ nor for the {\em ad-hoc}
methods developed to do so with {\sc flock}.  Therefore, in our simulations and assessments
we limit ourselves in the paper to assessing how well and consistently the programs assign
individuals to the correct cluster, {\em given an analysis run at the correct value of $K$}.
Thus, we no longer compare successive clustering solutions across different values of $K$. We 
understand the referee may have preferred that we assess how well it estimates $K$; however we don't
for several reasons: 1) we don't find it a particularly well-posed problem (what one calls a
subpopulation depends on what they think a subpopulation is, and that is typically not well-specified,
or is specified in terms of a model that makes assumptions that are violated by
the true arrangement or sampling of populations),
2) The approach to estimating $K$ in {\sc flock} is not amenable to easy, automatic application to
numerous simulated data sets, 3) It is more straightforward to interpret program performance in terms
of clustering accuracy; 4) we are not convinced that ``the basic purpose of {\sc flock} is to
estimate $K$.''  If that were the case, then we imagine estimating $K$ would have figured
more prominently in the original 2009 paper introducting {\sc flock}.
}

{\em Misinterpretation of `best run'}

However, the authors are in fact proposing to analyse the output of FLOCK for each K as if it were potentially informative. This, in turn, lead to a misunderstanding of what `best run' means. For instance on p.12, they state:
``For all groups of runs for $K> 2$ no plateaus were observed...it is unclear in these cases what constitutes the `best run' as all plateau lengths are equal to one.''

The answer is simple: there is no best run. The term `best run' applies exclusively to the run with the longest plateau corresponding to the estimate of K. First the stopping rules are applied then the estimation rules. If the estimate is not undecided, the only best run to be considered is the one associated with the estimated K. If the estimate is undecided there is no best run. The documentation and also the flow chart in the latest article (Duchesne and Turgeon 2012) make that clear. In fact, outside the estimated K, no partitions from FLOCK should be considered. Therefore comparisons between clusterings from FLOCK and STRUCTURE for all K except the estimated one are just irrelevant and not valid.

\reply{This seems strange to us.  What happens in the situation where the
``true $K$'' is known and you just want to cluster individuals? If there are no plateaus
amongst the runs at that value of $K$ should one discard {\sc flock} and use
a different program altogether? In most of the simulated data sets we investigated {\sc flock}
like {\sc flockture}
would not have converged to a different solution on each of its 50 runs of 20 iterations---there
would have been no plateaus at all!  and yet it is still delivering useful clustering
inference that is competitive with, and at times better than, {\sc structure}.  We discuss
this rather restrictive guidance on using {\sc flock} in the Discussion.  }

This same approach (= all K outputs are valuable) also brings the authors to lament ``FLOCK's tendency to converge to different solutions with large K'' and conclude that STRUCTURE would be a better choice for complex problems. This is a surprising conclusion since FLOCK is expected to signal the absence of K components precisely by not repeating solutions for that K and thus producing very short plateaus if any. Clearly, by increasing the value of K, the true value will eventually be surpassed and FLOCK will tend to generate numerous solutions and therefore short plateaus ($< 6$). As for the real dataset, there is no ground to believe that $K > 4$ (line 246--248), so that the reported large variability of the FLOCK results may just be the right signal. Here there is a huge misunderstanding that leads the authors to systematically misinterpret potentially correct answers as a sign of inconsistency or lack of power.

\reply{These points are now largely moot, as we don't address the problem of estimating $K$, and we
only analyze simulated data sets at the correct $K$ value.}

{\em Misinterpretation of the stochasticity of plateau lengths}

Another misinterpretation of FLOCK output shows when the plateau records for K = 2 are reported (p.12) for each of the six groups of (50) runs and described as varying greatly. However, the important statistic here is the length of the longest plateau for each K as this will determine the decision on K (the main purpose of FLOCK). Here the top lengths for the six groups are: (7, 5, 5, 4, 5, 3). The corresponding decisions are undecided for 5 groups and K = 2+ for one group. And so there is not so much variation among top lengths and, more importantly, practically none at the decision level (5/6 undecided).

\reply{Again, we don't address the issue of estimating $K$. }

{\bf 4) Comparisons of estimates of K}

(p. 12) The (ln P(D)) values from STRUCTURE was largest at K = 6 which was further supported by
the qi plots, however, the $\Delta K$ method supported a K of 3.

With STRUCTURE, the authors do not provide a definite estimate for K, and they do not provide support for several alleged estimates.

As is, the authors used Flock to reach the decision that STRUCTURE supports an estimate of K=6. Flock needed to go up to K=6 to reach a stopping rule and it is on that basis that they decided to run STRUCTURE up to K=6. This is not consistent with what the method actually prescribes nor with what is often (or should be) reported in the literature: K should be estimated from the peak value for (ln P(D)), not just its highest value but one preceded and succeeded by lower values. It is well known that sometimes the (ln P(D)) values from STRUCTURE increase monotonically with K. If one applied the simple largest value rule then obviously the estimate for K would be as large as the largest K that is tested. Here, that led, incorrectly, to the decision that it is the best estimate because Ln(P(D)) is maximum.

As for K = 3 obtained from the $\Delta K$ method, the results presented do not allow to evaluate whether this estimate is truly supported or not. It is well-known that this method, being based on the rate of change of Ln(P(D)) as a function of K, cannot provide support for the lowest value of K tested (here, K=2). For that reason it is customary to run STRUCTURE from K=1 such that K=2 is a possible estimate.
??
Finally, I note that, based on the same dataset and STRUCTURE, K=5 was preferred in Garza et al. 2014 (Trans. Amer. Fish. Soc.) , but that required considering the regional distribution of clusters, and plots of Ln(P(D)) or ?K were not provided.
At the very least, the results supporting the estimate of K=6 on the basis of the maximum value of Ln(P(D)) (and K= 3 on the basis of ?K for K=1 to a large value) should be provided.
The bottom line is that this dataset does not allow valid comparisons of estimates of K from the STRUCTURE and FLOCK programs, such estimates being the primary task of both programs. There may certainly be cases where FLOCK will provide a different and sometimes less (or more) accurate estimate for K than STRUCTURE, but the comparison made here is not done properly and would in any case concern a single dataset.

\reply{Again.  Largely moot as we no longer focus on estimation of $K$.}

{\bf 5) Speed comparisons}

(p. 11) ``All runs of FLOCK completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes. All runs of STRUCTURE were completed in 486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
(line 281) Though it is difficult to reliably assess differences in program run times when there is no clear consensus on what constitutes a reasonable comparison with respect to run length and number of runs, our results indicate that Flock did not carry a distinct run-time advantage over STRUCTURE''

I agree with the authors that it is difficult to decide on the unit of comparison.
At first sight, their comparison seems to indicate that STRUCTURE is 188.8/80.8 = 2.33 faster than FLOCK. My own opinion is that the right unit of comparison is the run since each run produces the output from a clustering program i.e. a partition of a collection of genotypes into K clusters. Consequently, the speed of execution of cluster programs should be compared on the basis of average run time and not on batches of runs of unequal numbers. If one takes the run as basic unit, given that FLOCK was required to perform 50 runs each time STRUCTURE was required to run once, then the correct speed ratio should be 80.8 X 50/188.8, which means that FLOCK was over 21 times faster than STRUCTURE in clustering this particular dataset. Theoretically, one might argue that one run of STRUCTURE is really the equivalent of 50 runs of FLOCK since 50 runs is the recommended (for validation purposes) number of runs when using FLOCK while STRUCTURE may be run any number of times, including just once. The reason that large numbers of runs are not a standard recommendation for STRUCTURE is precisely that it is too slow. FLOCK is much faster mainly because convergence is very fast (usually in less than 10 iterations = reallocations). It is doubtful that clusterings from STRUCTURE would be of even acceptable quality after 20 iterations. Otherwise why would the authors run a total of 200000 sweeps (one sweep = one iteration), a number consistent with customary usage and recommendations, rather than a mere 20?

Another option that has not been formally evaluated but probably deserves considering would be to compare the total computing time required for a user to reach a decision on the best K estimate, ideally with various datasets for which both programs provide the same K estimates. This would require deciding how many runs of STRUCTURE should be used as there are currently no strict rules on the matter. Perhaps the program should be run following the recommendation of Gilbert et al. (2012, Mol. Ecol. Res.), namely 20 replicates for each K, and a range of K-values truly allowing to estimate the peak or plateau value of Ln(P(D)). Using the dataset presented here, this recommendation implies testing for values larger than K=6, and the range used by Garza et al. (2014) in their original analysis, i.e. K=2 to K=10, seems reasonable. That would be a start for a comparison that truly matters to real users. Among the other people who have run the two programs, I feel that there is a general consensus that FLOCK is much faster than STRUCTURE , and I believe that they are referring to this ?real life? experience. It would be interesting to have a quantitative comparison based on processing time to reach a decision on K.

\reply{Since we modified {\sc structure} to become {\sc flockture} it became clear that the 
there is a direct correspondence between the amount of computation required for one sweep of
{\sc structure} or one reallocation in {\sc flock}. Our discussion of computing speed now reflects
this.  It reads: \\
\mbox{}\\
``One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each execution of {\sc flock}.
Duchesne \& Turgeon (2012) observed faster processing times relative to {\sc structure}.
Having implemented {\sc flockture} it is clear that the amount of computation required for a
single sweep of {\sc structure} is equivalent to an iteration of reallocation of individuals
in {\sc flock}.
Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
computations, one would expect that running flock once with 50 runs, each of 20 iterations,
would require about the same amount of time as running {\sc structure} on the same
data set for 1,000 sweeps.  Since it is recommended to run {\sc structure} longer than 1,000 sweeps, {\sc flock} should have a speed advantage. However, {\sc flock} requires far more 
time for each computation
than {\sc structure}, as is clear from the fact that it runs 200 times slower than 
{\sc flockture}.  Thus, for data sets comparable to the ones we simulated, one can expect
50 runs of 20 iterations in {\sc flock} (a standard execution of the program) to require as
much time as running {\sc structure} for 200,000 sweeps. This does not constitute
a speed advantage.
If {\sc flock} were to be implemented using good coding practices in a fast language,
rather than as a VBA routine in Microsoft Excel,
it could be made much faster. As it is, we found it preferable to implement our own version
in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.''}

{\bf Additional comment:}

The authors mention the existence of a set of stopping rules in FLOCK but not their absence in STRUCTURE and the possible consequences thereof. For example, if no peak has yet been observed for (ln P(D)) values from STRUCTURE, how long should one process ever larger values of K? Interestingly, this is exactly the situation with the empirical dataset discussed in the paper. The authors chose 6 as the upper bound for K, a value clearly derived from the upper bound from the FLOCK stopping rules. Now this in turn lead them to suggest that K = 6 was to be considered as a plausible number of populations since (ln P(D)) was highest at K = 6, which is obviously a misapplication of the estimation rules for K with STRUCTURE. Then they noticed the large number of solutions from FLOCK, while implicitly assuming the validity of the K = 6 estimate. From this they concluded that FLOCK `seems to find the same solution less consistently than does STRUCTURE'. As explained above, if K = 6 were wrong, and there is no evidence to the contrary, then FLOCK would be expected to produce numerous solutions and the more the better, thereby indicating that there is no support for K=6. And so the absence of stopping rules in STRUCTURE combined with a misapplication of the estimation rule for K for STRUCTURE and a misunderstanding of FLOCK plateau analysis lead the authors to a specific conclusion not supported by their data and, worse still, one that they generalized over the entire domain of application of FLOCK.

\reply{Moot, again, as we don't focus on estimating $K$.}

{\bf Recommendations:}

Based on the numerous problems I have described, I think this paper unfit to be published in Molecular Ecology Resources. If the authors intend to establish that `FLOCK yields results similar to STRUCTURE with the no-admixture model and uncorrelated allele frequency prior', they should do so by running FLOCK, STRUCTURE with the no-admixture model and STRUCTURE with the with-admixture model over a large collection of datasets spanning numerous combinations of pertinent parameter values.

\reply{We have done that}

The object of the comparisons should be the K-value estimated by each algorithm, not the solutions for each processed K-value.

\reply{We disagree for the reasons listed above.  We have compared the programs on the basis of the
solutions for the {\em correct} $K$-value.}

The datasets should be mostly
from simulations so as to compare the performance of each algorithm on a solid basis and also, possibly, on some real datasets whose genetic structure is already well known.

\reply{We chose to use simulated data entirely.}

This, of course, takes time but there is no analytical shortcut to this work. This data-driven research would hit a double target: compare FLOCK to the two models of STRUCTURE and compare the performances of the two STRUCTURE models on a broad empirical basis. The results would be very useful to all potential users of clustering programs and would practically ensure that they are not misled into false conceptions.

\reply{We agree with the referee that the present version of our manuscript should be very useful.}

They should be published in a full-fledged article, not just a comment.

\reply{If, upon reading the present version of our manuscript, the editor feels it would be appropriate
to publish it as a full article, we would agree to that.
}

I also suggest that the authors make an additional effort to better understand plateau analysis especially in connection with variability of solutions i.e. lengths of plateaus. Short plateaus (numerous solutions) are the right answer whenever K is the wrong answer to the `number of populations' problem. If run times are to be compared, the comparisons should follow the above recommendation such that a real user gets a simple message out of it.

Pierre Duchesne

\closing{Best regards,}
\setlength{\topmargin}{0in}
\textheight = 9 in
\end{letter}

 \end{document}

### 2015-03-12 tex/main-body-text.tex
Addressing reviewer comments.

@@
 considerable attention in the molecular ecology literature.  
-The best known example of this class of methods is {\sc structure} 
+The best known of this class of methods is {\sc structure} 
 \citep{Pritchardetal2000}, with over 12,000 citations identified on 
@@
 
-Being able to interpret {\sc flock} in this fashion should help users
+Recognizing {\sc flock}'s underpinnings in this fashion should help users
 interpret output from {\sc flock} and predict when it might and might not
@@
 
-After initializing the unknown allele frequencies, $\theta = (\theta_1,\ldots,\theta_K)$, to randomly drawn values,
+After initializing the unknown allele frequencies, $\theta = (\theta_1,\ldots,\theta_K)$, to randomly drawn values from the prior,
 $\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
-$Z_i$'s and $\theta$ using Gibbs sampling.  That is,
-at iteration $t = 0, 1, 2, \ldots$:
-\begin{enumerate}
+$Z_i$'s and $\theta$ using Gibbs sampling.  That is, if Gibbs sampling is run for $T$ iterations,
+at iteration $t =0, 1, 2, \ldots, T-1$:
+\begin{itemize}
 \item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
@@
 distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
-prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
+prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$, implying the belief that the subpopulations are equally
+represented in the sample before observing the genotypes present.    
 \item $\theta^{(t)}$ is updated to $\theta^{(t+1)}$ from its full conditional distribution.  For each 
@@
 individuals with $Z^{(t+1)}_i = k$.   
-\end{enumerate}
+\end{itemize}
 
@@
 The first step of the {\sc flock} algorithm is initialization, during which 
-the program randomly allocates each of the $N$ individuals to one of 
-$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
-these steps, each individual is given the chance to be reallocated 
-(\ie moved to a different cluster).  This is done on the basis of 
-maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
+the program randomly allocates or partitions the $N$ individuals into
+approximately equal-sized
+clusters.  Then a number of reallocation steps are performed.  During every one of 
+these steps, each individual is non-randomly assigned to the cluster for which its multilocus likelihood is maximal: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
 performed following multilocus maximum likelihood (Paetkau et al. 1995).''
@@
 the algorithm.''  As such, we should not be surprised if it is prone 
-to becoming trapped in local modes of the posterior probability, though
-the practical implications of that are not yet clear.
+to converging to different parts of the posterior probability surface on runs started
+from different initial conditions.
 
@@
   \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
-  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
-  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
- runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
+  \caption{
+  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
+  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
+  Individuals are grouped left to right into their simulated population samples, and each 
+  horizontal plot represents a run of the program
+  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
+  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
+  Within horizontal plots, each individual is represented by a vertical bar composed of colored
+  segments of lengths proportional to the various cluster membership probabilities.
+  Each horizontal plot
+  represents one of nine 
+ runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
   frequency (A.C) model, 
 the no-admixture and correlated allele frequency (NA.C) model, 
-and the no-admixture and non-correlated allele frequency (NA.NC) model. Each simulation is comprised of
+and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
+parentheses following the model names give the number, out of 9, of replicate runs
+that obtained the same visually-indistinguishable estimate. Each simulation is comprised of
  five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
@@
 {\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
-and became somewhat inconsistent at lower values of $F_{ST}$.
-This is likely a consequence of the low population divergence and also
-the large size of the search space over possible partitions. 
-The likelihood surface defined on partitions of this space will often
+and became inconsistent at lower values of $F_{ST}$.
+This is likely a consequence of the low population divergence which
+makes it possible for many different allocations of individuals to
+have roughly the same likelihood.
+The likelihood surface defined on partitions of this space is then often
 be multimodal---possessing numerous local peaks and troughs.
@@
 likelihood of assigning to any given cluster. This likely increases the number of
-local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
-caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
-different solutions in complex spaces; however our simulationssuggest this is a more  
+local peaks, leading {\sc flockture} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
+is not unique to {\sc flock}---{\sc structure} also can converge to 
+different solutions in complex spaces; however our simulations suggest this is a more  
 prevalent problem for {\sc flockture}.
@@
 One of the advantages of {\sc flock} over other methods for genetic clustering
-is reported to be the short processing time for each run of {\sc flock}.
-While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
-their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
-$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
-were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
-a 50,000 sweep burn-in followed by a 200,000 sweep sample 
-from the posterior for values of $K$ from 1 to three more than the true 
-value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
-constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
-{\sc flock} were to be implemented using good coding practices in a fast language, rather than as
-a VBA routine in Microsoft Excel,
-it would be approximately 200 times faster. This is not an inconsequential difference and is one reason
-we used {\sc flockture} for a large-scale simulation assessment of the {\sc flock} algorithm.
+is reported to be the short processing time for each execution of {\sc flock}.
+\citet{Duc&Tur2012} observed faster processing times relative to {\sc structure}.
+Having implemented {\sc flockture} it is clear that the amount of computation required for a
+single sweep of {\sc structure} is equivalent to an iteration of reallocation of individuals
+in {\sc flock}.
+Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
+computations, one would expect that running flock once with 50 runs, each of 20 iterations,
+would require about the same amount of time as running {\sc structure} on the same
+data set for 1,000 sweeps.  Since it is recommended to run {\sc structure} longer than 1,000 sweeps, {\sc flock} should have a speed advantage. However, {\sc flock} requires far more 
+time for each computation
+than {\sc structure}, as is clear from the fact that it runs 200 times slower than 
+{\sc flockture}.  Thus, for data sets comparable to the ones we simulated, one can expect
+50 runs of 20 iterations in {\sc flock} (a standard execution of the program) to require as
+much time as running {\sc structure} for 200,000 sweeps. This does not constitute
+a speed advantage.
+If {\sc flock} were to be implemented using good coding practices in a fast language,
+rather than as a VBA routine in Microsoft Excel,
+it could be made much faster. As it is, we found it preferable to implement our own version
+in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.
 

### 2015-03-12 tex/response_to_refs_letter.tex
almost done!

@@
 
-\reply{I'll give brief replies to each of the following here as an outline of how we
-changed the paper, and will give more complete details in the following sections}
+\reply{I'll give brief replies to each of the following here in outline,
+and will give more complete details in the following sections.}
 \begin{enumerate}
 \item The conceptual framework is inappropriate since FLOCK is not an MCMC algorithm
-\reply{It is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
+\reply{{\sc flock} is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
 recognizing it as a special case of well known statistical methods is appropriate.}
@@
 to converging to different parts of the posterior probability surface on runs started
-from different initial conditions.}
+from different initial conditions.''}
 
@@
 sampled) randomly from amongst
-possible partitions with roughly the same number of individuals; 2) the starting partition uniquely
+possible partitions with roughly the same number of individuals in each cluster; 2) the starting partition uniquely
 determines the partition that the algorithm will converge to since it is deterministic, 3) different
@@
 run in {\sc flock} is {\em not} a sample from the partitions?  }
-\item Line 234: ?individuals would alternate between reference groups?: this, precisely, is an orbit
+\item Line 234: ``individuals would alternate between reference groups'': this, precisely, is an orbit
 \reply{We removed discussions of individuals alternating between reference groups.}
@@
 
-\reply{We agree whole-heartedly and undertake a numerous simulations of data like the real
+\reply{We agree whole-heartedly and undertake numerous simulations of data like the real
 steelhead data in the previous submission.  Having done this, and analyzed the results,
@@
 \item (p.9 ) FLOCK seems to find the same solution less consistently than does STRUCTURE
-\reply{From our simulations using {\sc flockture} this appears to be true.  One need only look
+\reply{From our simulations using {\sc flockture} this is clearly true.  One need only look
 at the variability present in the purple boxplots in Figures~2 and~3 to see that there is more
@@
 increases the number of
-local peaks, leading {\sc flockture} to settle on many different solutions in 
-each run. The problem presented by having multiple modes in the likelihood surface 
+local peaks, leading {\sc flockture} to settle on different solutions in 
+different runs. The problem presented by having multiple modes in the likelihood surface 
 is not unique to {\sc flock}---{\sc structure} also can converge to 
@@
 \item (p.16) It is probably better to use STRUCTURE in such cases...
-\reply{We don't say this, but we do point out that {\sc flock} is not best-suited to high-gene-flow
+\reply{We don't use this phrase specifically, but we do point out that {\sc flock} is not best-suited to high-gene-flow
 species with very low $F_\mathrm{ST}$.}
@@
 \reply{We disagree with the referee that estimating $K$ was the ``main task that {\sc structure} was
-built for.''  One of us had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
+built for.''  I had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
 {\sc structure} paper was published. At that time, the members of the Donnelly lab were not eagerly using
@@
 is {\em that} was the task that structure was original built for.  Furthermore, when I contacted
-Jonathan a few years later to ask him some details about how $Ln P(D)$ was calculated by {\sc structure},
+Jonathan a few years later to ask him some details about how $\ln P(D)$ was calculated by {\sc structure},
 he pointed out to me (while getting his notes out of his filing cabinet) that the method for estimating

### 2015-03-12 tex/response_to_refs_letter.tex
Further tweaks and edits.  Very close to resubmission.

@@
 
-\reply{Some notes:  using flockture was mandatory because there is no other way to do a reproducible anaysis
-with {\sc flock}.  Until people start distributing scriptable, reproducible tools, that is what we 
-had to do.}
+Thank you for the opportunity to resubmit our revised manuscript, MER-14-0343 ``Interpreting the FLOCK algorithm
+from a statistical perspective.''  We are very grateful to you and to both of the referees for the many
+suggestions that we have incorporated.  We feel they have improved the manuscript greatly.  We have taken your
+advice and undertaken a set of simulations that provide a thorough assessment of {\sc flock} versus {\sc 
+structure} on data sets with large numbers (2,000) of individuals, like our original steelhead data set.
+We have dropped the steelhead data set from the paper.
+
+In order to conduct these simulations we implemented the {\sc flock} algorithm by modifying the source code of
+the program {\sc structure}.  We used the resulting program, {\sc flockture}, because {\sc flock} itself
+is 200 times slower than {\sc flockture} and it is also not amenable to scripting and repeated application to 
+multiple data sets in batch mode.  It is simply not possible to prepare a reproducible set of simulations using
+the point-and-click interface of {\sc flock}.
+We show simulations in the paper to verify that {\sc flockture} and {\sc flock} yield equivalent results before comparing {\sc structure} to {\sc flockture}.
+
+Another change in the paper is that we focus on the performance of the programs for clustering individuals while
+assuming the correct value of $K$.  We were actually somewhat uncomfortable in our initial submission that
+assessing {\sc flock}'s ability to estimate $K$ might give the impression that we recommend such pursuits---an
+endeavor that I (along with {\sc structure} co-author Matthew Stephens) are reluctant to recommend, especially
+when population structure is weak.  So, we are pleased to be able to offer a rigorous evaluation of program
+performance without going too deeply into the {\em ad-hoc} procedures for estimating $K$ with {\sc flock}.
+
+Overall, we discovered through our simulations that results from the {\sc flock} algorithm tend to be less 
+consistent than {\sc structure}'s, especially at low values of $F_\mathrm{ST}$.  But, on the other hand, we were 
+quite interested to see that the algorithm still worked quite well in some cases.  In fact, it outperformed {\sc structure} 
+using SNPs at intermediate levels of differention.  We also found that in large data sets like those that
+we simulated, {\sc flock}'s individual runs seldom (or never) converge to exactly the same partition.  This 
+presents some serious problems for applying the Duchesne \& Turgeon (2012) recipe for estimating $K$, and we 
+note that in the discussion.
+
+Below, we address individual comments from you and the referees.  Our replies are in \textcolor{blue}{{\sl blue slant}} font.
+
+Thank you again for your service.  As usual, we have been greatly pleased with the review process
+at {\em Molecular Ecology Resources} and are excited with how much improvement the referees have brought
+to our paper.  We apologize that it has required several months to return these revisions, but do appreciate
+the extra time the journal granted us---it was necessary to write {\sc flockture} and to undertake the
+simulations.
+
+Best regards,
+
+Eric C. Anderson
+
+\vspace*{3em}
 
@@
 
+\reply{Thank you for your recommendations.  We have undertaken a set of simulations as suggested.}
+
 Reviewer 1 raises numerous other important points which will improve this paper substantially when dealt with, as does Reviewer 2 (though this reviewer's concerns tend to be much more minor).
@@
 Either would be quite useful to the Molecular Ecology and MER readership.
+
+\reply{We address all reviewer comments in the revision, and detail our responses below.  We are re-submitting
+our manuscript as a revision of our first version; however if you thought that it warranted
+reclassification and publication as
+a full paper rather than as a comment we would be amenable to that.  Thank you.}
 
@@
 adding an R wrapper.  This allows a more direct comparison of the amount of computation involved.  If 
-{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  But since it
-runs 200 times slower than it should for what it is doing, there doesn't seem much speed advantage in 
-the data sets we ran.  We report that now in the paper in a way that should be easily understood. }
+{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  We report that now in the paper in a way that should be easily understood. }
 \end{enumerate}
@@
 
-\reply{I don't believe that we said {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
+\reply{We didn't say {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
  case of a simulated annealing algorithm.}
@@
 that (intentionally or not) yields solutions with 
-high (and possibly maximal) likelihood, we have difficulty understanding
-how the referee can honestly declare that, for purposes of analysis of the algorithm,
-the conceptual framework of dynamical systems is any more ``truly pertinent'' than
-the conceptual framework of statistics.   }
+high (and possibly maximal) likelihood, we find it quite clear that  
+the conceptual framework of statistics is at least as ``truly pertinent,''
+for purposes of analysis of the algorithm, as is
+the conceptual framework of dynamical systems.
+}
 
@@
 have confused readers in this way.  We have changed that section appropriately.  However, on whether
-or not {\sc flock} samples from the space of partitions, we respectfully submit that the referee is
+or not {\sc flock} samples from the space of partitions, we contend that the referee is
 incorrect when he states that, ``Flock does not sample nor search the surface or space of all partitions.'' 
@@
 starting partitions can lead to different final solutions/partitions.  How is it then that each
-run in {\sc flock} is {\em not} a sample from the partitions?  }
+run in {\sc flock} is {\em not} a sample from the partitions? It is quite clear to us that it is.  }
 \item Line 234: ``individuals would alternate between reference groups'': this, precisely, is an orbit
@@
 amongst the runs at that value of $K$ should one discard {\sc flock} and use
-a different program altogether? In most of the simulated data sets we investigated {\sc flock}
-like {\sc flockture}
-would not have converged to a different solution on each of its 50 runs of 20 iterations---there
-would have been no plateaus at all!  and yet it is still delivering useful clustering
+a different program altogether? In most of the simulated data sets we investigated {\sc flock},
+like {\sc flockture},
+would not have seen any plateaus at all!---and yet it is still delivering useful clustering
 inference that is competitive with, and at times better than, {\sc structure}.  We discuss
@@
 
-\reply{These points are now largely moot, as we don't address the problem of estimating $K$, and we
+\reply{These points are now largely moot, as we don't focus on the problem of estimating $K$, and we
 only analyze simulated data sets at the correct $K$ value.}
@@
 Short plateaus (numerous solutions) are the right answer whenever K is the wrong answer to the `number of populations' problem.
+
+\reply{We have taken the referee's suggestion and worked on obtaining a better understanding of plateau analysis.
+Short plateaus may well be right answer whenever $K$ is the wrong answer.  However, short plateaus are also
+the wrong answer when the $K$ value is correct.  This seems to be major problem for {\sc flock} on data sets
+of the size that we have simulated.  We've addessed this in the 2nd to last paragraph of the discussion. }
+
 If run times are to be compared, the comparisons should follow the above recommendation such that a real user gets a simple message out of it.
 
+\reply{See above.}
 Pierre Duchesne
 
-\closing{Best regards,}
-\setlength{\topmargin}{0in}
-\textheight = 9 in
 \end{letter}

### 2015-03-12 tex/main-body-text.tex
Further tweaks and edits.  Very close to resubmission.

@@
 relies on MCMC samples from the posterior distribution, while
-{\sc flock}'s inference involves finding a local maximum in the posterior surface.
+{\sc flock}'s inference involves finding local maxima in the posterior surface.
 
@@
 algorithm performance can then be measured with a zero-one 
-loss function (where a loss of 1 is incurred if an individual is group with the wrong cluster).
+loss function (where a loss of 1 is incurred for each individual that is grouped with the wrong cluster).
 
@@
 Microsoft Excel retards rigorous evaluation of its behavior due 
-the its slow execution speed and 
+to its slow execution speed and 
 the inability to efficiently batch-process multiple data sets. Consequenty, we adapted 
@@
 (code provided on GitHub repository https://github.com/eriqande/flockture) 
-in order to accurately assess the strengths and weaknesses of the {\sc flock} 
+in order to accurately and reproducibly assess the strengths and weaknesses of the {\sc flock} 
 algorithm. First, using simulated data we demonstrate that {\sc flock} 
@@
 
-For each set of runs, {\sc flock} reports the 
-normalized likelihood values (the exponentiated log-likelihood value for an individual's 
-membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
-longest plateau---refered to as the `best run' by \citep{Duc&Tur2012} (see below).
-These values can be compared to the $q_i$ values
-returned by {\sc structure}, which, in the no-admixture case, are posterior 
-probabilities of cluster membership. In {\sc flock} the user must specify both the number of runs 
+For a single excecution of {\sc flock} the user must specify both the number of runs 
 (different initial allocations of individuals 
@@
 membership to all $K$ clusters is computed
-and the log likelihood difference (LLOD), the difference between the 
+and the log likelihood difference (LLOD)---the difference between the 
 cluster with the highest likelihood 
-and second highest likelihood, is calculated for each individual. {\sc flock} 
+and second highest likelihood---is calculated for each individual. {\sc flock} 
 identifies runs that have converged to the same solution by detecting 
 that they have the same mean LLOD score, and, in 
-{\sc flock} parlance, a series of runs that have converged to the same 
-solution are termed a ``plateau.'' The plateau record is a list of 
-how many runs converged to the same allocation
+{\sc flock} parlance multiple runs that have converged to the same 
+allocation of individuals are termed a ``plateau.'' The plateau record is a list of 
+how many runs in the program execution have converged to the same allocation
 of individuals.
+When all 50  runs have completed, 
+{\sc flock} reports as its solution the cluster membership and 
+normalized likelihood values (the exponentiated log-likelihood value for an individual's 
+membership to each of $K$ clusters, scaled to sum to one) for one of the 50 runs belonging to the 
+longest plateau---refered to as the `best run' by \citep{Duc&Tur2012}.
+These normalized likelihood values can be compared to the $q_i$ values
+returned by {\sc structure}, which, in the no-admixture case, are posterior 
+probabilities of cluster membership.
 
 To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
-were simulated using a coalescent based program (code provided on GitHub repository 
-https://github.com/eriqande/flock-comment/simdata). The 5 populations
+were simulated using a coalescent-based program (code provided at 
+https://github.com/eriqande/flock-comment/). The 5 populations
 were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
@@
 representation of {\sc flock} based on the
-following criteria: (1) Clustering of individuals into corrected population as
+following criteria: (1) Clustering of individuals into correct population as
  defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
@@
 {\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
-(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
-range of population differentiation. The 96 datasets were simulated as above (five populations 
-composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
+(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader range
+of population differentiation. The 96 datasets were simulated as above (five populations 
+composed of 250, 250, 400, 550, and 550 individuals with 15 loci for the microsatellites
 and 96 loci for the SNPs)
@@
 {\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
-from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
+from the posterior distribution. Each dataset was analyzed 9 times by each of the different
+methods. All runs were performed on a Macbook with a 2.3 GHz i7 processor
 and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
@@
 while the program {\sc distruct} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.
+
+When {\sc flock} does not have a `best run' because the plateau record contains no plateaus, the program still returns a solution, but it is not clear how that solution is chosen.  In
+our analyses  with {\sc flockture}, when the plateau record did not indicate a `best run,' we used the
+run with the highest likelihood as the solution.
 
@@
 not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the
+inner workings of {\sc flock} 
 we took a subset of the 
@@
 As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
-well differentiated populations. Infrequently, at higher levels of population differentiation the 
+well-differentiated populations. Infrequently, at higher levels of population differentiation the 
 NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
@@
 Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
-increased loss for both programs increased and differences in the performance of
+increased, loss for both programs increased and differences in the performance of
 each model with different marker types emerged.
@@
 better in assignment. No model did consistently better than others, until $F_{ST}$ dropped 
-below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
+below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among-run
 variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
@@
 graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
-likelihood were suggestive of population structure.
+likelihood were indicative of population structure (Fig~\ref{fig:FvSdistruct}).
 
@@
 parentheses following the model names give the number, out of 9, of replicate runs
-that obtained the same visually-indistinguishable estimate. Each simulation is comprised of
+that obtained the same visually-indistinguishable estimate.
+(Note that although many of {\sc flocktures}'s solutions in B--D were visually indistinguishable, they were not identical partitions, and would
+hence be considered different for the purposes of creating a {\sc flock} plateau record.) Each simulation is comprised of
  five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
@@
 between the two programs at large values of $F_{ST}$, 
-{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
-and became inconsistent at lower values of $F_{ST}$.
+{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
+replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
 This is likely a consequence of the low population divergence which
@@
 be multimodal---possessing numerous local peaks and troughs.
-The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
+The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
 individuals, as the algorithm proceeds deterministically after the initial random allocation.
@@
 likelihood of assigning to any given cluster. This likely increases the number of
-local peaks, leading {\sc flockture} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
+local peaks, leading {\sc flockture} and {\sc flock} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
 is not unique to {\sc flock}---{\sc structure} also can converge to 
@@
 Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
-computations, one would expect that running flock once with 50 runs, each of 20 iterations,
+computations, one would expect that executing {\sc flock} once with 50 runs, each of 20 iterations,
 would require about the same amount of time as running {\sc structure} on the same
@@
 it could be made much faster. As it is, we found it preferable to implement our own version
+of the {\sc flock} algorithm
 in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.
@@
 model (e.g. clinal rather than demic structure). For these reasons, it seems inadvisable to us to
-put too much stock in the inference of $K$ from any program with a simple model of population structure which
-is almost undoubtedly violated in reality. Accordingly, we have based our comparison of 
+put too much stock in the inference of $K$ from any program, especially when population structure
+is subtle. Accordingly, we have based our comparison of 
 {\sc flockture} and {\sc structure} on their ability to correctly cluster individuals when the true value
-of $K$ is known. It should always be kept in mind that real data may depart in unexpected ways from
-simulated data.
+of $K$ is known.
 
@@
 sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
-By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
+By relying on the number of identical partitions obtained by {\sc flock} over multiple
 runs as a means of supporting 
 a particular $K$,  similar---but not identical---partitions
-will provide no support to a given $K$. We have shown that even at intermediate
-values of $F_{ST}$ {\sc flockture} has a tendency to converge to 
-different solutions where no plateaus are encountered. Despite the lack of
-plateaus, plots of the normalized likelihood values demonstrate that
-{\sc flockture} still produces valuable information, and it might be somewhat
-restrictive to insist that {\sc flock}'s clustering of individuals is useful
-only for the most likely $K$, and only if a long plateau has been
-observed (Pierre Duchesne, pers. comm).
+will provide no support to a given $K$. This could be problematic, 
+since we have shown that even at intermediate
+values of $F_{ST}$,  {\sc flockture} (and, by extension,
+{\sc flock}) may converge to 
+different cluster allocations in every run, so that no plateaus are encountered.
+
+The above presents a serious conundrum to the user of {\sc flock}.  On the one hand,
+{\sc flock} clearly continues to provide useful inference about population structure
+(as evidenced by the {\sc distruct} plots in Fig.~\ref{fig:FvSdistruct}), even when its
+plateau record includes no plateaus; however, on the other hand, if there were no
+plateaus for $K=5$ in our simulated data, then the {\sc flock} method would lead to an estimate of $K$ that
+was either "undecided" or some value less than $K$. The authors of {\sc flock}
+discourage users from making any inference at a value of $K$ that is not
+the value estimated by {\sc flock}, because, ``outside the estimated $K$, no partitions from {\sc flock}
+ should be considered. Therefore\ldots clusterings from {\sc flock}\ldots for all $K$ except the estimated one are just
+ irrelevant and not valid''(Pierre Duchesne, pers. comm).  To be specific, following 
+this advice of {\sc flock}'s authors would have led to discarding {\sc flockture}'s
+results for any $F_\mathrm{ST} < 0.032$ (almost 90\% of our simulated data sets!), even those that are
+manifestly quite accurate---some competitive
+with or even superior to {\sc structure}'s estimate!  We surmise that this problem with {\sc flock} 
+guidance issues from the fact that the {\em ad-hoc} rules developed by
+\citeauthor{Duc&Tur2012} for estimating $K$ with {\sc flock} were developed empirically
+by reference to a small set of simulations over a fairly limited range of sample sizes.  Since our simulated
+samples (made to mimic the sorts of large, real data sets we collect in our labs)
+fall outside the range of values explored
+by \citeauthor{Duc&Tur2012} it appears that {\sc flock}'s rules of use are inappropriate, 
+and perhaps over-restrictive for these data sets.  Researchers with data that similarly fall beyond the range of 
+data sets tested in \citeauthor{Duc&Tur2012} may wish to approach {\sc flock} analysis with
+caution, or may choose to rely on other programs.
 
@@
 We are grateful to Kristen Ruegg and Robin Waples for helpful comments on a draft of this paper.
+Comments and suggestions from an anonymous referee and Pierre Duchesne helped to strengthen the
+paper considerably.
 P. Barry was supported by a fellowship from the Rasmuson Foundation.

### 2015-03-12 tex/response_to_refs_letter.tex
a few more late night changes.

@@
 \item Comparisons of estimates of K cannot be performed with sufficient rigor
-\reply{We agree with the referee here.  In fact, we weren't entirely comfortably addressing the issue of estimating
+\reply{We agree with the referee here.  In fact, we weren't entirely comfortable addressing the issue of estimating
 $K$ in the previous version of the manuscript, lest it appear that we were implicitly encouraging it. 
@@
 likelihood of the data given a configuration of clusters.
-Regardless of what name is given to it, since it is effectively an algorithm
+Regardless of what name is given to it, since it is an algorithm
 that (intentionally or not) yields solutions with 
@@
 \item (p.16) It is probably better to use STRUCTURE in such cases...
-\reply{We don't use this phrase specifically, but we do point out that {\sc flock} is not best-suited to high-gene-flow
+\reply{We no longer use this phrase specifically, but we do point out that {\sc flock} is not best-suited to high-gene-flow
 species with very low $F_\mathrm{ST}$.}
@@
 for population admixture when doing association tests in human genetics---our understanding
-is {\em that} was the task that structure was original built for.  Furthermore, when I contacted
+is {\em that} was the task that structure was originally built for.  Furthermore, when I contacted
 Jonathan a few years later to ask him some details about how $\ln P(D)$ was calculated by {\sc structure},
@@
 amongst the runs at that value of $K$ should one discard {\sc flock} and use
-a different program altogether? In most of the simulated data sets we investigated {\sc flock},
+a different program altogether? In most of the simulated data sets we investigated, {\sc flock},
 like {\sc flockture},

### 2015-03-12 tex/main-body-text.tex
a few more late night changes.

@@
 \includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatLoss.pdf}%uSatloss_BW.pdf is a greyscale version
-  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
+  \caption{Zero-one loss (lower values = better performance) for nine runs of {\sc flockture} and the admixture and correlated allele 
   frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
@@
 likelihood of assigning to any given cluster. This likely increases the number of
-local peaks, leading {\sc flockture} and {\sc flock} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
+local peaks, leading {\sc flockture} and {\sc flock} to settle on settle on different solutions in 
+different runs. The problem presented by having multiple modes in the likelihood surface 
 is not unique to {\sc flock}---{\sc structure} also can converge to 
 different solutions in complex spaces; however our simulations suggest this is a more  
-prevalent problem for {\sc flockture}.
+prevalent problem for the {\sc flock} algorithm.
 {\sc flock} is not best-suited to species characterized by large

### 2015-03-12 tex/response_to_refs_letter.tex
Revised version as submitted.

@@
 consistent than {\sc structure}'s, especially at low values of $F_\mathrm{ST}$.  But, on the other hand, we were 
-quite interested to see that the algorithm still worked quite well in some cases.  In fact, it outperformed {\sc structure} 
+quite interested and pleased to see that the algorithm still works quite well in some cases.  In fact, it outperformed {\sc structure} 
 using SNPs at intermediate levels of differention.  We also found that in large data sets like those that
@@
 to our paper.  We apologize that it has required several months to return these revisions, but do appreciate
-the extra time the journal granted us---it was necessary to write {\sc flockture} and to undertake the
+the extra time the journal granted us---the extra time was necessary to write {\sc flockture} and to undertake the
 simulations.
@@
 
+\vspace*{3em}
+
 Eric C. Anderson
 
-\vspace*{3em}
+\vspace*{6em}
 
@@
 Short plateaus may well be right answer whenever $K$ is the wrong answer.  However, short plateaus are also
-the wrong answer when the $K$ value is correct.  This seems to be major problem for {\sc flock} on data sets
+the wrong answer when the $K$ value is correct.  This seems to be a major problem for {\sc flock} on data sets
 of the size that we have simulated.  We've addessed this in the 2nd to last paragraph of the discussion. }

### 2015-03-12 tex/abstract.tex
Addressing reviewer comments.
Revised version as submitted.

@@
 levels of population differentiation for both microsatellite and SNP genotypes. We analyzed them
-with {\sc flockture} and {\sc structure} and assessed each program on its ability to correctly cluster
+with {\sc flockture} and {\sc structure} and assessed each program on its ability to cluster
 individuals to their correct subpopulation.  We show that
-{\sc flockture} yields results similar to {\sc structure} although with greater 
+{\sc flockture} yields results similar to {\sc structure} albeit with greater 
 variability from run to run. {\sc flockture} did perform better than {\sc structure} 
@@
 ($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
-for both marker types. Interpreting {\sc flock}'s algorithm as a special case of the model in 
+for both marker types. On large data sets like those we simulated, it appears that 
+{\sc flock}'s reliance on inference rules regarding its ``plateau record'' are not helpful. 
+Interpreting {\sc flock}'s algorithm as a special case of the model in 
 {\sc structure} should aid in understanding the program's output and behavior.

### 2015-03-12 tex/main-body-text.tex
Revised version as submitted.

@@
 To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
-were simulated using a coalescent-based program (code provided at 
+were simulated using the coalescent program {\sc makesamples} \citep{Hudson2002} (code provided at 
 https://github.com/eriqande/flock-comment/). The 5 populations
@@
 the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
-the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
+the largest plateau. Multiple runs of {\sc flockture} demonstrated that the differences are well within the
 variability that we would expect from randomizing the initial grouping of individuals to 
@@
   Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
-  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
+  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flockture}). 
   Individuals are grouped left to right into their simulated population samples, and each 
   horizontal plot represents a run of the program
-  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
+  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flockture} (unless
   there were no plateaus, in which case the run reaching the highest likelihood was chosen).
@@
 (Note that although many of {\sc flocktures}'s solutions in B--D were visually indistinguishable, they were not identical partitions, and would
-hence be considered different for the purposes of creating a {\sc flock} plateau record.) Each simulation is comprised of
+hence be considered different for the purposes of creating a {\sc flock} plateau record.) In E, all 9
+of {\sc flocktures}'s solutions were visually distinguishable, though we show only one.  Each simulation is comprised of
  five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
@@
 The likelihood surface defined on partitions of this space is then often
-be multimodal---possessing numerous local peaks and troughs.
+multimodal---possessing numerous local peaks and troughs.
 The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
@@
 this advice of {\sc flock}'s authors would have led to discarding {\sc flockture}'s
-results for any $F_\mathrm{ST} < 0.032$ (almost 90\% of our simulated data sets!), even those that are
+results for any $F_\mathrm{ST} < 0.032$ (almost 90\% of our simulated data sets); even those that are
 manifestly quite accurate---some competitive
-with or even superior to {\sc structure}'s estimate!  We surmise that this problem with {\sc flock} 
-guidance issues from the fact that the {\em ad-hoc} rules developed by
+with or even superior to {\sc structure}'s estimate.  We surmise that this problem with the guidance on 
+{\sc flock} stems from the fact that the {\em ad-hoc} rules developed by
 \citeauthor{Duc&Tur2012} for estimating $K$ with {\sc flock} were developed empirically
@@
 fall outside the range of values explored
-by \citeauthor{Duc&Tur2012} it appears that {\sc flock}'s rules of use are inappropriate, 
+by \citeauthor{Duc&Tur2012}, it appears that {\sc flock}'s rules of use are inappropriate, 
 and perhaps over-restrictive for these data sets.  Researchers with data that similarly fall beyond the range of 
-data sets tested in \citeauthor{Duc&Tur2012} may wish to approach {\sc flock} analysis with
+data sets tested in \citet{Duc&Tur2012} may wish to approach {\sc flock} analysis with
 caution, or may choose to rely on other programs.

### 2015-03-23 tex/abstract.tex
minor changes to abstract

@@
 
-      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
+      We show that the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
 interpreted as an estimation procedure based on 
@@
 ($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
-for both marker types. On large data sets like those we simulated, it appears that 
+for both marker types. On large datasets like those we simulated, it appears that 
 {\sc flock}'s reliance on inference rules regarding its ``plateau record'' are not helpful. 

### 2015-03-23 tex/abstract.tex
minor edits for consistency and typos

@@
 {\sc flock}  and {\sc flockture} and very similar (though flockture is some 200 times faster). 
-Second, we simulated multiple large datasets under varying 
+Second, we simulated multiple large data sets under varying 
 levels of population differentiation for both microsatellite and SNP genotypes. We analyzed them
@@
 ($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
-for both marker types. On large datasets like those we simulated, it appears that 
+for both marker types. On large data sets like those we simulated, it appears that 
 {\sc flock}'s reliance on inference rules regarding its ``plateau record'' are not helpful. 

### 2015-03-23 tex/main-body-text.tex
minor edits for consistency and typos

@@
 programs from large 
-datasets simulated with varying degrees of population differentiation.
+data sets simulated with varying degrees of population differentiation.
 
@@
 which is the same full conditional distribution as the marginalized {\sc structure}
-model (equation~\ref{eq:fc}), raised
+model (\ref{eq:fc}), raised
 to the power $\beta$.  In typical applications of simulated annealing,
@@
 and {\sc flockture} are essentially equivalent programs.
-Second, we simulate and analyze datasets with {\sc flockture} 
+Second, we simulate and analyze data sets with {\sc flockture} 
 and three models in {\sc structure} (the admixture and correlated 
@@
 
-To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
+To compare {\sc flock} and {\sc flockture}, 4 data sets composed of 5 populations
 were simulated using the coalescent program {\sc makesamples} \citep{Hudson2002} (code provided at 
 https://github.com/eriqande/flock-comment/). The 5 populations
-were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
-Each of the four data sets was simulated with a different 
+were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed at 15 microsatellite loci.
+Each of the four data sets were simulated with a different 
 average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
@@
 Once we were confident that no major differences in inference existed between 
-{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
+{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 data sets for each of two marker types 
 (microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader range
-of population differentiation. The 96 datasets were simulated as above (five populations 
+of population differentiation. The 96 data sets were simulated as above (five populations 
 composed of 250, 250, 400, 550, and 550 individuals with 15 loci for the microsatellites
@@
 {\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
-from the posterior distribution. Each dataset was analyzed 9 times by each of the different
+from the posterior distribution. Each data set was analyzed 9 times by each of the different
 methods. All runs were performed on a Macbook with a 2.3 GHz i7 processor
@@
 {\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
-simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
+simulated data sets. The zero-one loss across all data sets had eqivalent results (Fig.~\ref{fig:FvFloss}).
 When genetic differentiation is high both programs correctly cluster the majority of individuals and 
@@
 clustering ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
-substantially among the simulated datasets.
+substantially among the simulated data sets.
 
@@
 persist in those endeavors when we realized that {\sc flock} itself does
-not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the
+not provide consistent results when expected by symmetry.  Briefly, in an attempt to understand the
 inner workings of {\sc flock} 
 we took a subset of the 
-first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
+first data set (29 individuals from population 1 and 21 individuals from population 2) and initiated 
 {\sc flock} with all individuals originating from population 1 being placed in reference population 1 and all 
@@
 
-Plateaus were only observed for the dataset 
+Plateaus were only observed for the data set 
 with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
@@
   \caption{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate 
-  runs of simulated datasets comprised of five populations with varying
+  runs of simulated data sets comprised of five populations with varying
   levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
@@
 \subsection*{{\sc flockture} and {\sc structure}} 
-As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
+As expected {\sc flockture} and {\sc structure} both perform extremely well on data sets with
 well-differentiated populations. Infrequently, at higher levels of population differentiation the 
@@
 with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct} A). This phenomenon 
-(only present in a single microsatellite dataset) is an instance of {\sc structure}'s
+(only present in a single microsatellite data set) is an instance of {\sc structure}'s
 Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
@@
 each model with different marker types emerged.
-For the microsatellite dataset we observed that both programs
+For the microsatellite data set we observed that both programs
 produce roughly equivalent results at higher levels of population differentiation. At intermediate 
@@
 below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among-run
-variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
+variance. The drastic increase in variance among runs was also observed with the SNP data sets. 
 {\sc flockture} did, however, have consistently smaller loss than  {\sc structure} at high to intermediate 
@@
 The number and length of plateaus {\sc flockture} encountered quickly 
-diminished as the amount of differentiation among populations decreased. For the microsatellite datasets,
+diminished as the amount of differentiation among populations decreased. For the microsatellite data sets,
 when $F_{ST}$ was less than 0.0337, no plateau record
@@
 (i.e. no two runs of {\sc flockture} gave exactly the same solution) when 
-$F_{ST}$ was smaller than 0.02187. Results from the SNP datasets
+$F_{ST}$ was smaller than 0.02187. Results from the SNP data sets
 were concordant.
-Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
-plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets, 
+Only simulated data sets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
+plateau greater than six. Despite the lack of plateaus observed for many of the simulated data sets, 
 graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
@@
 and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
-datasets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
+data sets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
 5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
@@
 and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
-datasets, composed of 2,000 individuals with 96 SNP genotypes comprising 
+data sets, composed of 2,000 individuals with 96 SNP genotypes comprising 
 5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
@@
 In comparing results from {\sc flockture} and {\sc structure} 
-on simulated datasets with varying levels of differentiation among populations
+on simulated data sets with varying levels of differentiation among populations
 we find that the two programs give similar results.
@@
 likelihood of assigning to any given cluster. This likely increases the number of
-local peaks, leading {\sc flockture} and {\sc flock} to settle on settle on different solutions in 
+local peaks, leading {\sc flockture} and {\sc flock} to settle on different solutions in 
 different runs. The problem presented by having multiple modes in the likelihood surface 
@@
 plateaus for $K=5$ in our simulated data, then the {\sc flock} method would lead to an estimate of $K$ that
-was either "undecided" or some value less than $K$. The authors of {\sc flock}
+was either "undecided" or some value less than $K$ as the lower bound or point estimate of $K$. The authors of {\sc flock}
 discourage users from making any inference at a value of $K$ that is not

### 2015-03-23 tex/main-body-text.tex
Do we want to reference the supplementarty tables? Comments placed at approx locations

@@
 Plateaus were only observed for the data set 
-with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
+with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although %ref supplementary table FLOCKvFLOCKTURE_PlateauTable.pdf
 the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
@@
 
-The number and length of plateaus {\sc flockture} encountered quickly 
+The number and length of plateaus {\sc flockture} encountered quickly %ref supp tables FLOCKTURE_PlateauTable_XX.pdf
 diminished as the amount of differentiation among populations decreased. For the microsatellite data sets,

### 2015-03-31 FLOCKTRUEvSTRUCTURE.R
Remove ECA reminder-comments from FLOCKTUREvSTRUCTURE.R

@@ -52,14 +52,12 @@ qi_indLoss <- F # set true if you want graphs of qi values and loss by simulatio
 
 ### Set up the parameters controlling the simulations 
 Seedset <-  3
-### ECA: gotta set reps back after testing
 Reps <- 9 # how many reps do we want to run for each program
 Npops <- 5
 N <- 2000
 
 ## set up migration rates to use 
 if(marker == 1) {
-  ## ECA got to set MGrate back after testing
   MGrate <- c(20,24,28,30,31,32,36,37,37.5,38,40,50,60,75,90,125)
 } else if(marker == 2) {
   MGrate<-c(20,24,28,30,31,34,36,36.8,37.5,38,40,50,60,75,80,87)

### 2015-04-01 FLOCKvFLOCKture.R
All files needed to do FLOCKvFlockture analysis added. Code now runs from within Flock-comment directory instead of flockture directory

@@ -1,0 +1,456 @@
+#### how does FLOCK compare with FLOCKTURE ####
+# First we will need to set up some paths, so that we can 
+# use the flockture script to analyze our datasets
+
+flockcommentDIR <- normalizePath("../flock-comment") 
+flocktureDIR <- normalizePath("../flockture") 
+
+#Next, load some useful libraries
+library(ggplot2)
+library(stringr)
+library(plyr)
+library(reshape2)
+library(miscFuncs)
+library(xtable)
+require("XLConnect")
+options(java.parameters = "-Xmx4g" ) # if you have a lot of java heap space errors,
+# you may need to run each of the Ds iterations individually, or run on a computer 
+# than has more memory. 
+
+# We will need to source the flockture code... it should be compiled first 
+# by running 'make' in terminal within scr folder
+setwd(flocktureDIR)
+source("R/flockture.R")
+
+#Now we will run Flockture#
+# We ran FLOCK 9 times with 50 reps and 20 interations... make sure
+# to system.time this to track how long it takes
+Datasets <- 4 #how many datasets do we have to read in?
+for (Ds in 1:Datasets){
+  #Files are named 'BaseFile_SimX.txt' where X is the run number
+  dat<-paste(flockcommentDIR,'/FLOCKvFLOCKTURE_data/data/BaseFile_Sim',Ds,'.txt',sep='')
+
+# read in a data set:
+D <- read.table(dat, row.names = 1)
+
+# define some starting conditions if you want (this is useful to test initial conditions):
+#sc <- c(c(rep(2,times=14),rep(1,times=15),rep(2,times=10),rep(1,times=11))
+#)
+
+FLCTR_PlateauRec<-vector()
+Flockture.Runtime<-vector()
+FlocktureReps<-9 #how many times are we running flockture
+for (FR in 1:FlocktureReps){
+# run flockture on it and grab the results out
+ptm<-proc.time()
+catch <- run_flockture_bin(D, K = 5, iter = 20, reps = 50)
+out <- slurp_flockture_dumpola()
+
[…]

### 2015-04-03 tex/abstract.tex
Updated README to reflect new stuff.
Prepping to push it off to Dryad.

@@
 the program {\sc flockture}. With simulated data we confirmed that results obtained with
-{\sc flock}  and {\sc flockture} and very similar (though flockture is some 200 times faster). 
+{\sc flock}  and {\sc flockture} are very similar (though flockture is some 200 times faster). 
 Second, we simulated multiple large data sets under varying 

### 2015-04-03 tex/main-body-text.tex
Final things to prep for final submission.

--- 2015-03-23 tex/main-body-text.tex
@@
 Plateaus were only observed for the data set 
-with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
+with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although %ref supplementary table FLOCKvFLOCKTURE_PlateauTable.pdf
 the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
-the largest plateau. Multiple runs of {\sc flockture} demonstrated that the differences are well within the
+the largest plateau (see Supporting Information). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
 variability that we would expect from randomizing the initial grouping of individuals to 
@@
 
-The number and length of plateaus {\sc flockture} encountered quickly 
-diminished as the amount of differentiation among populations decreased. For the microsatellite data sets,
+The number and length of plateaus {\sc flockture} encountered quickly %ref supp tables FLOCKTURE_PlateauTable_XX.pdf
+diminished as the amount of differentiation among populations decreased (Supporting Information). For the microsatellite data sets,
 when $F_{ST}$ was less than 0.0337, no plateau record

### 2015-04-03 tex/PREPARE_FILES_FOR_BLACKWELL_SUBMISSION.sh

@@ -1,0 +1,30 @@
+
+
+
+# this is a little script that will create a directory on my desktop
+# with all the files needed to typeset the final submission to
+# Molecular Ecology Resources
+
+DIR=~/Desktop/flock-comment-tex-stuff
+
+rm -r $DIR
+
+mkdir $DIR
+
+cp  -r flock-comment-one-column-main.tex  \
+	flock-comment-one-column-main.bbl \
+	abstract.tex  \
+	author-title-etc.tex \
+ 	main-body-text.tex \
+ 	lineno.sty \
+ 	images \
+ 	$DIR/
+
+
+# now remove some of the files that are not needed/referenced
+rm 	-r $DIR/images/NMFS_letterhead.pdf \
+	$DIR/images/banner.pdf \
+	$DIR/images/Tables-Pat \
+	$DIR/images/Figures-Pat/*BW.pdf
+
+
